# The Big (Bayesian) Picture {#sec-chap-001}


```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")

```



::::: {#obj-chap-001}
:::: {.my-objectives}
::: {.my-objectives-header}
Objectives
:::

::: {.my-objectives-container}
- Learn to think like a Bayesian. 
- Explore the foundations of a Bayesian data analysis and how they contrast with the frequentist alternative. 
- Learn a little bit about the history of the Bayesian philosophy.

:::
::::
:::::

## Think like an Bayesian

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-001-thinking-bayesian-1}
: Bayesian thinking: Example Italian Restaurant
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: bayesian-thinking-1

DiagrammeR::grViz("
  digraph thinking_bayesian{
  
  # node statement
  node [shape = oval]
  a [label = 'Incoming (prior) Information\nFive start restaurant'];
  b [label = 'Data:\nsoggy noodles'];
  c [label = 'Updates (posterior) information\nThree star restaurant'];
  d [label = 'Another visit (new data):\ntasty alfredo'];
  e [label = 'Updated (posterior) information\nFour star reastaurant'];
  f [label = 'Another visit: (new data):\nso-so salad']
  g [style = invisible ]
  
  # edge statement
  a -> c b -> c
  c -> e d -> e
  f-> g [style = dashed] e-> g [style = dashed]
  }")
```

::::
:::::

Summarizing and generalizing the above graphic we get the following graph:

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-001-bayesian-thinking-2}
: Bayesian thinking: The general process
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: bayesian-thinking-2

DiagrammeR::grViz("
  digraph thinking_bayesian{
  
  # node statement
  node [shape = oval]
  a [label = 'Prior'];
  b [label = 'Data'];
  c [label = 'Posterior'];
  d [label = 'New data'];
  e [label = 'Posterior'];
  f [label = 'New data']
  g [style = invisible ]
  
  # edge statement
  a -> c b -> c
  c -> e d -> e
  f-> g [style = dashed] e-> g [style = dashed]
  }")
```

::::
:::::





## Glossary Entries {.unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```

------------------------------------------------------------------------

## Session Info {.unnumbered}

::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::

