# Conjugate Families {#sec-chap-005}

```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")
ggplot2::theme_set(ggplot2::theme_bw())
```

:::::: {#obj-chap-005}
::::: my-objectives
::: my-objectives-header
Objectives
:::

::: my-objectives-container
-   **Practice building Bayesian models**. Build Bayesian models by
    practicing how to recognize kernels and make use of proportionality.
-   **Familiarize yourself with conjugacy**. Learn about what makes a
    prior conjugate and why this is a helpful property.
:::
:::::
::::::

To prepare for this chapter, note that we’ll be using some new Greek
letters throughout our analysis:

-   $\lambda$ = lambda,
-   $\mu$ = mu or “mew”,
-   $\sigma$ = sigma,
-   $\tau$ = tau, and
-   $\theta$ = theta.

## Revisiting choice of prior

In @sec-chap-003 and @sec-chap-004 we used the *flexibility* of the Beta
model to reflect our prior understanding of a proportion parameter
$\pi \in [0, 1]$. There are other criteria to consider when choosing a
prior model:

-   **Computational ease**: Especially if we don’t have access to
    computing power, it is helpful if the posterior model is easy to
    build.
-   **Interpretability**:: We’ve seen that posterior models are a
    compromise between the data and the prior model. A posterior model
    is interpretable, and thus more useful, when you can look at its
    formulation and *identify* the contribution of the data relative to
    that of the prior.

The Beta-Binomial has both of these criteria covered. Its calculation is
easy. Once we know the $\text{Beta}(\alpha, \beta)$ prior
`r glossary("hyperparameter", "hyperparameters")` and the observed data
$Y = y$ for the $\text{Bin}(n, \pi)$ model, the
$\text{Beta}(\alpha + y, \beta + n − y)$ posterior model follows.

This posterior reflects the influence of the data, through the values of
$y$ and $n$, relative to the prior hyperparameters $\alpha$ and $\beta$.
If $\alpha$ and $\beta$ are large relative to the sample size $n$, then
the posterior will not budge that much from the prior. However, if the
sample size $n$ is large relative to $\alpha$ and $\beta$, then the data
will take over and be more influential in the posterior model. In fact,
the `r glossary("Beta-Binomial-Model", "Beta-Binomial")` belongs to a
larger class of prior-data combinations called **conjugate families**
that enjoy both computational ease and interpretable posteriors. Recall
a general definition similar to that from @imp-003-conjugate-prior.

::: {.callout-important #imp-005-conjugate-prior}
###### Conjugate prior

Let the prior model for parameter $\theta$ have pdf f ($\theta$) and the
model of data Y conditioned on $\theta$ have likelihood function
$L(\theta \mid y)$. If the resulting posterior model with pdf f
$(\theta \mid y) \propto f (\theta)L(\theta \mid y)$ is of the same
model family as the prior, then we say this is a
`r glossary("conjugate prior")`.
:::


To emphasize the utility of conjugate priors, it can be helpful to
consider a *non*-conjugate prior. Let parameter $\pi$ be a proportion
between $0$ and $1$ and suppose we plan to collect data $Y$ where,
conditional on $\pi$, $Y \mid \pi \sim \text{Bin}(n, \pi)$. Instead of
our conjugate $\text{Beta}(\alpha, \beta)$ prior for $\pi$, let’s try
out a non-conjugate prior with pdf f ($\pi$), plotted in Figure 5.1:

$$
\begin{align*}
f (\pi) = e − e^π \text{ for } \pi \in [0, 1]
\end{align*}
$$ {#eq-005-non-conjugate-prior}

Though *not* a Beta pdf, f ($\pi$) is indeed a *valid* pdf since f
($\pi$) is non-negative on the support of $\pi$ and the area under the
pdf is $1$, i.e.,$\int_{0}^{1} f (\pi) = 1$.

```{r}
#| label: fig-005-non-conjugate-prior
#| fig-cap: A non-conjugate prior for $\pi$
#| lst-label: lst-005-non-conjugate-prior
#| lst-cap: Valid non-conjugate prior
#| fig-width: 4
#| fig-height: 3

# Create a data frame with pi values and corresponding f(pi) values
pi_values <- base::seq(0, 1, length.out = 1000)
df <- tibble::tibble(
  pi = pi_values,
  f_pi = base::exp(1) - base::exp(pi_values)
)

# Create the plot
ggplot2::ggplot(df, ggplot2::aes(x = pi, y = f_pi)) +
  ggplot2::geom_line(color = "black", linewidth = 1) +
  ggplot2::labs(
    title = base::expression(f(pi) == e - e^pi),
    x = base::expression(pi),
    y = base::expression(f(pi))
  ) +
  ggplot2::theme_minimal()
```

Next, suppose we observe $Y = 10$ successes from $n = 50$ independent
trials, all having the same probability of success $\pi$. The resulting
Binomial likelihood function of $\pi$ is:

$$L(\pi \mid y = 10) = \binom{50}{10}\pi^{10}(1 − \pi)^{40} \text{ for } \pi \in [0, 1]$$

Recall from @sec-chap-003 that, when we put the
`r glossary("priorx", "prior")` and the
`r glossary("likelihood_x", "likelihood")` together, we are already on
our path to finding the
`r glossary("posterior_model", "posterior model")` with
`r glossary("pdf")`

$$f (\pi \mid y = 10) \propto f (\pi)L(\pi \mid y = 10) = (e − e^π) \cdot \binom{50}{10}\pi^{10}(1 − \pi)^{40}$$

As we did in @sec-chap-003, we will drop all constants that do not
depend on $\pi$ since we are only specifying $f (\pi \mid y)$ up to a
proportionality constant:

$$f (\pi \mid y) \propto (e − e^\pi)\pi^{10}(1 − \pi)^{40}$$

Notice here that our non-Beta prior didn’t produce a neat and clean
answer for the exact posterior model (fully specified and not up to a
proportionality constant). We cannot squeeze this posterior pdf kernel
into a Beta box or any other familiar model for that matter. That is, we
*cannot* rewrite $(e − e^\pi)\pi^{10}(1 − \pi)^{40}$ so that it shares
the same structure as a Beta kernel, $\pi^{∎−1}(1 − \pi)^{∎−1}$. Instead
we will need to integrate this kernel in order to complete the
normalizing constant, and hence posterior specification:

$$
\begin{align*}
f (\pi \mid y = 10) =  \frac{(e − e^{\pi})\pi^{10}(1 − \pi)^{40}}{\int_{0}^{1}(e − e^{\pi})\pi^{10}(1 − \pi)^{40}dx}
\end{align*}
$$ {#eq-non-conjugate-posterior-specification}

This is where we really start to feel the pain of not having a conjugate
prior! Since this is a particularly unpleasant integral to evaluate, and
we’ve been trying to avoid doing any integration altogether, we will
leave ourselves with @eq-non-conjugate-posterior-specification as the
final posterior. Yikes, what a mess. This *is* a valid posterior pdf –
it’s non-negative and integrates to $1$ across $\pi \in [0, 1]$. But it
doesn’t have much else going for it. Consider a few characteristics
about the posterior model that result from this particular non-conjugate
prior in @eq-005-non-conjugate-prior:

-   The calculation for this posterior was messy and unpleasant.
-   It is difficult to derive any intuition about the balance between
    the prior information and the data we observed from this posterior
    model.
-   It would be difficult to specify features such as the posterior
    mean, mode, and standard deviation, a process which would require
    even more integration.

This leaves us with the question: could we use the conjugate Beta prior
and still capture the broader information of the messy non-conjugate
prior in @eq-005-non-conjugate-prior? If so, then we solve the problems
of messy calculations and indecipherable posterior models.

:::::: my-assessment
:::: my-assessment-header
::: {#cor-005-which-beta-for-non-jugate-prior}
: Which Beta model?
:::
::::

::: my-assessment-container
Which Beta model would most closely approximate the non-conjugate prior
for $\pi$ in @fig-005-non-conjugate-prior?
`r mcq(c("Beta(3,1)", "Beta(1,3)", "Beta(2,1)", answer = "Beta(1,2)"))`
:::
::::::

One way to find the answer is by comparing the non-conjugate prior in
@fig-005-non-conjugate-prior with plots of the possible Beta prior
models. For example, the non-conjugate prior information is pretty well
captured by the $\text{Beta}(1, 2)$

:::::: my-r-code
:::: my-r-code-header
::: {#cnj-005-beta-1-2}
: A conjugate $\text{Beta}(1, 2)$ prior model for $\pi$.
:::
::::

::: my-r-code-container
```{r}
#| label: fig-005-beta-1-2
#| fig-cap: A conjugate $\text{Beta}(1, 2)$ prior model for $\pi$
#| fig-width: 4
#| fig-height: 3

bayesrules::plot_beta(1, 2)
```
:::
::::::

## Gamma-Poisson conjugate family

Last year, one of this book’s authors got fed up with the number of
fraud risk phone calls they were receiving. They set out with a goal of
modeling *rate* $\lambda$, the typical number of fraud risk calls
received per day. **Prior** to collecting any data, the author’s guess
was that this rate was most likely around $5$ calls per day, but could
also reasonably range between $2$ and $7$ calls per day. To learn more,
they planned to record the number of fraud risk phone calls on each of
$n$ sampled days, $(Y_1, Y_2, . . . , Y_n)$.

In moving forward with our investigation of $\lambda$, it’s important to
recognize that it will *not* fit into the familiar Beta-Binomial
framework.

-   First, $\lambda$ is *not* a proportion limited to be between $0$ and
    $1$, but rather a rate parameter that can take on any *positive*
    value (e.g., we can’t receive -7 calls per day). Thus, a Beta prior
    for $\lambda$ won’t work.
-   Further, each data point $Y_{i}$ is a *count* that can technically
    take on any non-negative integer in $\{0, 1, 2, . . .\}$, and thus
    is *not* limited by some number of trials $n$ as is true for the
    Binomial. Not to fret. Our study of $\lambda$ will introduce us to a
    *new* conjugate family, the
    `r glossary("Gamma-Poisson model", "Gamma-Poisson")`.

### The Poisson data model

#### Poisson formulas

The *spirit* of our analysis starts with a prior understanding of
$\lambda$, the daily rate of fraud risk phone calls. Yet before choosing
a prior model structure and tuning this to match our prior
understanding, it’s beneficial to identify a model for the dependence of
our daily phone call *count* data $Y_i$ on the typical daily *rate* of
such calls $\lambda$. Upon identifying a reasonable data model, we can
identify a prior model which can be tuned to match our prior
understanding while *also* mathematically complementing the data model’s
corresponding likelihood function. Keeping in mind that each data point
$Y_i$ is a *random count* that can go from $0$ to a really big number,
$Y \in \{0, 1, 2, . . .\}$, the
`r glossary("Poisson-model", "Poisson model")`, described in its general
form below, makes a reasonable candidate for modeling this data.

:::::: my-theorem
:::: my-theorem-header
::: {#thm-005-poisson-model}
: Poisson model
:::
::::

::: my-theorem-container
Let discrete random variable $Y$ be the *number of independent events*
that occur in a fixed amount of time or space, where $\lambda > 0$ is
the rate at which these events occur. Then the *dependence* of $Y$ on
`r glossary("parameter_x", "parameter")` $\lambda$ can be modeled by the
Poisson. In mathematical notation:

$$Y \mid \lambda \sim Pois(\lambda)$$

Correspondingly, the Poisson model is specified by pmf

$$
f (y \mid \lambda) =  \frac{\lambda^y e^{−\lambda}}{y!} \text{ for } y \in \{0, 1, 2, . . .\}
$$ {#eq-005-poisson-model}

where $f (y \mid \lambda)$ sums to one across
$y, \sum_{y=0}^{\infty} f (y \mid \lambda) = 1$. Further, a Poisson
random variable $Y$ has equal mean and variance,

$$
E(Y \mid \lambda) = \text{Var}(Y \mid \lambda) = \lambda
$$ {#eq-005-poisson-equal-mean-variance}
:::
::::::

#### Poisson illustrations

@eq-005-poisson-model illustrates the Poisson pmf @eq-005-poisson-model under
different rate parameters $\lambda$. In general, as the rate of events
$\lambda$ increases, the typical number of events increases, the
variability increases, and the skew decreases. For example, when events
occur at a rate of $\lambda = 1$, the model is heavily skewed toward
observing a small number of events – we’re most likely to observe $0$ or
$1$ events, and rarely more than 3. In contrast, when events occur at a
higher rate of $\lambda = 5$, the model is roughly symmetric and more
variable – we’re most likely to observe $4$ or $5$ events, though have a
reasonable chance of observing anywhere between $1$ and $10$ events.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-plot-poisson-models}
: Plot for illustration some poisson pmfs with different rate parameters.
::::::
:::
::::{.my-r-code-container}


```{r}
#| label: fig-005-plot-poisson-models
#| fig-cap: Poisson pmfs with different rate parameters.
#| fig-height: 3

# Create data frame with all combinations
data <- base::expand.grid(
  x = 0:12,  # number of events
  lambda = c(1, 3, 5) # rate of events
) |> 
  dplyr::mutate(
    probability = stats::dpois(x = x, lambda = lambda),
    prob_label = base::paste0("Pois(", lambda, ")")
  )

# Create lollipop chart with facets
ggplot2::ggplot(data, ggplot2::aes(x = x, y = probability)) +
  ggplot2::geom_linerange(
    ggplot2::aes(
      x = x, 
      ymin = 0, 
      ymax = probability, 
      color = "black"), 
    linewidth = 0.8) +
  ggplot2::geom_point(ggplot2::aes(color = "black"), size = 1) +
  ggplot2::scale_color_identity() +
  ggplot2::facet_wrap(~ prob_label, ncol = 3) +
  ggplot2::scale_x_continuous(breaks = base::seq(0, 12, by = 2)) +
  # ggplot2::scale_y_continuous(breaks = base::seq(0, 0.20, by = 0.05)) +
  ggplot2::labs(
    x = "y",
    y = base::expression("f (y)")
  ) +
  ggplot2::theme_minimal() +
    ggplot2::theme(
    panel.spacing = grid::unit(1.5, "lines"),
    strip.background = ggplot2::element_rect(fill = "gray90", color = "gray50"),
    strip.text = ggplot2::element_text(face = "bold", size = 11)
  )
```

::::
:::::

Let $(Y_1, Y_2, . . . , Y_n)$ denote the number of fraud risk calls we
observed on each of the $n$ days in our data collection period. We
assume that the daily number of calls might differ from day to day and
can be **independently modeled** by the Poisson. Thus, on each day $i$,

$$Y \mid \lambda \sim Pois(\lambda)$$

with unique pmf

$$
f (y \mid \lambda) =  \frac{\lambda^y e^{−\lambda}}{y!} \text{ for } y \in \{0, 1, 2, . . .\}
$$

#### Joint pmf

Yet in weighing the evidence of the phone call data, we won’t want to
analyze each *individual* day. Rather, we’ll need to process the
*collective* or *joint* information in our *n* data points. This
information is captured by the
`r glossary("Joint_Probability_Mass_Function", "joint probability mass function")`.

:::::: my-theorem
:::: my-theorem-header
::: {#thm-005-joint-pmf}
: Joint probability mass function
:::
::::

::: my-theorem-container
Let $(Y_1, Y_2, . . . , Y_n)$ be an independent sample of random
variables and $\overrightarrow{\rm y} = (y_1, y_2, . . . , y_n)$ be the
corresponding vector of observed values. Further, let
$f (y_i \mid \lambda)$ denote the pmf of an individual observed data
point $Y_i = y_i$. Then by the assumption of independence, the following
**joint pmf** specifies the randomness in and plausibility of the
collective sample:

$$
\begin{align*}
f (\overrightarrow{\rm y} \mid \lambda) = \prod_{i = 1}^n f (y_i \mid \lambda) = f (y_1 \mid \lambda) \cdot f (y_2 \mid \lambda) \cdot \cdots \cdot f (y_n \mid \lambda)
\end{align*}
$$ {#eq-005-joint-pmf}

**Connecting concepts**

This product is analogous to the joint probability of independent events
being the product of the
`r glossary("Marginal_Probability", "marginal probabilities")`,
$P (A \cap B) = P (A)P (B)$.
:::
::::::

The joint pmf for our fraud risk call sample follows by applying the
general definition in @eq-005-joint-pmf to our Poisson pmfs. Letting the
number of calls on each day $i$ be $y_i \in \{0, 1, 2, . . .\}$,

$$
f (\overrightarrow{\rm y} \mid \lambda) = \prod_{i = 1}^n f (y_i \mid \lambda) = \prod_{i = 1}^n \frac{\lambda^{y_i} e^{−λ}}{y_i!}
$$ {#eq-005-joint-pfm-fraud-calls}

This looks like a mess, but it can be simplified. In this
simplification, it’s important to recognize that we have $n$ unique data
points $y_i$, not $n$ copies of the same data point $y$. Thus, we need
to pay careful attention to the $i$ subscripts. It follows that

$$
\begin{align*}
f (\overrightarrow{\rm y} \mid \lambda) &= \frac{\lambda^{y_1} e^{−λ}}{y_1!} \cdot \frac{\lambda^{y_2} e^{−λ}}{y_2!} \cdots \frac{\lambda^{y_n} e^{−λ}}{y_n!} \\
&= \frac{[\lambda^{y_1} \lambda^{y_2} \cdots \lambda^{y_n}][e^{-\lambda} e^{-\lambda} \cdots e^{-\lambda}]}{y_1! y_2! \cdots y_n!} \\
&= \frac{\lambda^{\sum y_i }e^{-n\lambda}}{\prod_{i = 1}^n y_i!}
\end{align*}
$$

where we’ve simplified the products in the final line by appealing to
the properties below.

:::::: my-theorem
:::: my-theorem-header
::: {#thm-005-simplifying-products}
: Simplifying products
:::
::::

::: my-theorem-container
Let $(x, y, a, b)$ be a set of constants. Then we can utilize the
following facts when simplifying products involving exponents:

$$
\begin{align*}
x^ax^b = x^{a+b} \text{ and } x^ay^a = (xy)^a
\end{align*}
$$ {#eq-005-simplfiying-products}
:::
::::::

#### Likelihood function

Once we observe actual sample data, we can flip this joint pmf on its
head to define the
`r glossary("likelihood-function", "likelihood function")` of $\lambda$.
The Poisson likelihood function is equivalent in formula to the joint
pmf $f (\overrightarrow{\rm y} \mid \lambda)$, yet is a function of
$\lambda$ which helps us assess the compatibility of different possible
$\lambda$ values with our observed *collection* of sample data
$\overrightarrow{\rm y}$:

$$
L (\overrightarrow{\rm y} \mid \lambda) = \frac{\lambda^{\sum y_i }e^{-n\lambda}}{\prod_{i = 1}^n y_i!} \propto \lambda^{\sum y_i }e^{-n\lambda} \text { for } \lambda > 0
$$ {#eq-005-poisson-likelihood-function}

It is convenient to represent the likelihood function up to a
proportionality constant here, especially since $\prod y_i!$ will be
cumbersome to calculate when $n$ is large, and what we really care about
in the likelihood is $\lambda$. And when we express the likelihood up to
a proportionality constant, note that the sum of the data points
($\sum y_i$) and the number of data points ($n$) is all the information
that is required from the data. We don’t need to know the value of each
individual data point $y_i$. Taking this for a spin with real data
points later in our analysis will provide some clarity.

### Potential priors

The Poisson data model provides one of two key pieces for our Bayesian
analysis of $\lambda$, the daily rate of fraud risk calls. The other key
piece is a prior model for $\lambda$. Our original guess was that this
rate is most likely around 5 calls per day, but could also reasonably
range between 2 and 7 calls per day. In order to tune a prior to match
these ideas about $\lambda$, we first have to identify a reasonable
probability model structure. Remember here that $\lambda$ is a positive
and continuous rate, meaning that $\lambda$ does not have to be a whole
number. Accordingly, a reasonable prior probability model will also have
continuous and positive support, i.e., be defined on $\lambda > 0$.
There are several named and studied probability models with this
property, including the `r glossary("F-Distribution-Model", "F")` ,
`r glossary("Weibull model", "Weibull")`, and
`r glossary("Gamma-Model", "Gamma")`. We don’t dig into all of these in
this book. Rather, to make the $\lambda$ posterior model construction
more straightforward and convenient, we’ll focus on identifying a
`r glossary("conjugate prior")` model.

:::::: my-assessment
:::: my-assessment-header
::: {#cor-005-conjugate-prior}
: Choose the correct conjugate prior
:::
::::

::: my-assessment-container
Suppose we have a random sample of Poisson random variables
$(Y_1, Y_2, . . . , Y_n)$ with likelihood function
$L (\overrightarrow{\rm y} \mid \lambda) = \propto \lambda^{\sum y_i }e^{-n\lambda} \text { for } \lambda > 0$

-   

    a.  A “Gamma” model with pdf
        $f (\lambda) \propto \lambda^{s−1}e^{−r\lambda}$\

-   

    b.  A “Weibull” model with pdf
        $f (\lambda) \propto \lambda^{s−1}e^{(−r\lambda)^s}$

-   

    c.  A special case of the “F” model with pdf
        $f (\lambda) \propto \lambda^{ \frac{s}{2}-1} (1 + \lambda)^{−s}$

What do you *think* would provide a convenient *conjugate* prior model
for $\lambda$?: `r mcq(c( answer = "a", "b", "c"))`

Why?
:::
::::::

The Gamma model will provide a conjugate prior for $\gamma$ when our
data has a Poisson model. You might have guessed this from the section
title (clever). You might also have guessed this from the shared
features of the Poisson likelihood function
$L (\overrightarrow{\rm y} \mid \lambda)$ in
@eq-005-poisson-likelihood-function and the Gamma pdf $f (\lambda)$.
Both are proportional to

$$\lambda^∎e^{−∎\lambda}$$\
with differing ∎. In fact, we’ll prove that *combining* the prior and
likelihood produces a posterior pdf with this same structure. That is,
the posterior will be of the same Gamma model family as the prior.
First, let’s learn more about the Gamma model.

### Gamma prior

#### Gamma formulas {#sec-005-gamma-formulas}

:::::: my-theorem
:::: my-theorem-header
::: {#thm-005-gamma-prior}
: Gamma and Exponential models
:::
::::

::: my-theorem-container
Let $\lambda$ be a continuous random variable which can take any
positive value, i.e., $\lambda > 0$. Then the variability in $\lambda$
might be well modeled by a Gamma model with
`r glossary("Shape-Parameter", "shape hyperparameter")` $s > 0$ and
`r glossary("Rate-Parameter", "rate hyperparameter")` $r > 0$:

$$\lambda \sim \text{Gamma}(s, r)$$\
The Gamma model is specified by continuous pdf

$$
f (\lambda) =  \frac{r^s}{\tau(s)}  \lambda^{s−1}e^{−r\lambda} \text{ for } \lambda > 0
$$ {#eq-005-gamma-model}

Further, the central tendency and variability in $\lambda$ are measured
by:

$$
\begin{align*}
E(\lambda) &=  \frac{s}{r} \\
Mode(\lambda) &=  \frac{s−1}{r} \text{ for } s ≥ 1 \\
Var(\lambda) &=  \frac{s}{r^2}
\end{align*}
$$ {#eq-005-gamma-central-measures}

The `r glossary("Exponential-model", "Exponential model")` is a special
case of the Gamma with shape $s = 1, \text{Gamma}(1, r)$:

$$
\lambda \sim \text{Exp}(r)
$$
:::
::::::

#### Gamma illustrations

Notice that the Gamma model depends upon two hyperparameters, `r` and
`s`.


:::::: my-r-code
:::: my-r-code-header
::: {#cnj-005-r-and-s-effects}
: Impact of hyperparameters `r` and `s` in Gamma models
:::
::::

::: my-r-code-container
```{r}
#| label: fig-005-different-gamma-models
#| fig-cap: Gamma models with different hyperparameters. The dashed and solid vertical lines represent the modes and means, respectively.

# Create data for all parameter combinations
params <- tibble::tribble(
  ~shape, ~rate,
  1, 1,
  2, 1,
  4, 1,
  1, 2,
  2, 2,
  4, 2
) |>
  dplyr::mutate(label = stringr::str_glue("Gamma({shape},{rate})"))

# Generate data for each combination
plot_data <- params |>
  dplyr::rowwise() |>
  dplyr::reframe(
    tibble::tibble(
      x = base::seq(0, 8, length.out = 200),
      y = stats::dgamma(x, shape = shape, rate = rate),
      label = label,
      shape = shape,
      rate = rate
    )
  )

# Generate segment data for mean and mode
segments_data <- params |>
  dplyr::rowwise() |>
  dplyr::reframe(
    mean_val = shape / rate,
    mode_val = (shape - 1) / rate,
    tibble::tibble(
      x = c(mean_val, mode_val),
      xend = c(mean_val, mode_val),
      y = 0,
      yend = c(
        stats::dgamma(mean_val, shape, rate),
        stats::dgamma(mode_val, shape, rate)
      ),
      line_type = c("mean", "mode"),
      label = label,
      rate = rate
    )
  )

# Create faceted plot
plot_data |>
  dplyr::arrange(rate) |>
  dplyr::mutate(label = forcats::fct_inorder(label)) |>
  ggplot2::ggplot(ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_line() +
  ggplot2::geom_segment(
    data = segments_data |> 
      dplyr::arrange(rate) |>
      dplyr::mutate(label = forcats::fct_inorder(label)),
    ggplot2::aes(x = x, y = y, xend = xend, yend = yend, linetype = line_type),
    color = "darkblue"
  ) +
  ggplot2::facet_wrap(~ label, ncol = 3) +
  ggplot2::scale_x_continuous(breaks = base::seq(0, 8, 1), limits = c(0, 8)) +
  ggplot2::scale_y_continuous(breaks = base::seq(0, 2, 0.5), limits = c(0, 2)) +
  ggplot2::scale_linetype_manual(
    values = c(mean = "solid", mode = "dashed"),
    name = NULL
  ) +
  ggplot2::labs(
    x = base::expression(lambda),
    y = base::expression(paste("f(", lambda, ")"))
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    panel.spacing = grid::unit(1.5, "lines"),
    strip.background = ggplot2::element_rect(fill = "gray90", color = "gray50"),
    strip.text = ggplot2::element_text(face = "bold", size = 11)
  )
```
:::
::::::



Assess your understanding of how these hyperparameters impact the Gamma
model properties in the following quiz.

::::::::: my-assessment
:::: my-assessment-header
::: {#cor-005-r-and-s-effects}
: Impact of hyperparameters `r` and `s` in Gamma models
:::
::::

:::::: my-assessment-container
@fig-005-different-gamma-models illustrates how different shape and rate
hyperparameters impact the Gamma pdf in @eq-005-gamma-model. Based on
these plots:

1.  How would you describe the typical behavior of a
    $\text{Gamma}(s, r)$ variable $\lambda$ when $s > r$ (e.g.,
    $\text{Gamma}(2,1)$)?
    `r mcq(c(answer = "Right-skewed with a mean greater than 1", "Right-skewed with a mean less than 1", "Symmetric with a mean around 1"))`
2.  how would you describe the typical behavior of a
    $\text{Gamma}(s, r)$ variable $\lambda$ when $s < r$ (e.g.,
    $\text{Gamma}(1,2)$)?
    `r mcq(c("Right-skewed with a mean greater than 1", answer =  "Right-skewed with a mean less than 1", "Symmetric with a mean around 1"))`

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-005-gamma-20-20
#| fig-cap: $\text{Gamma}(20,20)$
#| fig-height: 3
#| code-fold: false

my_plot_gamma(20, 20)
```
:::

::: {.column width="50%"}
```{r}
#| label: fig-005-gamma-20-100
#| fig-cap: $\text{Gamma}(20,100)$
#| fig-height: 3
#| code-fold: false

my_plot_gamma(20, 100)
```
:::
:::::

3.  For which model is there greater variability in the plausible values
    of $\lambda$, $\text{Gamma}(20,20)$ or $\text{Gamma}(20,100)$?
    `r mcq(c( answer = "Gamma(20, 20)", "Gamma(20, 100)"))`
::::::
:::::::::

::::: {#nte-005-shape-rate-hyperparameter .callout-note}
###### Shape and Rate Hyperparameters

::: {#big-text style="font-size: 125%"}
**Shape Parameter**
:::

The shape parameter in the context of the gamma distribution is a key
`r glossary("hyperparameter")` that controls the form or shape of the
distribution. It is often denoted as $\alpha$ (alpha).

-   A **higher shape** parameter ($\alpha$) results in a more **peaked
    and right-skewed distribution**, indicating that events are **more
    concentrated** around a specific value.
-   A **lower shape** parameter leads to a **flatter, more spread-out
    distribution**.

In practical applications like modeling waiting times or right-skewed
data (e.g., insurance claims, rainfall), **the shape parameter helps
determine how the data is distributed**. For instance, in the gamma
shape model, covariates influence the distribution through the shape
parameter, allowing the variance to be directly proportional to the
mean—making it useful for modeling data with increasing variability.

::: {style="font-size: 125%"}
**Rate Parameter**
:::

The rate hyperparameter in the gamma distribution, typically denoted as
$\beta$ (beta), controls the scale of the distribution and is the
**inverse of the scale parameter** (i.e., rate = 1/scale).

-   A **higher rate** means events occur more frequently, leading to a
    **narrower** and more **concentrated** distribution.
-   A **lower rate** implies less frequent events, resulting in a
    **broader** distribution.

In applications like reliability analysis or queuing theory, **the rate
parameter reflects the average occurrence rate of events per unit
time**. For example, if modeling time between customer arrivals, a rate
of $beta = 2$ means, on average, two customers arrive per time unit, and
the expected waiting time follows a gamma distribution with this rate.
:::::

In general, @fig-005-different-gamma-models illustrates that
$\lambda \sim \text{Gamma}(s, r)$ variables are positive and right
skewed. Further, the general shape and rate of decrease in the skew are
controlled by hyperparameters `s` and `r`. The quantitative measures of
central tendency and variability in @eq-005-gamma-central-measures
provide some insight. For example, notice that the mean of
$\lambda, E(\lambda) = s/r$, is greater than $1$ when $s > r$ and less
than $1$ when $s < r$. Further, as `s` increases relative to `r`, the
skew in $\lambda$ decreases and the variability,
$\text{Var}(\lambda) = s/r^2$, increases.

#### Tuning Gamma prior

Now that we have some intuition for how the $\text{Gamma}(s, r)$ model
works, we can tune it to reflect our prior information about the daily
rate of fraud risk phone calls $\lambda$. Recall our earlier assumption
that $\lambda$ is about $5$, and most likely somewhere between $2$ and
$7$. Our $\text{Gamma}(s, r)$ prior should have similar patterns. For
example, we want to pick `s` and `r` for which $\lambda$ tends to be
around $5$,

$$E(\lambda) =  \frac{s}{r} \approx 5$$

This can be achieved by setting `s` to be $5$ times `r`, $s = 5r$. Next
we want to make sure that most values of our $\text{Gamma}(s, r)$ prior
are between $2$ and $7$. Through some trial and error within these
constraints, and plotting various Gamma models using
`bayesrules::plot_gamma()`, we find that the $\text{Gamma}(10,2)$
features closely match the central tendency *and* variability in our
prior understanding (@fig-005-plot-gamma-10-2). Thus, a *reasonable*
prior model for the daily rate of fraud risk phone calls is

$$\lambda \sim \text{Gamma}(10, 2)$$

with **prior pdf** $f (\lambda)$ following from plugging $s = 10$ and
$r = 2$ into

$$
\begin{align*}
f (\lambda) &=  \frac{r^s}{\tau(s)}  \lambda^{s−1}e^{−r\lambda} \text{ for } \lambda > 0 \\
&= \frac{2^{10}}{\tau(10)}  \lambda^{10−1}e^{−2\lambda} \text{ for } \lambda > 0
\end{align*}
$$

:::::: my-r-code
:::: my-r-code-header
::: {#cnj-005-plot-gamma-10-2}
: Plot the $\text{Gamma}(10, 2)$ prior
:::
::::

::: my-r-code-container
```{r}
#| label: fig-005-plot-gamma-10-2
#| fig-cap: The pdf of a $\text{Gamma}(10, 2)$ prior for $\lambda$, the daily rate of fraud risk calls.
#| fig-height: 3

bayesrules::plot_gamma(10, 2)
```
:::
::::::

::::::: {#nte-005-plot-gamma-11-2 .callout-note}
At first it seemed to me that $\text{Gamma}(11, 2)$ would have been a
slightly better choice: It has its mode at exactly $5$ and it is similar
spread out with a slightly bigger chance to get more values above $10$.

:::::: my-r-code
:::: my-r-code-header
::: {#cnj-005-plot-gamma-11-2}
: Plot the $\text{Gamma}(11, 2)$ prior
:::
::::

::: my-r-code-container
```{r}
#| label: fig-005-plot-gamma-11-2
#| fig-cap: The pdf of a $\text{Gamma}(11, 2)$ prior for $\lambda$, the daily rate of fraud risk calls.
#| fig-height: 3

bayesrules::plot_gamma(11, 2)
```
:::
::::::

But in that case the mean would change to 5.5 ($E = s/r = 11/2 = 5.5$).
It is better to take the mean as central tendency in a heavily skewed
distribution because it would be a better measure to catch the typical
property of the distribution than the mode.
:::::::

### Gamma-Poisson conjugacy

#### Bayesian model

As we discussed at the start of this chapter, conjugate families can
come in handy. Fortunately for us, using a Gamma prior for a rate
parameter $\lambda$ and a Poisson model for corresponding count data $Y$
is another example of a conjugate family. This means that, *spoiler*,
the posterior model for $\lambda$ will also have a Gamma model with
*updated* parameters. We’ll state and prove this in the general setting
before applying the results to our phone call situation.

:::::: my-theorem
:::: my-theorem-header
::: {#thm-005-gamma-poisson-model}
: The Gamma-Poisson Bayesian model
:::
::::

::: my-theorem-container
Let $\lambda > 0$ be an unknown rate parameter and
$(Y_1, Y_2, . . . , Y_n)$ be an independent $\text{Pois}(\lambda)$
sample. The Gamma-Poisson Bayesian model complements the Poisson
structure of data $Y$ with a Gamma prior on $\lambda$:

$$
\begin{align*}
Y_i \mid &\lambda \overset{\text{ind}}{\sim} \text{Pois}(\lambda) \\
&\lambda \sim \text{Gamma}(s, r)
\end{align*}
$$

Upon observing data $\overrightarrow{\rm y} = (y_1, y_2, . . . , y_n)$,
the posterior model of $\lambda$ is also a Gamma with updated
parameters:

$$
\lambda \mid \overrightarrow{\rm y} \sim \text{Gamma}(s + \sum y_i, r + n)
$$ {#eq-005-gamma-poisson-model}
:::
::::::

#### Prove conjugacy

Let’s prove this result. In general, recall that the posterior pdf of
$\lambda$ is proportional to the product of the prior pdf and likelihood
function defined by @eq-005-gamma-model and
@eq-005-poisson-likelihood-function, respectively:

$$
f (\lambda \mid \overrightarrow{\rm y}) \propto f (\lambda)L(\lambda \mid \overrightarrow{\rm y}) =  \frac{r^s}{\Gamma(s)}\lambda^{s−1}e^{−rλ} \cdot  \frac{\lambda^{\sum y_i} e{−n\lambda}} {\prod yi!} \text{ for } \lambda > 0
$$ Next, remember that any non-$\lambda$ multiplicative constant in the
above equation can be “proportional-ed” out. Thus, boiling the prior pdf
and likelihood function down to their kernels, we get

$$
\begin{align*}
f (\lambda \mid \overrightarrow{\rm y}) &\propto \lambda^{s−1}e^{−r\lambda} \cdot \lambda^{\sum y_i} e^{−n\lambda} \\
&= \lambda^{s+\sum y_i-1}e^{-(r+n)\lambda}
\end{align*}
$$

where the final line follows by combining like terms. What we’re left
with here is the *kernel* of the posterior pdf. This particular kernel
corresponds to the pdf of a Gamma model in @eq-005-gamma-model, with
shape parameter $s + \sum y_i$ and rate parameter $r + n$. Thus, we’ve
proven that

$$
\lambda \mid \overrightarrow{\rm y} \sim \text{Gamma}(s + \sum y_i, r + n)
$$

#### Likelihood

Let’s apply this result to our fraud risk calls. There we have a
$\text{Gamma}(10,2)$ prior for $\lambda$, the daily rate of calls.
Further, on four separate days in the second week of August, we received
$\overrightarrow{\rm y} = (y_1, y_2, y_3, y_4) = (6, 2, 2, 1)$ such
calls. Thus, we have a sample of $n = 4$ data points with a total of
$11$ fraud risk calls and an *average* of $2.75$ phone calls per day:

$$
\sum_{i=1}^{4}y_i = 6 + 2 + 2 + 1 = 11 \text{ and } \overline{\rm y} = \frac{\sum_{i=1}^{4} y_i}{4} = 2.75
$$ Plugging this data into @eq-005-poisson-likelihood-function, the
resulting Poisson likelihood function of $\lambda$ is

$$
L(\lambda \mid \overrightarrow{\rm y}) =  \frac{\lambda^{11}e^{−4λ}}{  6! \times 2! \times 2! \times× 1!} \propto \lambda^{11}e^{−4λ} \text { for } \lambda > 0
$$

We visualize a portion of $L(\lambda \mid \overrightarrow{\rm y})$ for
$\lambda$ between $0$ and $10$ using the
`bayesrules::plot_poisson_likelihood()` function. Here, `y` is the
vector of data values and `lambda_upper_bound` is the maximum value of
$\lambda$ to view on the x-axis. (Why can’t we visualize the whole
likelihood? Because $\lambda \in (0, \infty)$ and this book would be
pretty expensive if we had infinite pages.)

:::::: my-r-code
:::: my-r-code-header
::: {#cnj-005-plot-poisson-likelihood-fraud-calls}
: Plot Poisson likelihood of fraud calls
:::
::::

::: my-r-code-container
```{r}
#| label: fig-005-plot-poisson-likelihood-fraud-calls
#| fig-cap: The likelihood function of $\lambda$, the daily rate of fraud risk calls, given a four-day sample of phone call data.
#| fig-height: 3

bayesrules::plot_poisson_likelihood(c(6, 2, 2, 1), lambda_upper_bound = 10)
```
:::
::::::

The punchline is this: Underlying *rates* $\lambda$ of between one to
five fraud risk calls per day are consistent with our phone call data.
And across this spectrum, rates near $2.75$ are the *most* compatible
with this data. This makes sense. The Poisson data model assumes that
$\lambda$ is the underlying average daily phone call count,
$E(Y_i \mid \lambda) = \lambda$. As such, we’re *most* likely to observe
a sample with an average daily phone call rate of $\overline{y} = 2.75$
when the underlying rate $\lambda$ is also $2.75$.

Combining these observations with our $\text{Gamma}(10,2)$ prior model
of $\lambda$, it follows from @eq-005-gamma-poisson-model that the
posterior model of $\lambda$ is a Gamma with an updated shape parameter
of $21 (s + \sum y_i = 10 + 11)$ and rate parameter of
$6 (r + n = 2 + 4)$:

$$
\begin{align*}
\lambda \mid \overrightarrow{\rm y} &\sim \text{Gamma}(s + \sum y_i, r + n) = \\
&\sim \text{Gamma}(10 + 11, 2 + 4) = \\
&\sim \text{Gamma}(21, 6)
\end{align*}
$$ We can visualize the `r glossary("Priorx", "prior")` pdf, scaled
`r glossary("Likelihood-Function", "likelihood function")`, and
`r glossary("posteriorx", "posterior")` pdf for $\lambda$ all in a
single plot with the `bayesrules::plot_gamma_poisson()` function. How
magical. For this function to work, we must specify a few things:

-   the prior `shape` hyperparameter and
-   the `rate` hyperparameter as well as the information from our data,
-   the observed total number of phone calls `sum_y` ($\sum y_i$) and
    the
-   the sample size `n`:

:::::: my-r-code
:::: my-r-code-header
::: {#cnj-005-gamma-poisson-fraud-calls}
: Gamma-Poisson model of fraud risk calls,
:::
::::

::: my-r-code-container
```{r}
#| label: fig-005-gamma-poisson-fraud-calls
#| fig-cap: The Gamma-Poisson model of $\lambda$, the daily rate of fraud risk calls.
#| fig-height: 3

bayesrules::plot_gamma_poisson(shape = 10, rate = 2, sum_y = 11, n = 4)
```
:::
::::::

#### Posterior

Our posterior notion about the daily rate of fraud calls is, of course,
a *compromise* between our vague prior and the observed phone call data.
Since our prior notion was quite variable in comparison to the strength
in our sample data, the posterior model of $\lambda$ is more in sync
with the data. Specifically, utilizing the properties of the
$\text{Gamma}(10,2)$ prior and $\text{Gamma}(21,6)$ posterior as defined
by @eq-005-gamma-central-measures, notice that our posterior
understanding of the typical daily rate of phone calls dropped from $5$
to $3.5$ per day:

$$
\text{E}(\lambda) =  \frac{10}{2} = 5 \text{ and } \text{E}(\lambda \mid \overrightarrow{\rm y}) =  \frac{21}{6} = 3.5
$$ Though a compromise between the prior mean and data mean, this
posterior mean is *closer* to the data mean of $\overline{y} = 2.75$
calls per day.

::: {#tip-005-posterior-between-prior-and-data .callout-tip}
###### Posterior has to be between prior mean and data mean

The posterior mean will always be between the prior mean and the data
mean. If your posterior mean falls outside that range, it indicates that
you made an error and should retrace some steps.
:::

Further, with the additional information about $\lambda$ from the data,
the variability in our understanding of $\lambda$ drops by more than
half, from a standard deviation of $1.581$ to $0.764$ calls per day:

$$
\text{SD}(\lambda) =  \sqrt{\frac{10}{2^2}} \approx 1.581 \text{ and } \text{SD}(\lambda \mid \overrightarrow{\rm y}) =  \sqrt\frac{21}{6^2} \approx  0.764.
$$ The convenient `bayesrules::summarize_gamma_poisson()` function ,
which uses the same arguments as `bayesrules::plot_gamma_poisson()`,
helps us contrast the prior and posterior models and confirms the
results above:

:::::: my-r-code
:::: my-r-code-header
::: {#cnj-005-summarize-gamma-poisson-fraud-calls}
: Summarize Gamma-Poisson model for fraud risk calls
:::
::::

::: my-r-code-container
```{r}
#| label: tbl-005-summarize-gamma-poisson-fraud-calls
#| tbl-cap: Summarize Gamma-Poisson model for fraud risk calls

bayesrules::summarize_gamma_poisson(
  shape = 10, 
  rate = 2, 
  sum_y = 11, 
  n = 4) |> 
  knitr::kable(digits = 4)
```
:::
::::::

## Normal-normal conjugate family {#sec-005-normal-normal-conjugacy}

We now have two conjugate families in our toolkit: the Beta-Binomial and
the Gamma-Poisson. But many more conjugate families exist!

:::::: my-resource
:::: my-resource-header
::: {#lem-005-prior-compendium}
: Prior Compendium
:::
::::

::: my-resource-container
Fink, D. (1997). A Compendium of Conjugate Priors. 47 pages.
<https://www.johndcook.com/CompendiumOfConjugatePriors.pdf>

This report reviews conjugate priors and priors closed under sampling
for a variety of data generating processes where the prior distributions
are univariate, bivariate, and multivariate. The effects of
transformations on conjugate prior relationships are considered and
cases where conjugate prior relationships can be applied under
transformations are identified Univariate and bivariate prior
relationships are verified using Monte Carlo methods. [@fink-1997]
:::
::::::

It’s impossible to cover them all, but there is a third conjugate family
that’s especially helpful to know: the
`r glossary("Normal-Normal conjugate family", "Normal-Normal")`.

Consider a data story. As scientists learn more about brain health, the
dangers of concussions (hence of activities in which participants
sustain repeated concussions) are gaining greater attention
[@bachynski-2019]. Among all people who have a history of concussions,
we are interested in $\mu$, the average volume (in cubic centimeters) of
a specific part of the brain: the hippocampus. Though we don’t have
prior information about this group in particular,
[Wikipedia](https://en.wikipedia.org/wiki/Hippocampus) tells us that
among the general population of human adults, both halves of the
hippocampus have a volume between 3.0 and 3.5 cubic centimeters. Thus,
the *total* hippocampal volume of *both* sides of the brain is between 6
and 7 $\text{cm}^3$. Using this as a starting point, we’ll assume that
the mean hippocampal volume among people with a history of concussions,
$\mu$, is also somewhere between 6 and 7 $\text{cm}^3$, with an average
of 6.5. We’ll balance this prior understanding with data on the
hippocampal volumes of $n = 25$ subjects, $(Y_1, Y_2, . . . , Y_n)$,
using the **Normal-Normal Bayesian model**.

### The Normal data model

#### Normal formulas

Again, the spirit of our Bayesian analysis starts with our prior
understanding of $\mu$. Yet the specification of an appropriate prior
model structure for $\mu$ (which we can then tune) can be guided by
first identifying a model for the dependence of our data $Y_i$ upon
$\mu$. Since hippocampal volumes $Y_i$ are measured on a continuous
scale, there are many possible common models of the variability in $Y_i$
from person to person: `r glossary("Beta-Model", "Beta")`,
`r glossary("Exponential-Model", "Exponential")`,
`r glossary("Gamma-Model", "Gamma")`,
`r glossary("Normal-Model", "Normal")`,
`r glossary("F-Distribution-Model", "F")`, etc. From this list, we can
immediately eliminate the Beta model – it assumes that $Y_i \in [0, 1]$,
whereas hippocampal volumes tend to be around 6.5 $\text{cm}^3$. Among
the remaining options, the Normal model is quite *reasonable* –
biological measurements like hippocampal volume are often symmetrically
or *Normally* distributed around some global average, here $\mu$.

:::::: my-theorem
:::: my-theorem-header
::: {#thm-005-normal-model}
: The Normal model
:::
::::

::: my-theorem-container
Let $Y$ be a continuous random variable which can take any value between
$−\infty$ and $\infty$, i.e., $Y \in (−\infty, \infty)$. Then the
variability in $Y$ might be well represented by a Normal model with
**mean parameter** $\mu \in (−\infty, \infty)$ and **standard deviation
parameter** $\sigma > 0$:

$$
Y \sim N (\mu, \sigma^2)
$$ 


The Normal model is specified by continuous pdf

$$
f(y) =  \frac{1}{\sqrt{2\pi\sigma^2}} exp \left[ −  \frac{(y − \mu)^2}{2\sigma^2} \right] \text{ for } y \in (−\infty, \infty)
$$ {#eq-005-normal-model}

and has the following features:

$$
\begin{align*}
\text{E}(Y) &= \text{Mode}(Y) = \mu \\
\text{Var}(Y) &= \sigma^2 \\
\text{SD}(Y) &= \sigma
\end{align*}
$$ {#eq-005-normal-model-features}

Further, $\sigma$ provides a sense of scale for $Y$ . Roughly 95% of $Y$
values will be within $2$ standard deviations of $\mu$:

$$
\mu \pm 2\sigma
$$ {#eq-005-95-sd}
:::
::::::

#### Normal illustrations

:::::: my-r-code
:::: my-r-code-header
::: {#cnj-005-normal-illustrations}
: Illustration of Normal `r glossary("pdf", "pdfs")`
:::
::::

::: my-r-code-container
```{r}
#| label: fig-005-normal-illustrations
#| fig-cap: Normal pdfs with varying mean and standard deviation parameters.


# Create data frame with all combinations
data <- base::expand.grid(
  x = seq(0, 10, length.out = 100),  # More points for smooth density curve
  mean = c(2, 2, 4),
  sd = c(0.5, 1, 2)
) |> 
  dplyr::mutate(
    probability = stats::dnorm(x = x, mean = mean, sd = sd),
    prob_label = base::paste0("N(", mean, ",", sd, "\u00B2", ")")
  )

# Create density plot with facets
ggplot2::ggplot(data, ggplot2::aes(x = x, y = probability)) +
  ggplot2::geom_line(color = "black", linewidth = 0.8) +
  ggplot2::facet_wrap(~ prob_label, ncol = 3) +
  ggplot2::scale_x_continuous(breaks = base::seq(0, 12, by = 2)) +
  ggplot2::labs(
    x = "y",
    y = base::expression("f (y)")
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.spacing = grid::unit(1.5, "lines"),
    strip.background = ggplot2::element_rect(fill = "gray90", color = "gray50"),
    strip.text = ggplot2::element_text(face = "bold", size = 11)
  )
```
:::
::::::

@fig-005-normal-illustrations illustrates the Normal model under a
variety of mean and standard deviation parameter values, $\mu$ and
$\sigma$. No matter the parameters, the Normal model is bell-shaped and
symmetric around $\mu$ – thus as $\mu$ gets larger, the model shifts to
the right along with it. Further, $\sigma$ controls the variability of
the Normal model – as $\sigma$ gets larger, the model becomes more
spread out. Finally, though a Normal variable $Y$ can technically range
from $−\infty$ to $\infty$, the Normal model assigns negligible
plausibility to $Y$ values that are more than $3$ standard deviations
$\sigma$ from the mean $\mu$. To play around some more, you can plot
Normal models using the `bayesrules::plot_normal()` function.

Returning to our brain analysis, we can reasonably assume that the
hippocampal volumes of our $n = 25$ subjects, $(Y_1, Y_2, . . . , Y_n)$,
are *independent* and Normally distributed around a mean volume $\mu$
with standard deviation $\sigma$. Further, to keep our focus on $\mu$,
we’ll assume throughout our analysis that the standard deviation is
*known* to be $\mu = 0.5 \text{cm}^3$[^005-conjugate-families-1]. This
choice of $\sigma$ suggests that most people have hippocampal volumes
within $2\sigma = 1 \text{cm}^3$ of the average. Thus, the dependence of
$Y_i$ on the *unknown* mean $\mu$ is:

[^005-conjugate-families-1]: It is a silly assumption that we know
    $\sigma$ but don’t know $\mu$. Later we learn to relax this
    assumption.
    
$$
Y_i \mid \mu \sim N (\mu, \sigma^2)
$$

::: {.callout-warning #wrn-005-reasonable-not-perfect}
###### Reasonable doesn't mean perfect

Though we’ll later see that our hippocampal volume data does exhibit Normal behavior, the Normal model *technically* assumes that each subject’s hippocampal volume can range from $-\infty$ to $\infty$. However, we’re not too worried about this incorrect assumption here. Per our earlier discussion of @fig-005-normal-illustrations, the Normal model will put negligible weight on unreasonable values of hippocampal volume. In general, not letting *perfect* be the enemy of *good* will be a theme throughout this book (mainly because there is no perfect).
:::

#### Joint pdf

Accordingly, the `r glossary("Joint_Probability_Density_Function", "joint pdf")` which describes the collective randomness in our $n = 25$ subjects’ hippocampal volumes, $(Y_1, Y_2, . . . , Y_n)$, is the product of the unique Normal pdfs $f (y_i \mid \mu)$ defined by @eq-005-normal-model,

$$
f(\overrightarrow{\rm y} \mid \mu) = \prod_{i = 1}^{n} f(y_i \mid \mu) = \prod_{i = 1}^{n} \frac{1}{\sqrt{2\pi\sigma^2}} exp \left[ −  \frac{(y − μ)^2}{2\sigma^2} \right]
$$

#### Likelihood function

Once we observe our sample data $\vec{y}$, we can flip the joint pdf on its head to obtain the Normal likelihood function of $\mu$, $L(\mu \mid \vec{y}) = f (\vec{y} \mid \mu)$. Remembering that we’re assuming $\sigma$ is a known constant, we can simplify the likelihood up to a proportionality constant by dropping the terms that don’t depend upon $\mu$. Then for $\mu \in (−\infty, \infty)$,

$$
L(\mu \mid \vec{y}) = \propto \prod_{i = 1}^{n} \text{exp} \left[ −  \frac{(y − μ)^2}{2\sigma^2} \right] = \text{exp} \left[ - \frac{\sum_{i = 1}^{n}(y_i - \mu)^2}{2\sigma^2}\right]
$$

Through a bit more rearranging (which we encourage you to verify if, like us, you enjoy algebra), we can make this even easier to digest by using the sample mean $\overline{y}$ and sample size $n$ to summarize our data values:

$$
L(\mu \mid \vec{y}) \propto \exp\left[-\frac{\sum_{i=1}^{n}(y_i - \mu)^2}{2\sigma^2 / n}\right]
$$ {#eq-005-normal-likelihood}


Don’t forget the whole point of this exercise! Specifying a model for the data along with its corresponding likelihood function provides the tools we’ll need to assess the compatibility of our data $\overline(y)$ with different values of $\mu$ (once we actually collect that data).


::: {.callout-note #nte-005-rearranging-normal-likelihood}
###### Rearranging likelihood function


The content of this note was the result of the `r glossary("LLM")` `r glossary("Qwen-14b")`. I have it intentionally not edited, because I wanted to show the difference to the book result.

::: {#big-text style="font-size: 125%"}
**Step 0: Starting point**
:::


Let's simplify the likelihood function using the **sample mean** $\overline{y}$ and **sample size** $n$, and show how to arrive at the desired form:

$$
L(\mu \mid \vec{y}) \propto \exp\left(-\frac{\sum_{i=1}^{n}(y_i - \mu)^2}{2\sigma^2 / n}\right)
$$

We start from the original expression:

$$
L(\mu \mid \vec{y}) \propto \prod_{i=1}^{n} \exp\left(-\frac{(y_i - \mu)^2}{2\sigma^2}\right) = \exp\left(-\frac{1}{2\sigma^2} \sum_{i=1}^{n}(y_i - \mu)^2\right)
$$

---

::: {#big-text style="font-size: 125%"}
**Step 1: Use the identity for sum of squared deviations**
:::


We know that:
$$
\sum_{i=1}^{n}(y_i - \mu)^2 = \sum_{i=1}^{n}(y_i - \overline{y} + \overline{y} - \mu)^2
$$

Expand:
$$
= \sum_{i=1}^{n} \left[(y_i - \overline{y}) + (\overline{y} - \mu)\right]^2 = \sum_{i=1}^{n} \left[(y_i - \overline{y})^2 + 2(y_i - \overline{y})(\overline{y} - \mu) + (\overline{y} - \mu)^2 \right]
$$

Split the sum:
$$
= \sum_{i=1}^{n}(y_i - \overline{y})^2 + 2(\overline{y} - \mu)\sum_{i=1}^{n}(y_i - \overline{y}) + n(\overline{y} - \mu)^2
$$

But $\sum_{i=1}^{n}(y_i - \overline{y}) = 0$, so the middle term vanishes:

$$
\sum_{i=1}^{n}(y_i - \mu)^2 = \sum_{i=1}^{n}(y_i - \overline{y})^2 + n(\overline{y} - \mu)^2
$$

---

::: {#big-text style="font-size: 125%"}
**Step 2: Plug into the likelihood**
:::


Now substitute into the exponent:

$$
\exp\left(-\frac{1}{2\sigma^2} \left[ \sum_{i=1}^{n}(y_i - \overline{y})^2 + n(\overline{y} - \mu)^2 \right] \right)
$$

This splits into:
$$
= \exp\left(-\frac{1}{2\sigma^2} \sum_{i=1}^{n}(y_i - \overline{y})^2 \right) \cdot \exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right)
$$

The first term does **not** depend on $\mu$, so it can be ignored when maximizing the likelihood (since it's constant with respect to $\mu$).

So:
$$
L(\mu \mid \vec{y}) \propto \exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right)
$$

---

::: {#big-text style="font-size: 125%"}
**Step 3: Rewrite in the desired form**
:::


We now manipulate the exponent to match the target form:

$$
\exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right) = \exp\left(-\frac{\sum_{i=1}^{n}(y_i - \mu)^2}{2\sigma^2 / n} \right)
$$

Wait — this is **not** correct as written. Let's check the algebra.

We have:
$$
\exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right)
$$

But we want:
$$
\exp\left(-\frac{\sum_{i=1}^{n}(y_i - \mu)^2}{2\sigma^2 / n} \right)
$$

Let’s compute the exponent in the target form:

$$
\frac{\sum_{i=1}^{n}(y_i - \mu)^2}{2\sigma^2 / n} = \frac{n}{2\sigma^2} \sum_{i=1}^{n}(y_i - \mu)^2
$$

So the exponent becomes:
$$
- \frac{n}{2\sigma^2} \sum_{i=1}^{n}(y_i - \mu)^2
$$

But from earlier, we have:
$$
\exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right)
$$

So unless $\sum (y_i - \mu)^2 = n(\overline{y} - \mu)^2$, which is **not** true, the two expressions are **not** equal.

Wait — this suggests a **mistake** in the target form.

Let’s double-check.

---

**(a) Correct Interpretation**

The **correct** simplified form is:

$$
L(\mu \mid \vec{y}) \propto \exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right)
$$

This is **not** the same as:

$$
\exp\left(-\frac{\sum (y_i - \mu)^2}{2\sigma^2 / n} \right) = \exp\left(-\frac{n}{2\sigma^2} \sum (y_i - \mu)^2 \right)
$$

Which is **much larger** than the true likelihood unless all $y_i = \mu$.

So the **target expression** you wrote:

$$
L(\mu \mid \vec{y}) \propto \exp\left[-\frac{\sum_{i=1}^{n}(y_i - \mu)^2}{2\sigma^2 / n}\right]
$$

is **incorrect** — it's off by a factor of $n$ in the denominator.

---


**(b) Corrected Version**



The **correct** simplification using $\overline{y}$ and $n$ is:

$$
L(\mu \mid \vec{y}) \propto \exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right)
$$

This is the **correct** form.

---

**(c) Why the target form is wrong**


The expression:

$$
\exp\left(-\frac{\sum (y_i - \mu)^2}{2\sigma^2 / n} \right) = \exp\left(-\frac{n}{2\sigma^2} \sum (y_i - \mu)^2 \right)
$$

is **not** proportional to the original likelihood — it overweights the sum of squares by a factor of $n$, which changes the shape.

---

::: {#big-text style="font-size: 125%"}
**Step 4: Final Answer**
:::



The correct simplified form using $\overline{y}$ and $n$ is:

$$
\boxed{L(\mu \mid \vec{y}) \propto \exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right)}
$$ {#eq-005-normal-likelihood-function}

This shows that the **maximum likelihood estimate** of $\mu$ is $\overline{y}$, and the likelihood is centered around it with variance $\sigma^2/n$.

---


::: {#big-text style="font-size: 150%"}
**Summary of Steps**
:::


1. Start with the product of normal densities.
2. Combine into a single exponential.
3. Use the identity: $\sum (y_i - \mu)^2 = \sum (y_i - \overline{y})^2 + n(\overline{y} - \mu)^2$.
4. The first term is constant in $\mu$, so it can be dropped.
5. The likelihood depends only on $(\overline{y} - \mu)^2$.
6. Final form: $\exp\left(-\frac{n(\overline{y} - \mu)^2}{2\sigma^2} \right)$.

This is the **correct** simplification.


:::

### Normal prior

With the likelihood in place, let’s formalize a prior model for $\mu$, the mean hippocampal volume among people that have a history of concussions. By the properties of the $Y_i \mid \mu \sim N(\mu, \sigma^2)$ data model, the Normal mean parameter $\mu$ can technically take any value between −∞ and ∞. Thus, a Normal prior for $\mu$, which is also defined for $\mu \in (−\infty, \infty)$, makes a reasonable choice. 

:::::{.my-theorem}
:::{.my-theorem-header}
:::::: {#thm-005-normal-prior}
: Normal prior
::::::
:::
::::{.my-theorem-container}

We’ll assume that $\mu$ itself is Normally distributed around some mean $\theta$ with standard deviation $\tau$ :

$$\mu \sim N (\theta, \tau^2)$$

where $\mu$ has prior pdf


$$
f(\mu) =  \frac{1}{\sqrt{2\pi\tau^2}} \text{exp} \left[ −  \frac{(\mu - \theta)^2}{2\tau^2} \right] \text{ for } \mu \in (−\infty, \infty)
$$ {#eq-005-normal-prior}



::::
:::::

Not only does the Normal prior assumption that $\mu \in (−\infty, \infty)$ match the same assumption of the Normal data model, we’ll prove below that this is a conjugate prior. You might anticipate this result from the fact that the likelihood function $L(\mu \mid \vec{y})$ in @eq-005-normal-likelihood and prior pdf $f(\mu)$ in @eq-005-normal-prior are both proportional to

$$
\text{exp} \left[− \frac{(μ − ∎)^2} {  2∎2} \right]
$$

with different ∎.

#### Tuning Normal prior

Using our understanding of a Normal model, we can now tune the prior hyperparameters $\theta$ and $\tau$ to reflect our prior understanding and uncertainty about the average hippocampal volume among people that have a history of concussions, $\mu$. 

- Based on [Wikipedia research](https://en.wikipedia.org/wiki/Hippocampus) that hippocampal volumes tend to be between $6$ and $7 \text{cm}^3$, we’ll set the Normal prior mean $\theta$ to the midpoint, $6.5$. 
- Further, we’ll set the Normal prior standard deviation to $\tau = 0.4$. In other words, by @eq-005-95-sd, we think there’s a 95% chance that $\mu$ is somewhere between $5.7$ and $7.3 \text{cm}^3 (6.5 \pm 2 \times 0.4)$. 

This range is wider, and hence more conservative, than what Wikipedia indicated. Our uncertainty here reflects the fact that we didn’t vet the Wikipedia sources, we aren’t confident that the features for the typical adult translates to people with a history of concussions, and we generally aren’t sure what’s going on here (i.e., we’re not brain experts). Putting this together, our tuned prior model for $\mu$ is:

$$\mu \sim N (6.5, 0.4^2)$$


:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-tuning-normal-prior}
: Tuning the Normal prior
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-tuning-normal-prior
#| fig-cap: A Normal prior model for μ, with mean 6.5 and standard deviation 0.4.
#| fig-height: 3

bayesrules::plot_normal(6.6, 0.4)

```

::::
:::::


### Normal-Normal conjugacy

#### Bayesian model

To obtain our posterior model of $\mu$ we must combine the information from our prior and our data. Again, we were clever to pick a Normal prior model – the Normal-Normal is another convenient conjugate family! Thus, the posterior model for $\mu$ will *also* be Normal with updated parameters that are informed by the prior and observed data.

:::::{.my-theorem}
:::{.my-theorem-header}
:::::: {#thm-005-normal-normal-bayesian-model}
: The Normal-Normal Bayesian model
::::::
:::
::::{.my-theorem-container}

Let $\mu \in (−\infty, \infty)$ be an unknown mean parameter and $(Y_1, Y_2, . . . , Y_n)$ be an independent $N (\mu, \sigma^2)$ sample where $\sigma$ is assumed to be known. The Normal-Normal Bayesian model complements the Normal data structure with a Normal prior on $\mu$:  

$$
\begin{align*}
Y_i \mid &\mu \overset{\text{ind}}{\sim} \mathcal{N} (\mu, \sigma^2) \\
&\mu \sim \mathcal{N} (\theta, \tau^2)
\end{align*}
$$ {#eq-005-normal-normal-model}

Upon observing data $\vec{y} = (y_1, y_2, . . . , y_n)$ with mean $\overline{y}$, the posterior model of $\mu$ is also Normal with updated parameters:

$$
\begin{align*}
\mu \mid \vec{y} \sim N \left(\theta \frac{\sigma^2}{n \tau^2 + \sigma^2} + \overline{y} \frac{n\tau^2}{n \tau^2 + \sigma^2}, \frac{\tau^2 \sigma^2}{n \tau^2 + \sigma^2} \right)
\end{align*}
$$ {#eq-005-normal-normal-posterior-model}
::::
:::::

Whooo, that is a mouthful! We provide an optional proof of this result in @sec-005-proving-normal-normal-conjugacy. 

::: {.callout-note #nte-005-norma-normal-posterior-model}

###### Reformulation of @eq-005-normal-normal-posterior-model
With the help of `r glossary("Claude Haiku")` I noticed during working on @exr-006-normal-normal-approximation a different formula for the Normal-Normal posterior model. What follow is the derivation from the formula above and the explication why the converted formula is of importance.

***

::: {#big-text style="font-size: 110%"}
**Reformulation of the mean part of the formula** 
:::

$$
\theta \frac{\sigma^2}{n \tau^2 + \sigma^2} + \overline{y} \frac{n\tau^2}{n \tau^2 + \sigma^2} = \frac{1}{\dfrac{n}{\sigma^2} + \dfrac{1}{\tau}} \cdot \left(\frac{\mu}{\sigma^2} + \frac{n \cdot \overline{y}}{\tau^2}\right)
$$


**Explain why $\large \theta \frac{\sigma^2}{n \tau^2 + \sigma^2} + \overline{y} \frac{n\tau^2}{n \tau^2 + \sigma^2}$ can be expressed as $\large \frac{1}{\frac{n}{\sigma^2} + \frac{1}{\tau}} \cdot \left(\frac{\mu}{\sigma^2} + \frac{n \cdot \overline{y}}{\tau^2}\right)$**

Starting with the left-hand side:

$$\mu \frac{\sigma^2}{n \tau^2 + \sigma^2} + \overline{y} \frac{n\tau^2}{n \tau^2 + \sigma^2}$$

Factor out the common denominator:

$$\frac{\mu \sigma^2 + \overline{y} n\tau^2}{n \tau^2 + \sigma^2}$$

Now, divide both numerator and denominator by $\sigma^2 \tau^2$:

$$\Large \frac{\frac{\mu \sigma^2 + \overline{y} n\tau^2}{\sigma^2 \tau^2}}{\frac{n \tau^2 + \sigma^2}{\sigma^2 \tau^2}}$$

Split the numerator into two fractions:

$$\Large \frac{\frac{\mu \sigma^2}{\sigma^2 \tau^2} + \frac{\overline{y} n\tau^2}{\sigma^2 \tau^2}}{\frac{n \tau^2 + \sigma^2}{\sigma^2 \tau^2}}$$

Simplify each term in the numerator:

$$\Large \frac{\frac{\mu}{\tau^2} + \frac{n\overline{y}}{\sigma^2}}{\frac{n \tau^2 + \sigma^2}{\sigma^2 \tau^2}}$$

Invert the complex fraction (divide by a fraction = multiply by its reciprocal):

$$\left(\frac{\mu}{\tau^2} + \frac{n\overline{y}}{\sigma^2}\right) \cdot \frac{\sigma^2 \tau^2}{n \tau^2 + \sigma^2}$$

Rewrite as:

$$\frac{1}{\dfrac{n \tau^2 + \sigma^2}{\sigma^2 \tau^2}} \cdot \left(\frac{\mu}{\tau^2} + \frac{n\overline{y}}{\sigma^2}\right)$$

Finally, split the denominator:

$$
\frac{1}{\dfrac{n}{\sigma^2} + \dfrac{1}{\tau^2}} \cdot \left(\frac{\mu}{\tau^2} + \frac{n\overline{y}}{\sigma^2}\right)
$$ {#eq-005-normal-normal-posterior-mean}

This is the right-hand side! ✓

The second form is the **precision-weighted average** representation of the posterior mean, where:

- The prior mean $\mu$ is weighted by data precision ($n/\sigma^2$)
- The sample mean $\overline{y}$ is weighted by prior precision ($1/\tau^2$)

**Key insight**: The key step is **dividing by $\sigma^2 \tau^2$**, which transforms the weighted average form into the precision-weighted form, revealing how the posterior mean combines prior and data information through their precisions.



***

::: {#big-text style="font-size: 110%"}
**Reformulation of the variance part of the formula** 
:::


$$\frac{\tau^2 \sigma^2}{n \tau^2 + \sigma^2} = \large \frac{1}{\frac{n}{\sigma^2} + \frac{1}{\tau^2}}$$

**Explain why $\large \frac{\tau^2 \sigma^2}{n \tau^2 + \sigma^2}$ can be expressed as $\large \frac{1}{\frac{n}{\sigma^2} + \frac{1}{\tau^2}}$**

Starting with the left-hand side:

$$\frac{\tau^2 \sigma^2}{n \tau^2 + \sigma^2}$$

Divide both numerator and denominator by $\tau^2 \sigma^2$:

$$ \dfrac{\dfrac{\tau^2 \sigma^2}{\tau^2 \sigma^2}}{\dfrac{n \tau^2 + \sigma^2}{\tau^2 \sigma^2}}$$

Simplify the numerator:

$$\frac{1}{\dfrac{n \tau^2 + \sigma^2}{\tau^2 \sigma^2}}$$

Split the denominator into two fractions:

$$\frac{1}{\dfrac{n \tau^2}{\tau^2 \sigma^2} + \dfrac{\sigma^2}{\tau^2 \sigma^2}}$$

Simplify each fraction in the denominator:

$$
\frac{1}{\dfrac{n}{\sigma^2} + \dfrac{1}{\tau^2}}
$$ {#eq-005-normal-normal-posterior-variance}

This is exactly the right-hand side! ✓


**Why @eq-005-normal-normal-posterior-variance as the different posterior variance form for the Normal-Normal model in Bayesian statistics matters**

This form highlights a key insight: **the posterior variance is the reciprocal of the sum of precisions**.

- **Prior precision**: $\large \frac{1}{\tau^2}$
- **Data precision**: $\large \frac{n}{\sigma^2}$
- **Posterior precision**: $\large \frac{n}{\sigma^2} + \frac{1}{\tau^2}$
- **Posterior variance**: $\large \frac{1}{\text{posterior precision}} = \frac{\tau^2 \sigma^2}{n \tau^2 + \sigma^2}$

The key insight is that **dividing both numerator and denominator by $\tau^2 \sigma^2$** transforms the expression into a form where the denominator becomes a sum of precisions, revealing the underlying structure of how prior and data information combine.

:::



Even without that proof, we can observe the balance that the Normal posterior in @eq-005-normal-normal-posterior-model strikes between the prior and the data. 

1. First, the posterior mean is a **weighted average** of the prior mean $E(\mu) = \theta$ and the sample mean $\overline{y}$. 
2. Second, the posterior variance is informed by the prior variability $\tau$ and variability in the data $\sigma$. Both are impacted by sample size $n$. 

a. First, as $n$ increases, the posterior mean places less weight on the prior mean and more weight on sample mean $\overline{y}$:

$$
\frac{\sigma^2}{n \tau^2 + \sigma^2} \to 0 \text{ and } \frac{n\tau^2}{n \tau^2 + \sigma^2} \to 1
$$

b. Further, as $n$ increases, the posterior variance decreases:
  
$$ 
\frac{\tau^2 \sigma^2}{n \tau^2 + \sigma^2} \to 0
$$

That is, the more and more data we have, our posterior certainty about $\mu$ increases and becomes more in sync with the data.


#### Apply result

Let’s apply and examine this result in our analysis of $\mu$, the average hippocampal volume among people that have a history of concussions. 

1. We’ve already built our prior model of $\mu$, $\mu \sim N (6.5, 0.4^2)$. 
2. Next, consider some data. The `football` data in {**bayesrules**}, a subset of the `FootballBrain` data in the {**Lock5Data**} package [@Lock5Data], includes results for a cross-sectional study of hippocampal volumes among 75 subjects [@singh-2014]: 

- 25 collegiate football players with a history of concussions (`fb_concuss`), 
- 25 collegiate football players that do not have a history of concussions (`fb_no_concuss`), and 
- 25 control subjects. 

For our analysis, we’ll focus on the $n = 25$ subjects with a history of concussions (`fb_concuss`):

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-load-football-data}
: Load and skim the `football` data from {**bayesrules**}
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: tbl-005-load-skim-football-data
#| tbl-cap: "Load an skim football data"
#| lst-label: lst-005-load-skim-football-data
#| lst-cap: Load, display `football` data and compute mean
#| results: hold

# Load the data
data(football, package = "bayesrules")
skimr::skim(football)

concussion_subjects <- football |> 
  dplyr::filter(group == "fb_concuss")

mean_hyppocampal_volume <- concussion_subjects |> 
  dplyr::summarize(mean(volume))
```
```{r}
#| label: tbl-load-football-data
#| tbl-cap: "Show 8 random rows with the first and last row."

my_glance_data(football) |> 
  knitr::kable()
```

::::
:::::

**Football players with a concussion history have an average hippocampal volume of y = `r mean_hyppocampal_volume` $\text{cm}^3$.**

Further, the hippocampal volumes appear to vary normally from subject to subject, ranging from roughly $4.5$ to $7 \text{cm}^3$. That is, our assumed Normal data model about individual hippocampal volumes, $Y_i \mid \mu \sim N (\mu, \sigma^2)$ with an assumed standard deviation of $\sigma = 0.5$, seems reasonable:


:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-hippocampal-volumes}
: Density plot of the hippocampal volumes
::::::
:::
::::{.my-r-code-container}

```{r}
#| label: fig-005-hippocampal-volumes
#| fig-cap: A density plot of the hippocampal volumes (in cubic centimeters) among 25 subjects that have experienced concussions.
#| fig-height: 3

ggplot2::ggplot(concussion_subjects, ggplot2::aes(x = volume)) + 
  ggplot2::geom_density()

```

::::
:::::

#### Likelihood

Plugging this information from the data ($n = 25, y = 5.735, \text{ and } \sigma = 0.5$) into @eq-005-normal-likelihood-function defines the Normal likelihood function of $\mu$:

$$
\begin{align*}
&L(\mu \mid \vec{y}) \propto \exp\left[-\frac{\sum_{i=1}^{n}(y_i - \mu)^2}{2\sigma^2 / n}\right] \\
&L(\mu \mid \vec{y}) \propto \exp\left(-\frac{(5.735 - \mu)^2}{2 (0.5 / 25)} \right) \text{ for } \mu \in (-\infty, \infty)
\end{align*}
$$ 


We plot this likelihood function using `bayesrules::plot_normal_likelihood()`, providing our observed volume data and data standard deviation $\sigma = 0.5$ (Figure 5.11). This likelihood illustrates the compatibility of our observed hippocampal data with different $\mu$ values. To this end, the hippocampal patterns observed in our data would most likely have arisen if the mean hippocampal volume across all people with a history of concussions, $\mu$, were between $5.3$ and $6.1 \text{cm}^3$. Further, we’re most likely to have observed a mean volume of $y = 5.735$ among our $25$ sample subjects if the underlying population mean $\mu$ were also $5.735$.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-normal-likelihood-function-hippocampal}
: Normal likelihood function for mean hippocampal volume $\mu$
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-normal-likelihood-function-hippocampal
#| fig-cap: The Normal likelihood function for mean hippocampal volume $\mu$
#| fig-height: 3

bayesrules::plot_normal_likelihood(y = concussion_subjects$volume, sigma = 0.5)
```

::::
:::::

#### Posterior

We now have all necessary pieces to plug into @eq-005-normal-normal-posterior-model, and hence to specify the posterior model of $\mu$:  

- our Normal prior model of $\mu$ had mean $theta = 6.5$ and standard deviation $\tau = 0.4$; 
- our $n = 25$ sample subjects had a sample mean volume $\overline{y} = 5.735$;
- we assumed a known standard deviation among individual hippocampal volumes of $\sigma = 0.5$.

It follows that the posterior model of $\mu$ is:

$$
\begin{align*}
\mu \mid \vec{y} &\sim N \left(\theta \frac{\sigma^2}{n \tau^2 + \sigma^2} + \overline{y} \frac{n\tau^2}{n \tau^2 + \sigma^2}, \frac{\tau^2 \sigma^2}{n \tau^2 + \sigma^2} \right) \\
&\sim N \left(6.5 \frac{0.5^2}{(25 \cdot 0.4^2) + 0.5^2} + 5.735 \cdot \frac{25 \cdot 0.4^2}{(25 \cdot 0.4^2) + 0.5^2}, \frac{0.4^2 \cdot 0.5^2}{(25 \cdot 0.4^2) + 0.5^2} \right) \\
&\sim N \left(6.5 \frac{0.25}{(25 \cdot 0.16) + 0.25} + 5.735 \cdot \frac{25 \cdot 0.16}{(25 \cdot 0.16) + 0.25}, \frac{0.16 \cdot 0.25}{(25 \cdot 0.16) + 0.25} \right) \\
&\sim N \left(6.5 \frac{0.25}{4.25} + 5.735 \cdot \frac{4}{4 + 0.25}, \frac{0.04}{4 + 0.25} \right) \\
&\sim N \left(6.5 \cdot 0.05882353 + 5.735 \cdot \frac{4}{4.25}, \frac{0.04}{4.25} \right) \\
&\sim N \left(0.3823529 + 5.735 \cdot 0.9411765, 0.009411765 \right) \\
&\sim N \left(5.78, 0.009411765 \right) \\
&\sim \bf{N \left(5.78, 0.0094 \right)}
\end{align*}
$$ 
Standard deviation is $\sqrt{0.0094} = 0.097$.

where the posterior mean places roughly 94% of its weight on the data mean ($\overline{y} = 5.375$) and only 6% of its weight on the prior mean ($E(mu) = 6.5$):

$$
\text{E}(\mu \mid \vec{y}) = 6.5 \cdot 0.0588 + 5.735 \cdot 0.9412 = 5.78
$$

Bringing all of these pieces together, we plot and summarize our Normal-Normal analysis of $\mu$ using `plot_normal_normal()` and `summarize_normal_normal()` in the {**bayesrules**} package. Though a compromise between the prior and data, our posterior understanding of $\mu$ is more heavily influenced by the latter. In light of our data, we are much more certain about the mean hippocampal volume among people with a history of concussions, and believe that this figure is somewhere in the range from $5.586$ to $5.974 \text{cm}^3 (5.78 \pm 2 \times 0.097)$.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-plot-and-summarize-posterior}
: Plot and summarize the result of the Normal-Normal analysis for football players with concussion history
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-005-plot-normal-normal
#| fig-cap: The Normal-Normal model of μ, average hippocampal volume.

bayesrules::plot_normal_normal(
  mean = 6.5,
  sd = 0.4,
  sigma = 0.5,
  y_bar = 5.375,
  n = 25
)
```
:::

::: {.column width="50%"}
```{r}
#| label: tbl-005-summarize-normal-normal
#| tbl-cap: The Normal-Normal model of μ, average hippocampal volume.

bayesrules::summarize_normal_normal(
  mean = 6.5,
  sd = 0.4,
  sigma = 0.5,
  y_bar = 5.375,
  n = 25
) |> 
  knitr::kable(digits = 4)
```
:::
:::::

::::
:::::




### Optional: Proving Normal-Normal conjugacy (skipped) {#sec-005-proving-normal-normal-conjugacy}


## Why no simulation in this chapter?

As you may have gathered, we love some simulations. There is something reassuring about the reality check that a simulation can provide. Yet we are at a crossroads. The Gamma-Poisson and Normal-Normal models we’ve studied here are tough to simulate using the techniques we’ve learned thus far. 

:::::{.my-procedure}
:::{.my-procedure-header}
:::::: {#prp-005-steps-of-simulations}
: Steps of simulations
::::::
:::
::::{.my-procedure-container}
Letting $\theta$ represent some parameter of interest, recall the steps we’ve used for past simulations:  

1. Simulate, say, 10000 values of $\theta$ from the prior model. 
2. Simulate a set of sample data $Y$ from each simulated $\theta$ value. 
3. Filter out only those of the 10000 simulated sets of ($\theta$, $Y$ ) for which the simulated $Y$ data matches the data we actually observed. 
4. Use the remaining $\theta$ values to approximate the posterior of $\theta$.
::::
:::::


The issue with extending this simulation technique to the `r glossary("Gamma-Poisson Model", "Gamma-Poisson")` and `r glossary("Normal-Normal Conjugate Family", "Normal-Normal")` examples in this chapter comes with step 3. In both of our examples, we had a sample size greater than one, $(Y_1, Y_2, . . . , Y_n)$. Further, in the Normal-Normal example, our data values $Y_i$ are continuous. In both of these scenarios, it’s very likely that no simulated sets of ($\theta, Y$ ) will perfectly match our observed sample data. This is true even if we carry out millions of simulations in step 3. 

Spoiler alert! There are other ways to approximate the posterior model which we will learn in Unit 2.



## Critiques of conjugate family models

Before we end the chapter, we want to acknowledge that conjugate family models also have drawbacks. In particular:  

- A conjugate prior model isn’t always flexible enough to fit your prior understanding. For example, a Normal model is always unimodal and symmetric around the mean $\mu$. So if your prior understanding is not symmetric or is not unimodal, then the Normal prior might not be the best tool for the job. 
- Conjugate family models do not always allow you to have an entirely flat prior. While we can tune a flat Beta prior by setting $\alpha = \beta = 1$, neither the Normal nor Gamma priors (or any proper models with infinite support) can be tuned to be *totally* flat. The best we can do is tune the priors to have very high variance, so that they’re *almost* flat.

## Summary

In this chapter, you learned about conjugacy and applied it to a few different situations. Our main takeaways for this chapter are:  

- Using conjugate priors allows us to have easy-to-derive and readily interpretable posterior models. 
- The Beta-Binomial, Gamma-Poisson, and Normal-Normal conjugate families allow us to analyze data $Y$ in different scenarios. 
  - The `r glossary("Beta-Binomial-Model", "Beta-Binomial")` is convenient when our data $Y$ is the number of successes in a set of $n$ trials, 
  - the `r glossary("Gamma-Poisson Model", "Gamma-Poisson")` when $Y$ is a count with no upper limit, and 
  - the `r glossary("Normal-Normal Conjugate Family", "Normal-Normal")` when $Y$ is continuous. 
- We can use several functions from the bayesrules package to explore these conjugate families: `plot_poisson_likelihood()`, `plot_gamma()`, `plot_gamma_poisson()`, `summarize_gamma_poisson()`, `plot_normal()`, `plot_normal_normal()`, and `summarize_normal_normal()`.  

We hope that you now appreciate the utility of conjugate priors!


::: {.callout-note #nte-005-summary}
###### Comparison of the three conjugate families

Here's a comprehensive comparison of the three main conjugate families:

::: {#big-text style="font-size: 125%"}
**1. Beta-Binomial**
:::


::: {#big-text style="font-size: 110%"}
**Components** 
:::

- **Parameter**: π ∈ [0, 1] (proportion/probability)
- **Prior**: Beta(α, β)
- **Likelihood**: Binomial(n, π)
- **Posterior**: Beta(α + y, β + n - y)

::: {#big-text style="font-size: 110%"}
**Key Features** 
:::

- **Use case**: Modeling proportions, probabilities, success rates
- **Data type**: Count of successes in fixed number of trials
- **Prior hyperparameters**: 
  - α = prior successes
  - β = prior failures
- **Update rule**: Add observed successes to α, add failures to β

::: {#big-text style="font-size: 110%"}
**Interpretation** 
:::

- **Mean**: E(π) = α/(α + β)
- **Effective sample size**: α + β (strength of prior belief)
- The larger α + β relative to n, the more the prior dominates

---

::: {#big-text style="font-size: 125%"}
**2. Gamma-Poisson**
:::

::: {#big-text style="font-size: 110%"}
**Components** 
:::


- **Parameter**: λ > 0 (rate)
- **Prior**: Gamma(s, r)
- **Likelihood**: Poisson(λ)
- **Posterior**: Gamma(s + Σyᵢ, r + n)

::: {#big-text style="font-size: 110%"}
**Key Features** 
:::

- **Use case**: Modeling rates, waiting times, counts per unit time/space
- **Data type**: Count data (0, 1, 2, ...) with no upper limit
- **Prior hyperparameters**:
  - s = shape (prior "total count")
  - r = rate (prior "exposure time")
- **Update rule**: Add total observed counts to s, add number of observations to r

::: {#big-text style="font-size: 110%"}
**Interpretation** 
:::

- **Mean**: E(λ) = s/r
- **Variance**: Var(λ) = s/r²
- r represents prior sample size or exposure

---

::: {#big-text style="font-size: 125%"}
**3. Normal-Normal**
:::


::: {#big-text style="font-size: 110%"}
**Components** 
:::


- **Parameter**: μ ∈ ℝ (mean)
- **Prior**: N(μ₀, σ₀²)
- **Likelihood**: N(μ, σ²) with known σ²
- **Posterior**: N(μₙ, σₙ²) where:
  - μₙ = (μ₀/σ₀² + nȳ/σ²)/(1/σ₀² + n/σ²)
  - σₙ² = 1/(1/σ₀² + n/σ²)

::: {#big-text style="font-size: 110%"}
**Key Features** 
:::

- **Use case**: Modeling continuous measurements, averages
- **Data type**: Continuous real numbers
- **Prior hyperparameters**:
  - μ₀ = prior mean
  - σ₀² = prior variance (uncertainty about μ)
- **Update rule**: Precision-weighted average of prior mean and sample mean

::: {#big-text style="font-size: 110%"}
**Interpretation** 
:::

- **Posterior mean**: Weighted average based on precisions (1/variance)
- **Posterior variance**: Always smaller than prior variance
- More data → posterior mean closer to sample mean

---


::: {#big-text style="font-size: 125%"}
**Common Patterns Across All Three** 
:::


::: {#big-text style="font-size: 110%"}
**1. Simple Update Rules**
:::
All three have clean, interpretable posterior formulas:
- Prior hyperparameters + data summaries = posterior hyperparameters

::: {#big-text style="font-size: 110%"}
**2. Balance Between Prior and Data**
:::

- Small n → prior dominates
- Large n → data dominates
- The posterior is always a compromise

::: {#big-text style="font-size: 110%"}
**3. Computational Ease** 
:::
- No integration needed
- Closed-form posteriors
- Easy to calculate means, variances, credible intervals

::: {#big-text style="font-size: 110%"}
**4. Interpretable Hyperparameters** 
:::

- Can be thought of as "prior data"
- Beta: prior successes/failures
- Gamma: prior counts/exposure
- Normal: prior mean/uncertainty

---

::: {#big-text style="font-size: 125%"}
**When to Use Each** 
:::

**Beta-Binomial**: 
- Binary outcomes (yes/no, success/failure)
- Proportions bounded between 0 and 1
- Example: conversion rates, disease prevalence

**Gamma-Poisson**:
- Count data with no fixed upper limit
- Rates of occurrence
- Example: customer arrivals, website visits, phone calls

**Normal-Normal**:
- Continuous measurements
- Symmetric distributions
- Example: heights, test scores, measurement errors (when σ² is known)


:::


::: {.column-page-inset}

| Feature                   | Beta-Binomial                                                                                                                                     | Gamma-Poisson                                                                                                       | Normal-Normal                                                                                                                                                                   |
|:--------------------------|:--------------------------------------------------------------------------------------------------------------------------------------------------|:--------------------------------------------------------------------------------------------------------------------|:--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| **Parameter**             | π ∈ [0, 1]                                                                                                                                        | λ > 0 (rate)                                                                                                        | μ ∈ ℝ (mean)                                                                                                                                                                    |
| **Parameter space**       | [0, 1]                                                                                                                                            | (0, ∞)                                                                                                              | (-∞, ∞)                                                                                                                                                                         |
| **Parameter type**        | Proportion / Probability                                                                                                                          | Rate                                                                                                                | Mean                                                                                                                                                                            |
| **Data support**          | {0, 1, ..., n}                                                                                                                                    | {0, 1, 2, ...}                                                                                                      | ℝ (continuous)                                                                                                                                                                  |
| **Data type**             | Count of successes in<br>fixed number of trials                                                                                                   | Count data (0, 1, 2, …) <br>with no upper limit                                                                     | Continuous real numbers                                                                                                                                                         |
| **Sufficient statistics** | y, n                                                                                                                                              | Σyᵢ, n                                                                                                              | ȳ, n                                                                                                                                                                            |
| **Prior**                 | Beta(α, β)                                                                                                                                        | Gamma(s, r)                                                                                                         | N(μ₀, σ₀²)                                                                                                                                                                      |
| **Prior Hyperparameters** | α = prior successes,<br>β = prior failures                                                                                                        | s = shape (prior “total count”),<br>r = rate (prior “exposure time”)                                                | μ₀ = prior mean,<br>σ₀² = prior variance <br>(uncertainty about μ)                                                                                                              |
| **Likelihood**            | Binomial(n, π)                                                                                                                                    | Poisson(λ)                                                                                                          | N(μ, σ²) with known σ²                                                                                                                                                          |
| **Posterior**             | Beta(α + y, β + n - y)                                                                                                                            | Gamma(s + Σyᵢ, r + n)                                                                                               | N(μₙ, σₙ²) where:<br>μₙ = (μ₀/σ₀² + nȳ/σ²)/(1/σ₀² + n/σ²)<br>σₙ² = 1/(1/σ₀² + n/σ²)                                                                                             |
| **Use case**              | Modeling proportions, <br>probabilities, success rates,<br>polls                                                                                  | Modeling rates, waiting times, <br>event counts per unit in<br>time/space, arrivals                                 | Modeling continuous measurements,<br>heights, averages                                                                                                                          |
| **Update rule**           | Add observed successes to α, <br>add failures to β                                                                                                | Add total observed counts to s, <br>add number of observations to r                                                 | Precision-weighted average of <br>prior mean and sample mean                                                                                                                    |
| **Interpretation**        | Mean: E(π) = α/(α + β)<br>Effective sample size: α + β (strength of prior belief)<br>The larger α + β relative to n, the more the prior dominates | Mean: E(λ) = s/r<br>Variance: Var(λ) = s/r²<br>r represents prior sample size or exposure                           | Posterior mean: Weighted average based on precisions (1/variance)<br>Posterior variance: Always smaller than prior variance<br>More data → posterior mean closer to sample mean |
| **R function**            | plot_beta()<br>plot_beta_binomial()<br>plot_binomial_likelihood()<br>summarize_beta()<br>summarize_beta_binomial()                                | plot_gamma()<br>plot_gamma_poisson()<br>plot_poisson_likelihood()<br>summarize_gamma()<br>summarize_gamma_poisson() | plot_normal()<br>plot_normal_normal()<br>plot_normal_likelihood()<br><br>summarize_normal_normal()                                                                              |
: Side-by-side comparison of Beta-Binomial, Gamm-Poisson, and Normal-Normal  {#tbl-005-model-comparison} {tbl-colwidths="[10,30,30,30]"} {.striped .hover}





:::

## Exercises

::: {#wrn-005-exercises .callout-warning}
###### Take my solutions with caution!

Keep in mind, that I am not an expert but a learner. This means that the
presented solutions may have errors or are completely wrong.

For some of the exercises I have used `r glossary("LLMs")`, mostly the free `r glossary("Claude Haiku")`.

If you find errors please report to me at the [GitHub
repo](https://github.com/petzi53/bayes-rules) of this website.
:::

:::::{.my-resource}
:::{.my-resource-header}
:::::: {#lem-005-distribution-zoo}
: Distribution Zoo: Shiny app for a variety of distributions
::::::
:::
::::{.my-resource-container}


My first book I read on Bayesian statistics was "A Student’s Guide to Bayesian Statistics" [@lambert2018]. At the time of my reading (September 2023) it was too complex for me and I stopped reading. The book recommends a Shiny app to view distribution properties and access dynamically in R, Python, Matlab, Mathematica and Stan. ([Website](https://ben18785.shinyapps.io/distribution-zoo/), [GitHub Repo](https://github.com/ben18785/distribution-zoo)).

At the time of my reading I didn't understand the importance of this tool, but now with my greater knowledge it comes very handy! The app is more extensive and elaborated than the proposed Shiny app from [Chapter 3 of STAT 454 Bayesian Statistics](https://ajohns24.github.io/454_Fall_21/the-beta-binomial-model.html#shiny-apps). 


::::
:::::

### Practice Gamma-Poisson

#### Tuning a Gamma prior

##### Mode & mean {#sec-005-tune-gamma-prior-mean-mode}

:::::{.my-theorem}
:::{.my-theorem-header}
:::::: {#thm-005-tune-gamma-prior-mean-mode}
: How to tune a Gamma prior given mode and mean?
::::::
:::
::::{.my-theorem-container}

To find the appropriate Gamma(s, r) prior given a desired **mode** and **mean**, we use the known formulas for the Gamma distribution in @eq-005-gamma-central-measures:

::: {#big-text style="font-size: 125%"}
**🔍 Gamma Distribution Properties** 
:::

- **Mean** = $\frac{s}{r}$
- **Mode** = $\frac{s - 1}{r}$, for $s > 1$

Let:

- $\lambda$ = desired **mode**
- $\mu$ = desired **mean**

We solve the system of equations:

$$
\begin{align*}
\frac{s}{r} &= \mu \quad \text{(1)} \\
\frac{s - 1}{r} &= \lambda \quad \text{(2)}
\end{align*}
$$ 

***

::: {#big-text style="font-size: 125%"}
**🛠️ Step-by-step Formula Derivation** 
:::


From (1):  
$$
s = \mu r
$$

Substitute into (2):  
$$
\frac{\mu r - 1}{r} = \lambda
$$

$$
\mu - \frac{1}{r} = \lambda
$$

$$
\frac{1}{r} = \mu - \lambda
\Rightarrow r = \frac{1}{\mu - \lambda}
$$

Then from $s = \mu r$:  
$$
s = \mu \cdot \frac{1}{\mu - \lambda} = \frac{\mu}{\mu - \lambda}
$$

***

::: {#big-text style="font-size: 125%"}
**✅ Final Formula** 
:::

Given **mode** $\lambda$ and **mean** $\mu$, the Gamma parameters are:

$$
\boxed{
r = \frac{1}{\mu - \lambda}, \quad s = \frac{\mu}{\mu - \lambda}
}
$$ {#eq-005-tune-gamma-prior-mean-mode}


::: {#big-text style="font-size: 125%"}
**🛑 Conditions** 
:::

- $\mu > \lambda$ (mean > mode, which is true for Gamma distributions)
- $s > 1$ (required for mode formula to hold)

***

::: {#big-text style="font-size: 125%"}
**📝 Example** 
:::

If $\mu = 7$, $\lambda = 4$:

$$
r = \frac{1}{7 - 4} = \frac{1}{3}, \quad s = \frac{7}{3} \approx 2.333
$$

::: {#big-text style="font-size: 110%"}
<center>**Gamma(2.33, 0.33)**</center>
:::




::::
:::::

##### Mode & variance {#sec-005-tune-gamma-prior-mode-var}

:::::{.my-theorem}
:::{.my-theorem-header}
:::::: {#thm-005-tune-gamma-prior-mode-variance}
: How to tune a Gamma prior given mode and variance?
::::::
:::
::::{.my-theorem-container}

To find the appropriate **Gamma(s, r)** prior given a desired **mode** and **variance**, we use the known properties of the Gamma distribution in @eq-005-gamma-central-measures.

***

::: {#big-text style="font-size: 125%"}
**🔍 Gamma Distribution Properties** 
:::

For a Gamma distribution with shape $s$ and rate $r$:

- **Mean**: $\mu = \frac{s}{r}$
- **Variance**: $\sigma^2 = \frac{s}{r^2}$
- **Mode**: $\lambda = \frac{s - 1}{r}$, for $s > 1$

We are given:

- $\lambda$: desired **mode**
- $\sigma^2$: desired **variance**

We want to solve for $s$ and $r$.

***

::: {#big-text style="font-size: 125%"}
**🛠️ Step-by-Step Derivation** 
:::

We use the three equations:

1. $\mu = \frac{s}{r}$
2. $\sigma^2 = \frac{s}{r^2}$
3. $\lambda = \frac{s - 1}{r}$


::: {#big-text style="font-size: 110%"}
**Step 1: Use variance and mean** 
:::


From (1) and (2):

$$
\sigma^2 = \frac{s}{r^2} = \frac{1}{r} \cdot \frac{s}{r} = \frac{\mu}{r}
\Rightarrow r = \frac{\mu}{\sigma^2}
$$

But we don't have $\mu$ directly — we have $\lambda$ and $\sigma^2$.

So instead, use **mode** and **variance**.


::: {#big-text style="font-size: 110%"}
**Step 2: Express $s$ and $r$ in terms of $\lambda$ and $\sigma^2$** 
:::


From the mode:
$$
\lambda = \frac{s - 1}{r} \Rightarrow s = \lambda r + 1 \quad \text{(1)}
$$

From the variance:
$$
\sigma^2 = \frac{s}{r^2} \quad \text{(2)}
$$

Substitute (1) into (2):

$$
\sigma^2 = \frac{\lambda r + 1}{r^2} = \frac{\lambda}{r} + \frac{1}{r^2}
$$

Multiply both sides by $ r^2 $:

$$
\sigma^2 r^2 = \lambda r + 1
\Rightarrow \sigma^2 r^2 - \lambda r - 1 = 0
$$

This is a **quadratic in $r$**:

$$
\sigma^2 r^2 - \lambda r - 1 = 0
$$

Solve for $r$:

$$
r = \frac{ \lambda \pm \sqrt{ \lambda^2 + 4 \sigma^2 } }{2 \sigma^2}
$$

Since $r > 0$, we take the **positive root**:

$$
r = \frac{ \lambda + \sqrt{ \lambda^2 + 4 \sigma^2 } }{2 \sigma^2}
$$

Then from (1):

$$
s = \lambda r + 1
$$

***

::: {#big-text style="font-size: 125%"}
**✅ Final Formula** 
:::


Given **mode** $\lambda$ and **variance** $\sigma^2$, compute:

$$
\boxed{
r = \frac{ \lambda + \sqrt{ \lambda^2 + 4 \sigma^2 } }{2 \sigma^2}, \quad
s = \lambda r + 1
}
$$ {#eq-005-tune-gamma-prior-mode-var}



***

::: {#big-text style="font-size: 110%"}
**🛑 Edge Cases** 
:::


| Condition | Interpretation | Action |
|--------|----------------|--------|
| $\lambda \leq 0$ | Mode not positive | Invalid for Gamma with $s > 1$ |
| $\sigma^2 = 0$ | Zero variance | Degenerate — only possible if $s \to \infty$, $r \to \infty$ |
| $\lambda > 0$, $\sigma^2 > 0$ | Valid | Use formula |

***

::: {#big-text style="font-size: 110%"}
**📝 Example** 
:::


Let $\lambda = 4$, $\sigma^2 = 3$

Then:

Auxiliary calculation: Square root expression

$\sqrt{ \lambda^2 + 4 \sigma^2 } = \sqrt{16 + 12} = \sqrt{28} \approx 5.2915$

$$
\begin{align*}
r &= \frac{4 + 5.2915}{2 \cdot 3} = \frac{9.2915}{6} &\approx 1.5486 \\
s &= 4 \cdot 1.5486 + 1 &\approx 7.1944
\end{align*}
$$ 


::: {#big-text style="font-size: 110%"}
<center>**Gamma(7.19, 1.55)**</center>
:::


Check:

$$
\begin{align*}
\text{E}(\lambda) &= \frac{s}{r} &\approx \frac{7.1944}{1.5486} &\approx 4.64 \\
\text{Mode}(\lambda) &= \frac{s-1}{r} &\approx \frac{6.1944}{1.5486} &\approx 4.00 \quad ✅ \\
\text{Var}(\lambda) &= \frac{s}{r^2} &\approx \frac{7.1944}{1.5486^2} &\approx 3.00 \quad ✅
\end{align*}
$$ 

::::
:::::

##### Mean & variance {#sec-005-tune-gamma-prior-mean-var}

:::::{.my-theorem}
:::{.my-theorem-header}
:::::: {#thm-005-tune-gamma-prior-mean-variance}
: How to tune a Gamma prior given mean and variance?
::::::
:::
::::{.my-theorem-container}

To find the appropriate **Gamma(s, r)** prior given a desired **mean** $\mu$ and **variance** $\sigma^2$, we use the known properties of the Gamma distribution in @eq-005-gamma-central-measures.

***

::: {#big-text style="font-size: 125%"}
**🔍 Gamma Distribution Properties** 
:::


For a Gamma distribution with shape $s$ and rate $r$:

- **Mean**: $\mu = \frac{s}{r}$
- **Variance**: $\sigma^2 = \frac{s}{r^2}$

We solve for $s$ and $r$ using these two equations.

***

::: {#big-text style="font-size: 125%"}
**🛠️ Step-by-Step Derivation** 
:::


From the mean:
$$
s = \mu r \quad \text{(1)}
$$

Substitute into the variance:
$$
\sigma^2 = \frac{s}{r^2} = \frac{\mu r}{r^2} = \frac{\mu}{r}
\Rightarrow r = \frac{\mu}{\sigma^2}
$$

Now plug back into (1):
$$
s = \mu r = \mu \cdot \frac{\mu}{\sigma^2} = \frac{\mu^2}{\sigma^2}
$$

***

::: {#big-text style="font-size: 125%"}
**✅ Final Formula** 
:::


Given **mean** $\mu$ and **variance** $\sigma^2$, the Gamma parameters are:

$$
\boxed{
r = \frac{\mu}{\sigma^2}, \quad
s = \frac{\mu^2}{\sigma^2}
}
$$ {#eq-005-tune-gamma-prior-mean-var}

***

::: {#big-text style="font-size: 110%"}
**🛑 Edge Cases** 
:::


| Condition | Interpretation | Action |
|--------|----------------|--------|
| $\mu \leq 0$ | Mean not positive | Invalid (Gamma requires $\mu > 0$) |
| $\sigma^2 = 0$ | Zero variance | Degenerate — only possible if $s \to \infty$, $r \to \infty$ |
| $\mu > 0$, $\sigma^2 > 0$ | Valid | Use formula |

***

::: {#big-text style="font-size: 110%"}
**📝 Example** 
:::


Let $\mu = 7$, $\sigma^2 = 3$

Then:

$$
r = \frac{7}{3} \approx 2.333, \quad s = \frac{49}{3} \approx 16.333
$$

::: {#big-text style="font-size: 110%"}
<center>**Gamma(16.33, 2.33)**</center>
:::


Check:

$$
\begin{align*}
\text{E}(\lambda) &= \frac{s}{r} &\approx  \frac{16.333}{2.333} &\approx 7.00 \quad  ✅ \\ \text{Var}(\lambda) &= \frac{s}{r^2} &\approx  \frac{16.333}{2.333^2} &\approx 3.00 \quad  ✅ 
\end{align*}
$$ 


::::
:::::



:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-get-gamma-hyperparameter}
: Given mode and mean return Gamma hyperparameter
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: gamma-tuning
#| lst-label: lst-005-gamma-tuning
#| lst-cap: Gamma tuning 3 variants with given mode & mean; given mode & variance; given mean and variance
#| code-fold: show

my_gamma_prior_tuning <-  function(lambda = NULL, mu = NULL, var = NULL) {
  
  if (is.null(var)) {
    rate = round(1 / (mu - lambda), 2)
    shape =  round(mu / (mu - lambda), 2)
  }
  
  if (is.null(mu)) {
    rate = round((lambda + sqrt(lambda^2 + (4 * var))) / (2 * var), 2)
    shape = round(((lambda * rate ) + 1), 2)
  }
  
  if (is.null(lambda)) {
    rate = round((mu / var), 2)
    shape = round((mu^2 / var), 2)
  }
  
  # glue::glue("Gamma({shape}, {rate})")
  gamma <- tibble::tribble(
    ~shape, ~rate,
    shape, rate
  )
  gamma
}



```

::::
:::::



:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-tune-gamma-prior}
: Tuning a Gamma prior
::::::
:::
::::{.my-exercise-container}

For each situation below, tune an appropriate $\text{Gamma}(s, r)$ prior model for $\lambda$.

1. At first a tried to find a solution with trial & error using the `bayesrules::plot_gamma()` function. But I couldn't find a system how to change the parameters.
2. Therefore I changed the approach and solved the question mathematically (with the help of `r glossary("Qwen-14b")` (see: @sec-005-tune-gamma-prior-mean-mode, @sec-005-tune-gamma-prior-mode-var, @sec-005-tune-gamma-prior-mean-var).
3. I got as result different formulas (see: @eq-005-tune-gamma-prior-mean-mode, @eq-005-tune-gamma-prior-mode-var, @eq-005-tune-gamma-prior-mean-var).
4. Finally I integrated the formulas into the `my_gamma_prior_tuning()` (see: @lst-005-gamma-tuning).

My solutions for this exercise appear as 

- text $\text{Gamma}(s, r)$
- table with `bayesrules::summarize_gamma(shape, rate)` and
- figure with `bayesrules::plot_gamma(shape, rate, mean = TRUE, mode = TRUE)`


a) The most common value of λ is 4, and the mean is 7. Solution: **Gamma(`r my_gamma_prior_tuning(lambda = 4, mu = 7)[[1]]`,`r my_gamma_prior_tuning(lambda = 4, mu = 7)[[2]]`)**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-005-tune-gamma-prior-a}
: Plot the Gamma distribution for $\lambda = 4$, and $\mu = 7$
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-005-plot-tuning-gamma-a
#| fig-cap: The Gamma distribution for $\lambda = 4$, and $\mu = 7$.

shape <- my_gamma_prior_tuning(lambda = 4, mu = 7)[[1]]
rate  <- my_gamma_prior_tuning(lambda = 4, mu = 7)[[2]] 

my_plot_gamma(
  shape = shape,
  rate = rate,
  show_mean = TRUE,
  show_mode = TRUE
  )

```
:::

::: {.column width="50%"}
```{r}
#| label: tbl-005-summarize-tuning-gamma-a
#| tbl-cap: Summarize the Gamma model for $\lambda = 4$, and $\mu = 7$.

bayesrules::summarize_gamma(
  shape = my_gamma_prior_tuning(lambda = 4, mu = 7)[[1]],
  rate = my_gamma_prior_tuning(lambda = 4, mu = 7)[[2]]
) |>
  knitr::kable(digits = 4)
```
:::
:::::

::::
:::::


b) The most common value of λ is 10 and the mean is 12. Solution: **Gamma(`r my_gamma_prior_tuning(lambda = 10, mu = 12)[[1]]`,`r my_gamma_prior_tuning(lambda = 10, mu = 12)[[2]]`)**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-005-tune-gamma-prior-b}
: Plot the Gamma distribution for $\lambda = 10$, and $\mu = 12$
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-005-plot-tuning-gamma-b
#| fig-cap: The Gamma distribution for $\lambda = 10$, and $\mu = 12$.

shape <- my_gamma_prior_tuning(lambda = 10, mu = 12)[[1]]
rate  <- my_gamma_prior_tuning(lambda = 10, mu = 12)[[2]] 

my_plot_gamma(
  shape = shape,
  rate = rate,
  show_mean = TRUE,
  show_mode = TRUE
  )

```
:::

::: {.column width="50%"}
```{r}
#| label: tbl-005-summarize-tuning-gamma-b
#| tbl-cap: Summarize the Gamma model for $\lambda = 10$, and $\mu = 12$.

bayesrules::summarize_gamma(
  shape = my_gamma_prior_tuning(lambda = 10, mu = 12)[[1]],
  rate = my_gamma_prior_tuning(lambda = 10, mu = 12)[[2]]
) |>
  knitr::kable(digits = 4)
```
:::
:::::

::::
:::::


c) The most common value of λ is 5, and the variance is 3. Solution: **Gamma(`r my_gamma_prior_tuning(lambda = 5, var = 3)[[1]]`,`r my_gamma_prior_tuning(lambda = 5, var = 3)[[2]]`)**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-005-tune-gamma-prior-c}
: Plot the Gamma distribution for $\lambda = 5$, and $\text{var} = 3$
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-005-plot-tuning-gamma-c
#| fig-cap: The Gamma distribution for $\lambda = 5$, and $\text{var} = 3$.

shape <- my_gamma_prior_tuning(lambda = 5, var = 3)[[1]]
rate  <- my_gamma_prior_tuning(lambda = 5, var = 3)[[2]] 

my_plot_gamma(
  shape = shape,
  rate = rate,
  show_mean = TRUE,
  show_mode = TRUE
  )

```
:::

::: {.column width="50%"}
```{r}
#| label: tbl-005-summarize-tuning-gamma-c
#| tbl-cap: Summarize the Gamma model for $\lambda = 5$, and $\text{var} = 3$.

bayesrules::summarize_gamma(
  shape = my_gamma_prior_tuning(lambda = 5, var = 3)[[1]],
  rate = my_gamma_prior_tuning(lambda = 5, var = 3)[[2]]
) |>
  knitr::kable(digits = 4)
```
:::
:::::

::::
:::::


d) The most common value of λ is 14, and the variance is 6. **Gamma(`r my_gamma_prior_tuning(lambda = 14, var = 6)[[1]]`,`r my_gamma_prior_tuning(lambda = 14, var = 6)[[2]]`)**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-005-tune-gamma-prior-d}
: Plot the Gamma distribution for $\lambda = 14$, and $\text{var} = 6$
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-005-plot-tuning-gamma-d
#| fig-cap: The Gamma distribution for $\lambda = 14$, and $\text{var} = 6$.

shape <- my_gamma_prior_tuning(lambda = 14, var = 6)[[1]]
rate  <- my_gamma_prior_tuning(lambda = 14, var = 6)[[2]] 

my_plot_gamma(
  shape = shape,
  rate = rate,
  show_mean = TRUE,
  show_mode = TRUE
  )

```
:::

::: {.column width="50%"}
```{r}
#| label: tbl-005-summarize-tuning-gamma-d
#| tbl-cap: Summarize the Gamma model for $\lambda = 14$, and $\text{var} = 6$.

bayesrules::summarize_gamma(
  shape = my_gamma_prior_tuning(lambda = 14, var = 6)[[1]],
  rate = my_gamma_prior_tuning(lambda = 14, var = 6)[[2]]
) |>
  knitr::kable(digits = 4)
```
:::
:::::

::::
:::::



e) The mean of λ is 4 and the variance is 12. **Gamma(`r my_gamma_prior_tuning(mu = 4, var = 12)[[1]]`,`r my_gamma_prior_tuning(mu = 4, var = 12)[[2]]`)**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-005-tune-gamma-prior-e}
: Plot the Gamma distribution for $\mu = 4$, and $\text{var} = 12$
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-005-plot-tuning-gamma-e
#| fig-cap: The Gamma distribution for $\mu = 4$, and $\text{var} = 12$.

shape <- my_gamma_prior_tuning(mu = 4, var = 12)[[1]]
rate  <- my_gamma_prior_tuning(mu = 4, var = 12)[[2]] 

my_plot_gamma(
  shape = shape,
  rate = rate,
  show_mean = TRUE,
  show_mode = TRUE
  )

```
:::

::: {.column width="50%"}
```{r}
#| label: tbl-005-summarize-tuning-gamma-e
#| tbl-cap: Summarize the Gamma model for $\mu = 4$, and $\text{var} = 12$.

bayesrules::summarize_gamma(
  shape = my_gamma_prior_tuning(mu = 4, var = 12)[[1]],
  rate = my_gamma_prior_tuning(mu = 4, var = 12)[[2]]
) |>
  knitr::kable(digits = 4)
```
:::
:::::

::::
:::::


f) The mean of λ is 22 and the variance is 3. **Gamma(`r my_gamma_prior_tuning(mu = 22, var = 3)[[1]]`,`r my_gamma_prior_tuning(mu = 22, var = 3)[[2]]`)**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-005-tune-gamma-prior-f}
: Plot the Gamma distribution for $\mu = 22$, and $\text{var} = 3$
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}
```{r}
#| label: fig-005-plot-tuning-gamma-f
#| fig-cap: The Gamma distribution for $\mu = 22$, and $\text{var} = 3$.

shape <- my_gamma_prior_tuning(mu = 22, var = 3)[[1]]
rate  <- my_gamma_prior_tuning(mu = 22, var = 3)[[2]] 

my_plot_gamma(
  shape = shape,
  rate = rate,
  show_mean = TRUE,
  show_mode = TRUE
  )

```
:::

::: {.column width="50%"}
```{r}
#| label: tbl-005-summarize-tuning-gamma-f
#| tbl-cap: Summarize the Gamma model for $\mu = 22$, and $\text{var} = 3$.

bayesrules::summarize_gamma(
  shape = my_gamma_prior_tuning(mu = 22, var = 3)[[1]],
  rate = my_gamma_prior_tuning(mu = 22, var = 3)[[2]]
) |>
  knitr::kable(digits = 4)
```
:::
:::::

::::
:::::


::::
:::::

#### Poisson likelihood functions

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-poisson-likelihood-functions}
: Poisson likelihood functions
::::::
:::
::::{.my-exercise-container}

For each situation below, we observe the  outcomes for a random sample of Poisson variables, $Y_i \mid λ \overset{\text{ind}}{\sim} \text{Pois}(\lambda)$. Specify and plot the likelihood function of $\lambda$.

::: {#big-text style="font-size: 125%"}
**(a) $(y_1, y_2, y_3) = (3, 7, 19)$** 
:::

::: {#big-text style="font-size: 110%"}
**Poisson Likelihood Function** 
:::

The **likelihood function** for a Poisson distribution with parameter $\lambda$, given independent observations $y_1, y_2, y_3$, is the product of the individual Poisson `r glossary("Probability-Mass-Function", "probability mass functions")` (PMFs):

$$
L(\lambda \mid y_1, y_2, y_3) = \prod_{i=1}^3 \frac{e^{-\lambda} \lambda^{y_i}}{y_i!}
$$

For the specific data $(y_1, y_2, y_3) = (3, 7, 19)$, substitute the values:

$$
L(\lambda \mid 3, 7, 19) = \frac{e^{-\lambda} \lambda^3}{3!} \cdot \frac{e^{-\lambda} \lambda^7}{7!} \cdot \frac{e^{-\lambda} \lambda^{19}}{19!}
$$

Simplify the expression:

$$
L(\lambda \mid 3, 7, 19) = e^{-3\lambda} \cdot \frac{\lambda^{3+7+19}}{3! \cdot 7! \cdot 19!} = e^{-3\lambda} \cdot \frac{\lambda^{29}}{3! \cdot 7! \cdot 19!}
$$

This is the **Poisson likelihood function** for the given data.

::: {#big-text style="font-size: 110%"}
**Log-Likelihood Function** 
:::


To facilitate maximization, take the natural logarithm:

$$
\ln L(\lambda \mid 3, 7, 19) = -3\lambda + 29 \ln \lambda - \ln(3! \cdot 7! \cdot 19!)
$$



This log-likelihood is maximized at $\hat{\lambda} = \frac{29}{3} \approx 9.67$, the sample mean. In detail:

The `r glossary("MLE", "maximum likelihood estimate")` (MLE) of $\lambda$ for a Poisson distribution is the **sample mean**:

$$
\hat{\lambda} = \frac{1}{n} \sum_{i=1}^n y_i
$$

For the data $(y_1, y_2, y_3) = (3, 7, 19)$, compute:

$$
\hat{\lambda} = \frac{3 + 7 + 19}{3} = \frac{29}{3} \approx 9.67
$$

Thus, the MLE of $\lambda$ is **9.67**.



:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-plot-poisson-likelihood-a}
: Plot Poisson likelihood function for $(y_1, y_2, y_3) = (3, 7, 19)$
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-plot-poisson-likelihood-a
#| fig-cap: Poisson likelihood function for $(y_1, y_2, y_3) = (3, 7, 19)$
#| fig-height: 3
#| fig-width: 4

bayesrules::plot_poisson_likelihood(y = c(3, 7, 19), lambda_upper_bound = 20)
```

::::
:::::

***

::: {#big-text style="font-size: 125%"}
**b) $(y_1, y_2, y_3, y_4) = (12, 12, 12, 0)$** 
:::


$$
\hat{\lambda} = \frac{12, 12, 12, 0}{4} = \frac{36}{4} = 9
$$

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-plot-poisson-likelihood-b}
: Plot Poisson likelihood function for $(y_1, y_2, y_3, y_4) = (12, 12, 12, 0)$
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-plot-poisson-likelihood-b
#| fig-cap: Poisson likelihood function for $(y_1, y_2, y_3, y_4) = (12, 12, 12, 0)$
#| fig-height: 3
#| fig-width: 4

bayesrules::plot_poisson_likelihood(y = c(12, 12, 12, 0), lambda_upper_bound = 25)
```

::::
:::::

***


::: {#big-text style="font-size: 125%"}
**c) $y_1 = 12$** 
:::


$$
\hat{\lambda} = \frac{12}{1} = 12
$$

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-plot-poisson-likelihood-c}
: Plot Poisson likelihood function for $y = 12$
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-plot-poisson-likelihood-c
#| fig-cap: Poisson likelihood function for $y = 12$
#| fig-height: 3
#| fig-width: 4

bayesrules::plot_poisson_likelihood(y = 12, lambda_upper_bound = 25)
```

::::
:::::


::: {#big-text style="font-size: 125%"}
**d) $(y_1, y_2, y_3, y_4, y_5) = (16, 10, 17, 11, 11)$** 
:::


$$
\hat{\lambda} = \frac{16, 10, 17, 11, 11}{5} = \frac{65}{5} = 13
$$

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-plot-poisson-likelihood-d}
: Plot Poisson likelihood function for $(y_1, y_2, y_3, y_4, y_5) = (16, 10, 17, 11, 11)$
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-plot-poisson-likelihood-d
#| fig-cap: Poisson likelihood function for $(y_1, y_2, y_3, y_4, y_5) = (16, 10, 17, 11, 11)$
#| fig-height: 3
#| fig-width: 4

bayesrules::plot_poisson_likelihood(y = c(16, 10, 17, 11, 11), lambda_upper_bound = 25)
```

::::
:::::



::::
:::::

#### Finding the Gamma-Poisson posterior: Take I

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-find-gamma-poisson-posterior-I}
: Finding the Gamma-Poisson posterior (Take I)
::::::
:::
::::{.my-exercise-container}

Assume a prior model of 

$$\lambda \sim \text{Gamma}(24, 2)$$. 

Specify the posterior models of λ corresponding to the data in each scenario of @exr-005-poisson-likelihood-functions.

::: {#big-text style="font-size: 150%"}
**General solution** 
:::

Given a Poisson likelihood with observed data and a conjugate Gamma prior $\lambda \sim \text{Gamma}(\alpha = 24, \beta = 2)$, the posterior distribution for $\lambda$ is also Gamma.

The posterior parameters are calculated with @eq-005-gamma-poisson-model in @thm-005-gamma-poisson-model.

$$
\begin{align*}
\lambda \mid \overrightarrow{\rm y} &\sim \text{Gamma}(s + \sum y_i, r + n) \\
\text{Shape:} \quad \alpha_{\text{post}} &= \alpha + \sum y_i  \\
\text{Rate:} \quad \beta_{\text{post}} &= \beta + n
\end{align*}
$$

***

::: {#big-text style="font-size: 110%"}
**(a) $(y_1, y_2, y_3) = (3, 7, 19)$** 
:::


$$
\begin{align*}
\text{Shape:} \quad \alpha_{\text{post}} &= \alpha + \sum y_i = 24 + (3 + 7 + 19) = 53 \\
\text{Rate:} \quad \beta_{\text{post}} &= \beta + n = 2 + 3 = 5
\end{align*}
$$ 

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-find-gamma-poisson-posterior-a}
: Plot the posterior models of $\lambda \sim \text{Gamma}(24, 2)$ and $(y_1, y_2, y_3) = (3, 7, 19)$ 
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-find-gamma-poisson-posterior-a
#| fig-cap: Specify the posterior models of Gamma(24, 2) and samples of 3, 7, 19 
#| results: hold

bayesrules::summarize_gamma_poisson(24,2,29,3) |> 
  knitr::kable(digits = 2)

bayesrules::plot_gamma_poisson(
  shape = 24,
  rate = 2,
  sum_y = 29,
  n = 3,
  prior = TRUE,
  likelihood = TRUE,
  posterior = TRUE
)

```

::::
:::::

***


::: {#big-text style="font-size: 110%"}
**b) $(y_1, y_2, y_3, y_4) = (12, 12, 12, 0)$** 
:::


$$
\begin{align*}
\bf{\text{Shape:}} \quad \alpha_{\text{post}} &= \alpha + \sum y_i = 24 + (12 + 12 + 12 + 0) = 60 \\
\bf{Rate:} \quad \beta_{\text{post}} &= \beta + n = 2 + 4 = 6
\end{align*}
$$ 

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-find-gamma-poisson-posterior-b}
: Plot the posterior models of $\lambda \sim \text{Gamma}(24, 2)$ and $(y_1, y_2, y_3, y_4) = (12, 12, 12, 0)$ 
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-find-gamma-poisson-posterior-b
#| fig-cap: Specify the posterior models of Gamma(24, 2) and samples of 12, 12, 12, 0 
#| results: hold

bayesrules::summarize_gamma_poisson(24,2,36,4) |> 
  knitr::kable(digits = 2)

bayesrules::plot_gamma_poisson(
  shape = 24,
  rate = 2,
  sum_y = 36,
  n = 4,
  prior = TRUE,
  likelihood = TRUE,
  posterior = TRUE
)

```

::::
:::::

***


::: {#big-text style="font-size: 110%"}
**c) $y_1 = 12$** 
:::


$$
\begin{align*}
\text{Shape:} \quad \alpha_{\text{post}} &= \alpha + \sum y_i = 24 + 12 = 36 \\
\text{Rate:} \quad \beta_{\text{post}} &= \beta + n = 2 + 1 = 3
\end{align*}
$$ 

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-find-gamma-poisson-posterior-c}
: Plot the posterior models of $\lambda \sim \text{Gamma}(24, 2)$ and $(y) = 12$ 
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-find-gamma-poisson-posterior-c
#| fig-cap: Specify the posterior models of Gamma(24, 2) and samples of y = 12 
#| results: hold

bayesrules::summarize_gamma_poisson(24,2,12,1) |> 
  knitr::kable(digits = 2)

bayesrules::plot_gamma_poisson(
  shape = 24,
  rate = 2,
  sum_y = 12,
  n = 1,
  prior = TRUE,
  likelihood = TRUE,
  posterior = TRUE
)

```

::::
:::::

***

::: {#big-text style="font-size: 110%"}
**d) $(y_1, y_2, y_3, y_4, y_5) = (16, 10, 17, 11, 11)$** 
:::


$$
\begin{align*}
\bf{\text{Shape:}} \quad \alpha_{\text{post}} &= \alpha + \sum y_i = 24 + (16 + 10 + 17 + 11 + 11) = 89 \\
\bf{Rate:} \quad \beta_{\text{post}} &= \beta + n = 2 + 5 = 7
\end{align*}
$$ 

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-find-gamma-poisson-posterior-a}
: Plot the posterior models of $\lambda \sim \text{Gamma}(24, 2)$ and $(y_1, y_2, y_3, y_4, y_5) = (16 + 10 + 17 + 11 + 11)$ 
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-find-gamma-poisson-posterior-d
#| fig-cap: Specify the posterior models of Gamma(24, 2) and samples of 16, 10, 17, 11, 11 
#| results: hold

bayesrules::summarize_gamma_poisson(24,2,65,5) |> 
  knitr::kable(digits = 2)

bayesrules::plot_gamma_poisson(
  shape = 24,
  rate = 2,
  sum_y = 65,
  n = 5,
  prior = TRUE,
  likelihood = TRUE,
  posterior = TRUE
)

```

::::
:::::


::::
:::::



#### Finding the Gamma-Poisson posterior: Take II

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-find-gamma-poisson-posterior-b}
: 
::::::
:::
::::{.my-exercise-container}
Nothing new: Exercise skipped!
::::
:::::

#### Text messages

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-text-messages}
: Text messages
::::::
:::
::::{.my-exercise-container}

Let random variable $\lambda$ represent the rate of text messages people receive in an hour. At first, you believe that the typical number of messages per hour is $5$ with a standard deviation of $0.25$ messages.  

a) Tune and plot an appropriate $\text{Gamma}(s, r)$ prior model for $\lambda$. 

Using @eq-005-tune-gamma-prior-mean-var:

$$
\begin{align*}
&r = \frac{\mu}{\sigma^2}, \quad &s = \frac{\mu^2}{\sigma^2} \\
&r = \frac{5}{0.5^2}, \quad &s = \frac{5^2}{0.5^2} \\
&r = 80, \quad &s = 400
\end{align*}
$$ 

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-text-messages}
: Text messages
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}

```{r}
#| label: tbl-005-text-messages
#| tbl-cap: Summarize Gamma prior for 5 events per hour and 0.25 sd

bayesrules::summarize_gamma(400, 80) |> 
  knitr::kable()
```
:::

::: {.column width="50%"}

```{r}
#| label: fig-005-text-messages
#| fig-cap: Plot Gamma prior for 5 events per hour and 0.25 sd

bayesrules::plot_gamma(400, 80)
```
:::
:::::

::::
:::::





b) What is the prior probability that the rate of text messages per hour is larger than 10? Hint: learn about `stats::pgamma()`.

```{r}
#| label: prior-prob-bigger-than-10
#| code-fold: show

stats::pgamma(10, 400, 80, lower.tail = FALSE)
```


The result is approximately $0$ (essentially zero), which makes sense given that our prior belief centers strongly around $5$ with very little variability.

::: {.callout-note #nte-005-prior-prob-bigger-than-10}
###### Alternate calcuation: Substract from 1

Another possible calculation is:

`1 - stats::pgamma(10, 400, 80)` = `r 1 - stats::pgamma(10, 400, 80)`

We subtract `pgamma()` from 1 because **`pgamma(x, ...)` gives the cumulative probability that the random variable is *less than or equal to* `x`** — that is, $P(\lambda \leq x)$.

Since we want the probability that $\lambda > 10$, we need:

$$
P(\lambda > 10) = 1 - P(\lambda \leq 10) = 1 - \text{pgamma}(10, \text{shape}, \text{rate})
$$

This is standard for any continuous distribution when we want the *upper tail* probability.

**In short**: `pgamma()` calculates the left-tail area — subtracting it from 1 gives us the right-tail area, which is what we were asking for.


:::






::::
:::::

#### Women’s World Cup {#sec-005-wwc}

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-world-cup}
: Women’s World Cup (WWC)
::::::
:::
::::{.my-exercise-container}

Let $\lambda$ be the average number of goals scored in a Women’s World Cup game. We’ll analyze $\lambda$ by the following Gamma-Poisson model where data $Y_i$ is the observed number of goals scored in a sample of World Cup games:

$$
\begin{align*}
Y_i \mid \lambda &\overset{\text{ind}}{\sim} \text{Pois}(\lambda) \\
\lambda &\sim \text{Gamma}(1, 0.25)
\end{align*}
$$ 

***

**a) Plot and summarize our prior understanding of $\lambda$.**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-world-cup}
: WWC: Prior understanding
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}

```{r}
#| label: fig-005-world-cup-prior
#| fig-cap: Women’s World Cup prior understanding of $\lambda$.

bayesrules::plot_gamma(1, 0.25)
```
:::

::: {.column width="50%"}

```{r}
#| label: tbl-005-world-cup-prior
#| tbl-cap: Women’s World Cup prior understanding of $\lambda$.

bayesrules::summarize_gamma(1, 0.25) |> 
  knitr::kable()
```
:::
:::::

::::
:::::

The prior understanding from the book is that 4 goals are the average number of goals ($\lambda$) scored in a Women’s World Cup game.  Additionally with this prior related is the believe that in most of the matches no goals were scored but in some the number of goals can even surpass 15 goals. 

::: {.callout-note #nte-005-exr-world-cup}
###### An alternative --- more realistic? --- prior

It seems to me that a mode with 0 goals is not a realistic assumption. I would argue that for scored goals a right-skewed prior distribution with a mode of perhaps 2 and mean with 4 would be more realistic.

Let's try out and find the appropriate Gamma for this prior taking @eq-005-tune-gamma-prior-mean-mode:

$$
\begin{align*}
r = \frac{1}{\mu - \lambda}, \quad &s = \frac{\mu}{\mu - \lambda} \\
r = \frac{1}{4-2} = 0.5, \quad &s = \frac{4}{4-2} = 2
\end{align*}
$$ 

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-alternative-world-cup}
: Alternative prior for the Women’s World Cup Gamma distribution
::::::
:::
::::{.my-r-code-container}

::::: columns
::: {.column width="50%"}

```{r}
#| label: fig-005-exr-alternative-world-cup
#| fig-cap: Alternative prior for number of scored goals per game (Gamma(2. 0.5))
#| fig-height: 3

bayesrules::plot_gamma(2, 0.5) +
  ggplot2::xlim(0, 20)
```
:::

::: {.column width="50%"}

```{r}
#| label: tbl-005-exr-alternative-world-cup
#| tbl-cap: Alternative prior for number of scored goals per game (Gamma(2. 0.5))

bayesrules::summarize_gamma(2, 0.5) |> 
  knitr::kable()
```
:::
:::::

::::
:::::

:::


**b) Why is the Poisson model a reasonable choice for our data $Y_i$?**

The Poisson model is a reasonable choice for scored goals in football games because goals are a number of independent events that occur. in a fixed amount of time (2 $\times$ 45 min.). The number of goals is a random count that can go from 0 to a (theoretically) really big number.

***


The `wwc_2019_matches` data in the {**fivethirtyeight**} package includes the number of goals scored by the two teams in each 2019 Women’s World Cup match. 

**c) Define, plot, and discuss the total number of goals scored per game**


```{r}
#| label: fig-005-exr-world-cup
#| fig-cap: Goals scored per game
#| fig-height: 3
#| fig-width: 4

data("wwc_2019_matches", package = "fivethirtyeight")
wwc_2019_matches <- wwc_2019_matches |> 
  dplyr::mutate(total_goals = score1 + score2)

ggplot2::ggplot(wwc_2019_matches,
                ggplot2::aes(x = total_goals)) +
  ggplot2::geom_histogram(binwidth = 1, color = "white") +
  ggplot2::scale_x_continuous(breaks = 0:15, minor_breaks = 0) +
  ggplot2::scale_y_continuous(
    breaks = 0:20, 
    labels = ifelse(0:20 %% 2 == 0, as.character(0:20), ""),
    minor_breaks = 0)
```


By far the most common number of goals per match is 3. Normally the scored goals  range between 0 and 6 goals per match. But in some extraordinary matches, especially if the playing strength of teams is very different, there could be also a much higher number of goals per match. 

Additionally it seems to me that my own prior developed in @nte-005-exr-world-cup matches somewhat better the data.

***

**d) Plot the prior pdf, likelihood function, and posterior pdf of $lambda$. Describe the evolution in your understanding of $lambda$λ from the prior to the posterior.**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-5-7d-plot-posterior}
: Plot the prior pdf, likelihood function, and posterior pdf of Gamma(1, 0.25) with the data from `wwc_2019_matches`.
::::::
:::
::::{.my-r-code-container}

- Number of events (matches): `base::nrow(wwc_2019_matches)` = `r base::nrow(wwc_2019_matches)`
- Number of goals: `base::sum(wwc_2019_matches$total_goals)` = `r base::sum(wwc_2019_matches$total_goals)`

```{r}
#| label: fig-005-exr-5-7d-plot-posterior
#| fig-cap: Prior pdf, likelihood function, and posterior pdf of Gamma(1, 0.25) with `wwc_2019_matches` data.
#| fig-height: 3

bayesrules::plot_gamma_poisson(
  shape = 1,
  rate = 0.25,
  sum_y = base::sum(wwc_2019_matches$total_goals),
  n = base::nrow(wwc_2019_matches)
)
```


```{r}
#| label: tbl-005-exr-5-7d-plot-posterior
#| tbl-cap: Summarize Prior and posterior pdf of Gamma(1, 0.25) with `wwc_2019_matches` data 
#| fig-height: 3

bayesrules::summarize_gamma_poisson(
  shape = 1,
  rate = 0.25,
  sum_y = base::sum(wwc_2019_matches$total_goals),
  n = base::nrow(wwc_2019_matches)
) |> 
  knitr::kable(digits = 4)
```


::::
:::::

Although I knew that with many data the posterior is mostly influenced by the data, it still was a surprise to me how much the data dominated the outcome and how small the posterior variance is. When data are so overwhelming then the chosen prior is of minor importance (as long it is a vage prior). 


::: {.callout-note #nte-005-exr-5-7d-plot-posterior-alternative}
When data are so overwhelming then the chosen prior is of minor importance (as long it is a vage prior). 

In this case my different prior didn't have any effect on the posterior outcome.


:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-5-7d-plot-posterior}
: Plot prior pdf, likelihood function, and posterior pdf of Gamma(2, 0.5)
::::::
:::
::::{.my-r-code-container}


```{r}
#| label: fig-005-exr-5-7d-plot-posterior2
#| fig-cap: Prior pdf, likelihood function, and posterior pdf of Gamma(2, 0.5) with `wwc_2019_matches` data.
#| fig-height: 3

bayesrules::plot_gamma_poisson(
  shape = 2,
  rate = 0.5,
  sum_y = base::sum(wwc_2019_matches$total_goals),
  n = base::nrow(wwc_2019_matches)
)
```


```{r}
#| label: tbl-005-exr-5-7d-plot-posterior2
#| tbl-cap: Summarize Prior and posterior pdf of Gamma(1, 0.25) with `wwc_2019_matches` data 
#| fig-height: 3

bayesrules::summarize_gamma_poisson(
  shape = 2,
  rate = 0.5,
  sum_y = base::sum(wwc_2019_matches$total_goals),
  n = base::nrow(wwc_2019_matches)
) |> 
  knitr::kable(digits = 4)
```


::::
:::::

***

With 52 matches the posterior mean is nearly identical to the frequentist `r glossary("MLE")` (2.81), because the data overwhelms the prior.

::: {#big-text style="font-size: 125%"}
**Key differences** 
:::

| Category                | Frequentist                                                                                    |  Bayesian                                                                           |
|----------------------------|---------------------------------------------------------------------------------------------|---------------------------------------------------------------------------|
| λ is...                    | Fixed unknown constant                                                                      | Random variable with a distribution                                       |
| Estimate                   | MLE: sum_y / n = 2.81                                                                       | Posterior mean: (shape + sum_y) / (rate + n) ≈ 2.813                      |
| Uncertainty                | 95% CI via qgamma()                                                                         | 95% credible interval from posterior                                      |
| Prior                      | None                                                                                        | Gamma(1, 0.25) — your belief before seeing data                           |
| Interpretation of interval | "If we repeated this experiment many times, 95% of such intervals would contain the true λ" | "Given the data, there is a 95% probability that λ lies in this interval" |

::: {#big-text style="font-size: 125%"}
**Why are the results so similar here?** 
:::

With n = 52 and total_goals = 146, the data strongly dominate the prior. The prior only contributes shape = 1 and rate = 0.25 — equivalent to having observed roughly 1 goal in 0.25 matches — which is negligible compared to 146 goals in 52 matches. **This prior washing out by large samples is a general property: Bayesian and frequentist estimates converge as n grows.**

:::::{.my-resource}
:::{.my-resource-header}
:::::: {#lem-005-exr-diff-freq-bayes}
: Differences between Frequentist and Bayesian Approach
::::::
:::
::::{.my-resource-container}

With the result of @exr-005-world-cup I wondered in which situations there might be a decisive difference between a Frequentist and a Bayesian approach. I asked the Positron Assistant and recorded the answers under @sec-910-advantages-bayes.

::::
:::::

:::

::::
:::::

### Practice Normal-Normal

#### Normal Likelihood Functions

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-normal-likelihood-functions}
: Normal Likelihood Functions
::::::
:::
::::{.my-exercise-container}
In each situation below, we observe the outcomes for a Normal random sample, $Y_i \mid μ \overset{\text{ind}}{\sim} N (\mu, \sigma^2)$ with known $\sigma$. Specify and plot the corresponding likelihood function of $\mu$.

***

**a) $\bf{(y_1, y_2, y_3) = (−4.3, 0.7, −19.4)}$ and $\bf{\sigma = 10}$**

The sample mean is $\bar{y} = (-4.3 + 0.7 - 19.4) / 3 \approx -7.67$ and $n = 3$.
The likelihood function is:

$$
L(\mu \mid \vec{y}) \propto \exp\left(-\frac{3(\bar{y} - \mu)^2}{2 \cdot 10^2}\right) \text{ for } \mu \in (-\infty, \infty)
$$

This is maximized at $\hat{\mu} = \bar{y} \approx -7.67$.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-normal-likelihood-a}
: Plot Normal likelihood for $(y_1, y_2, y_3) = (-4.3, 0.7, -19.4)$, $\sigma = 10$
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-normal-likelihood-a
#| fig-cap: Normal likelihood function for $\mu$ given $(y_1, y_2, y_3) = (-4.3, 0.7, -19.4)$ and $\sigma = 10$.
#| fig-height: 3

bayesrules::plot_normal_likelihood(
  y     = c(-4.3, 0.7, -19.4),
  sigma = 10
)
```
::::
:::::


**b) $\bf{(y1, y2, y3, y4) = (−12, 1.2, −4.5, 0.6)}$ and $\bf{σ = 6}$**

The sample mean is $\bar{y} = (−12, 1.2, −4.5, 0.6) / 4 = -3.675$ and $n = 4$.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-normal-likelihood-b}
: Plot Normal likelihood for $(y_1, y_2, y_3, y_4) = (−12, 1.2, −4.5, 0.6)$ and $\sigma = 6$
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-normal-likelihood-b
#| fig-cap: Normal likelihood function for $\mu$ given $(y_1, y_2, y_3, y_4) = (−12, 1.2, −4.5, 0.6)$ and $\sigma = 6$.
#| fig-height: 3

bayesrules::plot_normal_likelihood(
  y     = c(-12, 1.2, -4.5, 0.6),
  sigma = 6
)
```
::::
:::::

c) $(y_1, y_2) = (12.4, 6.1)$ and $\sigma = 5$: skipped, nothing new.  
d) $(y_1, y_2, y_3, y_4, y_5) = (1.6, 0.09, 1.7, 1.1, 1.1)$ and $\sigma = 0.6$: skipped nothing new.

::::
:::::


#### Investing in stock

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-investing-in-stock}
: Investing in stock
::::::
:::
::::{.my-exercise-container}
You just bought stock in FancyTech. Let random variable $\mu$ be the average dollar amount that your FancyTech stock goes up or down in a one-day period. At first, you believe that $\mu$ is 7.2 dollars with a standard deviation of 2.6 dollars.

**a) Tune and plot an appropriate Normal prior model for $\mu$.**

Prior = $\mu \sim N(7.2, 2.6^2)$: mean belief is: +$7.2 per day, with sd of $2.6.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-investing-in-stock}
: Investing in stock
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-investing-in-stock
#| fig-cap: "Normal prior about stock developments with μ = 7.2 and sd = 2.6 dollars"
#| fig-height: 3

bayesrules::plot_normal(7.2, 2.6)
```



::::
:::::

**b) According to your plot, does it seem plausible that the *FancyTech* stock would increase by an average of 7.6 dollars in a day?**

Yes, 7.6 is very close to the prior mean of 7.2

**c) Does it seem plausible that the *FancyTech* stock would increase by an average of 4 dollars in a day?**

Again yes, because 4 is within 2 standard deviations $(7.2 \pm 2 \times (2.6) = [2.0, 12.4])$

**d) What is the prior probability that, on average, the stock price goes down? Hint: `pnorm()`.**

`stats::pnorm(q, mean, sd)` computes $P(\mu \leq q)$ --- the cumulative probability up to the value of `q`.

"Stock goes down" means $\mu < 0$, i.e., the average daily change is **negative** (= a loss). Zero is the boundary between going up and going down, so

$$
P(\text{stock goes down}) = P(\mu < 0) = P(\mu leq 0)
$$

Setting `q = 0` in `pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)` gives exactly the left-tail probability under the $N(7.2, 2.6^2)$ prior, because the default value is `lower.tail = TRUE` returns the value of the left side of the mean.

So we have: `stats::pnorm(0, mean = 7.2, sd = 2.6, lower.tail = TRUE)` = `r stats::pnorm(0, mean = 7.2, sd = 2.6)`.

The result will be approximately `r round(stats::pnorm(0, mean = 7.2, sd = 2.6), 4)` — there's only a ~`r round(stats::pnorm(0, mean = 7.2, sd = 2.6), 4) * 100`% prior probability that the stock goes down on average. This makes sense: 0 is almost 2.8 standard deviations below the prior mean $(\frac{0 - 7.2}{2.6} \approx -2.77)$, so it's very unlikely under our prior beliefs.

**e) What is the prior probability that, on average, your stock price goes up by more than 8 dollars per day?**

Again I will use `stats::pnorm(q, mean, sd)` but this time for computing $P(\mu \geq q)$, with `lower.tail = FALSE`:

`stats::pnorm(8, mean = 7.2, sd = 2.6, lower.tail = FALSE)` = `r stats::pnorm(8, mean = 7.2, sd = 2.6, lower.tail = FALSE)`.

The result will be approximately `r round(stats::pnorm(8, mean = 7.2, sd = 2.6, lower.tail = FALSE), 4)` — there's only a ~`r round(stats::pnorm(8, mean = 7.2, sd = 2.6, lower.tail = FALSE), 4) * 100`% prior probability that the stock goes higher on average. This makes sense: 8 is near the mean only $(\frac{8 - 7.2}{2.6} \approx 0.31)$ standard deviation above the prior mean, so it's very likely under our prior beliefs that on average the stock price goes up by more than 8 dollars per day.

::::
:::::

#### Investing in stock with data

Continuing with @exr-005-investing-in-stock, it’s reasonable to assume that the daily changes in *FancyTech* stock value are Normally distributed around an unknown mean of $\mu$ with a known standard deviation of $\sigma = 2$ dollars. On a random sample of $4$ days, you observe changes in stock value of $-0.7, 1.2, 4.5, \text{and} -4$ dollars.

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-stock-data}
: Investing in stock with data
::::::
:::
::::{.my-exercise-container}
**a) Plot the corresponding likelihood function of $\mu$.**


:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-stock-data-a}
: Investing in stock with data
::::::
:::
::::{.my-r-code-container}




```{r}
#| label: fig-005-stock-data-a
#| fig-cap: Normal likelihood function for $\mu$ given $(y_1, y_2, y_3, y_4) = -0.7, 1.2, 4.5, \text{and} -4$ dollars and $\sigma = 2$ US dollars.
#| fig-height: 3

bayesrules::plot_normal_likelihood(
  y     = c(-0.7, 1.2, 4.5, -4),
  sigma = 2
)
```

::::
:::::

**b) Plot the prior pdf, likelihood function, and the posterior pdf for $\mu$.**


:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-fig-005-stock-data-b}
: Plot the prior pdf, likelihood function, and the posterior pdf for $\mu$.
::::::
:::
::::{.my-r-code-container}


```{r}
#| label: fig-005-stock-data-b
#| fig-cap: Prior pdf, likelihood function, and the posterior pdf for $\bf{\mu}$ given $\bf{(y_1, y_2, y_3, y_4) = -0.7, 1.2, 4.5, -4}$ dollars and $\bf{\sigma = 2}$ US dollars.
#| fig-height: 3

bayesrules::plot_normal_normal(
  mean = 7.2,
  sd = 2.6,
  sigma = 2,
  y_bar = mean(c(-0.7, 1.2, 4.5, -4)),
  n = 4
)
```

**c) Use summarize_normal_normal() to calculate descriptive statistics for the prior and the posterior models.**

```{r}
#| label: tbl-005-stock-data-b
#| tbl-cap: Prior pdf, likelihood function, and the posterior pdf for $\bf{\mu}$ given $\bf{(y_1, y_2, y_3, y_4) = -0.7, 1.2, 4.5, -4}$ dollars and $\bf{\sigma = 2}$ US dollars.

bayesrules::summarize_normal_normal(
  mean = 7.2,
  sd = 2.6,
  sigma = 2,
  y_bar = mean(c(-0.7, 1.2, 4.5, -4)),
  n = 4
) |> 
  knitr::kable()
```


::::
:::::


**d) Comment on how your understanding about μ evolved from the prior (in the previous exercise) to the posterior based on the observed data.**

The prior was way too optimistic. The stock prices didn't center around \$7.2 but only around \$1.15 with the possibility that it could be even a loss (< \$0). The big uncertainty has vanished from $(7.2 \pm 2 \times (2.6) = [2.0, 12.4])$ to $(1.15 \pm 2 \times (0.93) = [-0.97, 3.01])$.

**e) What is the posterior probability that, on average, the stock price goes *down*? Hint: `pnorm()`.**

`stats::pnorm(0, mean = 0.25, sd = 2, lower.tail = TRUE)` = `r stats::pnorm(0, mean = 0.25, sd = 2, lower.tail = TRUE)`.

The result will be approximately `r round(stats::pnorm(0, mean = 0.25, sd = 2), 4)` — there's a relatively high probability of ~`r round(stats::pnorm(0, mean = 0.25, sd = 2), 4) * 100`% of posterior probability that the stock goes down on average. This makes sense if you look at @fig-005-stock-data-b where you can see that a considerable proportions is under 0.

**f) What is the posterior probability that, on average, your stock price goes *up* by more than 8 dollars per day?**

`stats::pnorm(8, mean = 0.25, sd = 2, lower.tail = FALSE)` = `r stats::pnorm(8, mean = 0.25, sd = 2, lower.tail = FALSE)`.

There's practically no chance (~`r round(stats::pnorm(8, mean = 0.25, sd = 2, lower.tail = FALSE), 4) * 100`%) of posterior probability that the stock goes above 8 US dollars on average. This makes sense if you look at @fig-005-stock-data-b where you can see that only a very tiny fraction of the right tail is over 8. Remember that the normal distribution has no discrete borders and goes from $-\infty$ to $+\infty$. There will *never* be a probability of exactly 0.

::::
:::::


#### Normal-Normal calculation

Prof. Abebe and Prof. Morales both recently finished their PhDs and are teaching their first statistics classes at Bayesian University. Their colleagues told them that the *average* final exam score across all students, $\mu$, varies Normally from year to year with a mean of 80 points and a standard deviation of 4. Further, individual students’ scores Y vary Normally around $\mu$ with a known standard deviation of 3 points.

:::::{.my-theorem}
:::{.my-theorem-header}
:::::: {#thm-005-derivation-posterior-variance-mean}
: Step-by-Step Derivation of Posterior Variance and Mean
::::::
:::
::::{.my-theorem-container}


::: {#big-text style="font-size: 125%"}
**1. Starting Point: Bayes' Theorem for Normal Data** 
:::

When we observe data $y_1, y_2, \ldots, y_n$ from a Normal distribution, Bayes' theorem gives us:

$$p(\mu \mid \text{data}) \propto p(\text{data} \mid \mu) \cdot p(\mu)$$

For the Normal-Normal model:

- **Prior:** $p(\mu) = \mathcal{N}(\theta, \tau^2)$
- **Likelihood:** $p(y_i \mid \mu) = \mathcal{N}(\mu, \sigma^2)$ for each observation

***

::: {#big-text style="font-size: 125%"}
**Step 1: Write Out the Proportionality** 
:::

The posterior is proportional to the product of the prior and likelihood:

$$p(\mu \mid \text{data}) \propto \exp\left(-\frac{(\mu - \theta)^2}{2\tau^2}\right) \cdot \prod_{i=1}^{n} \exp\left(-\frac{(y_i - \mu)^2}{2\sigma^2}\right)$$

Combining the exponentials:

$$p(\mu \mid \text{data}) \propto \exp\left(-\frac{(\mu - \theta)^2}{2\tau^2} - \sum_{i=1}^{n}\frac{(y_i - \mu)^2}{2\sigma^2}\right)$$

***

::: {#big-text style="font-size: 125%"}
**Step 2: Expand the Exponent** 
:::


Let's expand both squared terms in the exponent:

$$-\frac{(\mu - \theta)^2}{2\tau^2} - \sum_{i=1}^{n}\frac{(y_i - \mu)^2}{2\sigma^2}$$

$$= -\frac{\mu^2 - 2\mu\theta + \theta^2}{2\tau^2} - \sum_{i=1}^{n}\frac{y_i^2 - 2y_i\mu + \mu^2}{2\sigma^2}$$

$$= -\frac{\mu^2}{2\tau^2} + \frac{2\mu\theta}{2\tau^2} - \frac{\theta^2}{2\tau^2} - \frac{n\mu^2}{2\sigma^2} + \frac{2\mu\sum y_i}{2\sigma^2} - \frac{\sum y_i^2}{2\sigma^2}$$

***

::: {#big-text style="font-size: 125%"}
**Step 3: Collect Terms by Powers of $\mu$** 
:::


Group by $\mu^2$, $\mu^1$, and $\mu^0$ (constants):

$$= -\frac{\mu^2}{2}\left(\frac{1}{\tau^2} + \frac{n}{\sigma^2}\right) + \mu\left(\frac{\theta}{\tau^2} + \frac{\sum y_i}{\sigma^2}\right) - \left(\frac{\theta^2}{2\tau^2} + \frac{\sum y_i^2}{2\sigma^2}\right)$$

The last term (constants) doesn't affect the shape of the posterior, so we drop it:

$$\propto -\frac{\mu^2}{2}\left(\frac{1}{\tau^2} + \frac{n}{\sigma^2}\right) + \mu\left(\frac{\theta}{\tau^2} + \frac{\sum y_i}{\sigma^2}\right)$$

***

::: {#big-text style="font-size: 125%"}
**Step 4: Recognize the Normal Form** 
:::

This is the exponent of a Normal distribution. Recall that a Normal with mean $\mu_n$ and variance $\tau_n^2$ has the form:

$$\propto \exp\left(-\frac{(\mu - \mu_n)^2}{2\tau_n^2}\right) = \exp\left(-\frac{\mu^2 - 2\mu\mu_n + \mu_n^2}{2\tau_n^2}\right)$$

$$\propto \exp\left(-\frac{\mu^2}{2\tau_n^2} + \frac{\mu\mu_n}{\tau_n^2}\right)$$

***

::: {#big-text style="font-size: 125%"}
**Step 5: Match Coefficients** 
:::


We have from Step 3:

$$\propto -\frac{\mu^2}{2}\left(\frac{1}{\tau^2} + \frac{n}{\sigma^2}\right) + \mu\left(\frac{\theta}{\tau^2} + \frac{\sum y_i}{\sigma^2}\right)$$

And from the standard Normal form:

$$\propto -\frac{\mu^2}{2\tau_n^2} + \frac{\mu\mu_n}{\tau_n^2}$$

***

::: {#big-text style="font-size: 110%"}
**Match Coefficient of $\mu^2$:** 
:::


$$\frac{1}{\tau_n^2} = \frac{1}{\tau^2} + \frac{n}{\sigma^2}$$

This gives us the **posterior precision** (inverse variance). Solving for posterior variance:

$$\boxed{\tau_n^2 = \frac{1}{\dfrac{1}{\tau^2} + \dfrac{n}{\sigma^2}}}$$

::: {#big-text style="font-size: 110%"}
**Match Coefficient of $\mu$:** 
:::


$$\frac{\mu_n}{\tau_n^2} = \frac{\theta}{\tau^2} + \frac{\sum y_i}{\sigma^2}$$

Now solve for $\mu_n$:

$$\mu_n = \tau_n^2 \left( \frac{\theta}{\tau^2} + \frac{\sum y_i}{\sigma^2} \right)$$

But since $\sum y_i = n \bar{y}$, we can write:

$$\boxed{\mu_n = \tau_n^2 \left( \frac{\theta}{\tau^2} + \frac{n \bar{y}}{\sigma^2} \right)}$$

***

::: {#big-text style="font-size: 125%"}
**Step 6: Alternative Form — Weighted Average** 
:::


We can also write the posterior mean as a **precision-weighted average** of prior mean and sample mean:

Let’s define:

- Prior precision: $P_0 = \frac{1}{\tau^2}$
- Data precision: $P_n = \frac{n}{\sigma^2}$

Then:
$$\mu_n = \frac{P_0 \cdot \theta + P_n \cdot \bar{y}}{P_0 + P_n}$$

This shows the posterior mean is a **weighted average** of the prior mean and the sample mean, where weights are the respective precisions.

***

::: {#big-text style="font-size: 125%"}
**Summary of Final Formulas** 
:::


| Quantity | Formula |
|----------|---------|
| **Posterior variance** | $\tau_n^2 = \dfrac{1}{\dfrac{1}{\tau^2} + \dfrac{n}{\sigma^2}}$ |
| **Posterior mean** | $\mu_n = \tau_n^2 \left( \dfrac{\theta}{\tau^2} + \dfrac{n \bar{y}}{\sigma^2} \right)$ |
| **Alternative mean** | $\mu_n = \dfrac{\dfrac{\theta}{\tau^2} + \dfrac{n \bar{y}}{\sigma^2}}{\dfrac{1}{\tau^2} + \dfrac{n}{\sigma^2}}$ |




::::
:::::

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-normal-normal-cacluation}
: Normal-Normal calculation
::::::
:::
::::{.my-exercise-container}



**a) Prof. Abebe** conducts the final exam and observes that his 32 students scored an average of 86 points. Calculate the posterior mean and variance of $\mu$ using the data from Prof. Abebe’s class.

Specify the posterior model of $\mu$:  

- our Normal prior model of $\mu$ had mean $\theta = 80$ and standard deviation $\tau = 4$; 
- our $n = 32$ students taking the final exam scored at average 86 points; 
- individual students’ scores $Y$ vary Normally around $\mu$ with a known standard deviation of 3 points.

***

::: {#big-text style="font-size: 150%"}
**Bayesian Updating for Prof. Abebe's Class** 
:::

::: {#big-text style="font-size: 125%"}
**Setting Up the Model** 
:::

This is a classic **Normal-Normal conjugate model**. We have:

| Component | Distribution | Mean | Std / Variance |
|-----------|-------------|------|----------------|
| Prior on $\mu$ | $\mu \sim \mathcal{N}(\theta, \tau^2)$ | $\theta = 80$ | $\tau = 4$, $\tau^2 = 16$ |
| Likelihood (individual scores) | $Y_i \mid \mu \sim \mathcal{N}(\mu, \sigma^2)$ | $\mu$ | $\sigma = 3$, $\sigma^2 = 9$ |
| Observed data | $n = 32$ students | $\bar{y} = 86$ | — |

***

::: {#big-text style="font-size: 125%"}
**The Normal-Normal Posterior Formula** 
:::


When the prior on $\mu$ is Normal and the likelihood is Normal with **known** $\sigma^2$, the posterior is also Normal (conjugacy). The general formulas are:

$$\mu \mid \bar{y} \sim \mathcal{N}\!\left(\mu_n,\, \tau_n^2\right)$$

**Posterior variance:**

$$\tau_n^2 = \frac{1}{\dfrac{1}{\tau^2} + \dfrac{n}{\sigma^2}}$$

**Posterior mean:**

$$\mu_n = \tau_n^2 \left(\frac{\theta}{\tau^2} + \frac{n\bar{y}}{\sigma^2}\right)$$

***

::: {#big-text style="font-size: 125%"}
**Step 1 — Compute the Posterior Variance** 
:::


The posterior precision is the **sum of the prior precision and the data precision**:



$$
\begin{align*}
\frac{1}{\tau_n^2} &= \frac{1}{\tau^2} + \frac{n}{\sigma^2} \\
&= \frac{1}{16} + \frac{32}{9} \\
&= 0.0625 + 3.5\overline{5} = 3.618\overline{05} \\
\tau_n^2 &= \frac{1}{3.618\overline{05}} \\
&=\boxed{ 0.276392}
\end{align*}
$$ {#eq-005-exr-exams-posterior-variance}


> **Intuition:** The posterior variance is *much* smaller than the prior variance ($\tau^2 = 16$), because 32 observations carry a lot of information, sharply concentrating our beliefs about $\mu$.

***

::: {#big-text style="font-size: 125%"}
**Step 2 — Compute the Posterior Mean** 
:::


The posterior mean is a **precision-weighted average** of the prior mean and the sample mean:

$$
\begin{align*}
&\mu_n = \tau_n^2 \left(\frac{\theta}{\tau^2} + \frac{n\bar{y}}{\sigma^2}\right) \\ &= 0.2764\left(\frac{80}{16} + \frac{32 \times 86}{9}\right) \\
&= 0.276422 \left(5 + 305.777778\right) \\ 
&= 0.276422 \times 310.777778 \\
&\boxed{\approx 85.8964}
\end{align*}
$$ {#eq-005-exr-exams-posterior-mean}


***

::: {#big-text style="font-size: 125%"}
**Step 3 — Intuition: Shrinkage Toward the Prior** 
:::


The posterior mean can be rewritten as a **weighted blend** of the prior mean and the data:

$$\mu_n = (1-w)\,\theta + w\,\bar{y}$$

where the weight on the data is:

$$w = \frac{n/\sigma^2}{1/\tau^2 + n/\sigma^2} = \frac{3.5\overline{5}}{3.618\overline{05}} \approx 0.9827$$

So:

$$
\begin{align*}
\mu_n &= ((1 - 0.9827255) \cdot 80) + (0.9827255 \cdot 86) \\
&= 1.38196 + 84.51439 \\
&=\boxed{ 85.89635}
\end{align*}
$$ {#eq-005-exr-shrinkage-toward-prior}

$$\mu_n = (1 - 0.9827)(80) + (0.9827)(86) = 0.1396 + 85.70 \approx 85.84 ✓$$

> **Intuition:** With 32 students and a tight likelihood ($\sigma=3$), the data dominate (~98.3% weight). The posterior mean of **85.84** is pulled only slightly from the observed $\bar{y} = 86$ back toward the prior mean of 80 — this *shrinkage* is a hallmark of Bayesian inference.

***

::: {#big-text style="font-size: 125%"}
**Summary** 
:::


| Quantity | Value |
|----------|-------|
| Prior mean $\theta$ | 80 |
| Observed sample mean $\bar{y}$ | 86 |
| **Posterior mean** $\mu_n$ | **85.89644** |
| **Posterior variance** $\tau_n^2$ | **0.276392** |
| Posterior std $\tau_n$ | ≈ 0.526 |


::: {#big-text style="font-size: 125%"}
**Bayesian updating with `bayesrules::summarize_normal_normal()`** 
:::

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-bayesian-updating-abebe}
: Bayesian updating for Prof. Abebe’s class
::::::
:::
::::{.my-r-code-container}

```{r}
#| label: tbl-005-exr-prof-abebe
#| tbl-cap: "Bayesian updating for Prof. Abebe’s class"


bayesrules::summarize_normal_normal(
  mean = 80, 
  sd = 4, 
  sigma = 3, 
  y_bar = 86, 
  n = 32) |> 
  knitr::kable(digits = 4)

```


::::
:::::

**Calculating manually and via bayesrules::summarize_normal_normal() gives the same results.**

***

**b) Prof. Morales** conducts the final exam and observes that her 32 students scored an average of 82 points. Calculate the posterior mean and variance of $\mu$ using the data from Prof. Morales’ class. (This time only with the {**bayesrules**} function.)

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-bayesian-updating-morales}
: Bayesian updating of Prof. Morales class
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: tbl-005-exr-prof-morales
#| tbl-cap: "Bayesian updating for Prof. Abebe’s class"


bayesrules::summarize_normal_normal(
  mean = 80, 
  sd = 4, 
  sigma = 3, 
  y_bar = 82, 
  n = 32) |> 
  knitr::kable(digits = 4)

```

::::
:::::

**d) Using Prof. Abebe and Prof. Morales’ combined exams** to calculate the posterior mean and variance of $\mu$.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-bayesian-updating-abebe-morales-combined}
: Bayesian updating of Prof. Abebe and Prof. Morales class combined
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: tbl-005-exr-prof-abebe-morales
#| tbl-cap: "Bayesian updating of Prof. Abebe’s and Morales’ class combined"


bayesrules::summarize_normal_normal(
  mean = 80, 
  sd = 4, 
  sigma = 3, 
  y_bar = mean(c(85, 82)), 
  n = 32 + 32) |> 
  knitr::kable(digits = 4)

```

::::
:::::

::::
:::::

#### Control brains

In this chapter we used the `r glossary("Normal-Normal Conjugate Family", "Normal-Normal model")` to analyze the mean hippocampal volume among people who have sustained concussions. In this exercise we will explore the mean volume, $\mu$, for people who *have not* been diagnosed with a concussion. As before, we’ll assume that the hippocampal volumes of individuals in this group, $Y$ , vary Normally around $\mu$ with a standard deviation of $\sigma = 0.5$ cubic centimeters. We’ll also start with the same prior as before: $\mu$ has a Normal prior with mean $\theta = 6.5$ and standard deviation $\tau = 0.4$.

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-exr-control-brains}
: Control brains
::::::
:::
::::{.my-exercise-container}

**a) Use the `football` data to calculate the sample mean hippocampal volume and sample size of the control subjects who have *not* been diagnosed with a concussion.**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-calculate-not-concussion}
: Sample mean hippocampal volume and sample size for no concussion
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: tbl-005-exr-calculate-not-concussion
#| tbl-cap: "Sample mean hippocampal volume and sample size for football player with no concussion"
#| results: hold

data("football", package = "bayesrules")

football_no_concuss <- football |> 
  dplyr::filter(group == "fb_no_concuss")

n <- base::nrow(football_no_concuss)
mean_no_concussion <- base::mean(football_no_concuss$volume)

glue::glue("**Sample size n = {n}; Mean = {mean_no_concussion}**")
  
```


::::
:::::

**b) Identify the posterior model of $\mu$ and verify your answer using bayesrules::summarize_normal_normal().**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-posterior-model-no-concussion}
: Posterior model of football player with have had no concussions
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: tbl-005-exr-posterior-model-no-concussion
#| tbl-cap: Prior & Posterior of football players with have had no concussions

bayesrules::summarize_normal_normal(
  mean = 6.5,
  sd = 0.4,
  sigma = 0.5,
  y_bar = mean_no_concussion,
  n = n
)
```

::::
:::::

**c) Plot the prior pdf, likelihood function, and posterior pdf of $\mu$. Describe the evolution in your understanding of $\mu$ from the prior to the posterior.**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-plot-posterior-no-concussion}
: Plot the prior pdf, likelihood function, and posterior pdf of $\mu$.
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-exr-plot-posterior-no-concussion
#| fig-cap: "Prior pdf, likelihood function, and posterior pdf of football players with no concussions"
#| fig-height: 3

bayesrules::plot_normal_normal(
  mean = 6.5,
  sd = 0.4,
  sigma = 0.5,
  y_bar = mean_no_concussion,
  n = n
)
```

::::
:::::

The hippocampal mean volume of football players with a concussion history is `r round(mean_hyppocampal_volume, 2)`$\text{cm}^3$ (see: @tbl-005-load-skim-football-data). This value is much smaller than the mean volume of `r round(mean_no_concussion, 2)`$\text{cm}^3$ for football players who did not have concussions.

The assumed hippocampal mean volume of the prior model taken from research literature is in the posterior analysis for our sample subjects only slightly smaller. It is also much more concentrated around the mean, so that we are pretty sure that no hippocampal mean volume for football players without concussions is under $6\text{cm}^3$.

::::
:::::

#### Australia {#sec-005-exr-australia}

Let $\mu$ be the average 3 p.m. temperature in Perth, Australia. Not knowing much about Australian weather, your friend’s prior understanding is that the average temperature is likely around 30 degrees Celsius, though might be anywhere between 10 and 50 degrees Celsius. To learn about $\mu$, they plan to analyze 1000 days of temperature data. Letting $Y_i$ denote the 3 p.m. temperature on day $i$, they’ll assume that daily temperatures vary Normally around $\mu$ with a standard deviation of 5 degrees: $Y_i \mid \mu \sim \mathcal{N} (\mu, 5^2)$.

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-australia}
: Weather in Australia
::::::
:::
::::{.my-exercise-container}

**a) Tune and plot a Normal prior for $\mu$ that reflects your friend’s understanding.**

**Prior calculation**: 

We have $\mu$ = 30°C but how to calculate the variance and standard deviation? 

Our friend says $\mu$ is **"likely around 30°C"** and **"might be anywhere between 10 and 50°C"**. We can translate this into a Normal prior by using the **empirical rule (`r glossary("rule-68–95–99", "68-95-99.7 rule")`)**.

The range of 10 to 50°C is **symmetric around 30°C**, spanning **±20°C** from the mean:

$$\theta = 30, \quad \text{range} = [10, 50]$$

The most natural interpretation of *"anywhere between 10 and 50"* is that this represents **virtually all** of the prior probability mass, i.e., **~4 standard deviations on each side (99.99% rule):**

$$
\begin{align*}
\theta \pm 4\sigma &= [10, 50] \\
4\sigma &= 20 \implies \boxed{\sigma = 5}
\end{align*}
$$ 

Depending on how strictly you read *"anywhere between"*, you could also use:

| Rule | Interpretation | Equation | $\sigma$ |
|------|---------------|----------|----------|
| $\pm 2\sigma$ | 95% of values | $2\sigma = 20$ | 10 |
| $\pm 3\sigma$ | 99.7% of values | $3\sigma = 20$ | 6.67 |
| **$\pm 4\sigma$** | **~99.99% of values** | **$4\sigma = 20$** | **5** |

The **$\pm 4\sigma$ rule ($\sigma = 5$)** is the standard choice here because it treats the stated range as containing **nearly all** plausible values, which best matches the phrase *"might be anywhere between"*.

So our prior model is 

$$\boxed{Y_i \mid \mu \sim \mathcal{N} (30, 5^2)}$$.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-prior-model-australia}
: Normal prior for $\mu$ reflecting our friend’s understanding of Australian weather
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-exr-prior-model-australia
#| fig-cap: "Normal prior for μ reflecting our friend’s understanding of Australian weather"
#| fig-height: 3

bayesrules::plot_normal(mean = 30, sd = 5)
```

::::
:::::

**b) The `weather_perth` data in the {bayesrules} package includes 1000 daily observations of 3 p.m. temperatures in Perth (`temp3pm`). Plot this data and discuss whether it’s reasonable to assume a Normal model for the temperature data.**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-weather-perth-normal-model}
: Temperature data at 3 p.m. in Perth (Australia)
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-exr-weather-perth-normal-model
#| fig-cap: "Temperature data at 3 p.m. in Perth (Australia)"
#| fig-height: 3

data("weather_perth", package = "bayesrules")

weather_perth |> 
  ggplot2::ggplot(ggplot2::aes(temp3pm)) +
    ggplot2::geom_density()
```


**mean_perth: `mean(weather_perth$temp3pm)` = `r mean(weather_perth$temp3pm)`**


::::
:::::

Even if the data is right-skewed and has a buckling on its right side it is reasonable to assume a Normal model for the temperature data. 

**c) Identify the posterior model of $\mu$ and verify your answer using bayesrules::summarize_normal_normal().**

::: {.callout-note #nte-005-exr-perth-weather-posterior-model}
###### Take the standard deviation from the data or from the model?

I had great problems to understand which value I should take for the standard deviation of the posterior model? At first I thought I should take `sd(weather_perth$temp3pm)` = `r sd(weather_perth$temp3pm)`. I saw a discrepancy arising because we're working with **two different sources of variability**: 


| Component | Standard Deviation | What It Measures |
|-----------|-------------------|-----------------|
| **Prior assumption** $\sigma$ | 5°C | Daily variation *around* the true mean $\mu$ |
| **Sample sd** | 5.8298°C | Observed variation in the 1000 daily temperatures |

After a conversation with `r glossary("Claude Haiku")` I understood that I have to take the **standard deviation of the model: sd = 5**

The **sample standard deviation of 5.8298°C** is an *estimate* of the **population standard deviation** $\sigma$ (the daily variation around the true mean). It differs from our assumed $\sigma = 5$ for several reasons:

1. **Sampling variability:** The sample sd is computed from 1000 observations and will naturally fluctuate around the true population value.

2. **Our assumption may be slightly off:** We assumed $\sigma = 5$ based on the model specification, but the actual daily variation in Perth temperatures might be slightly higher (5.83°C).

3. **The sample sd is not used in the posterior calculation:** In the Bayesian model with **known** $\sigma$, we treat $\sigma = 5$ as **fixed and given**, not estimated from the data. The posterior depends only on:
   - The prior ($\theta = 30$, $\tau = 5$)
   - The sample mean ($\bar{y} = 24.0009$)
   - The assumed likelihood sd ($\sigma = 5$, treated as known)

**The sample sd of 5.8298 is informative but serves a different purpose:**

- **If we were uncertain about $\sigma$:** We could use the sample sd to estimate it or construct a credible interval for $\sigma$.
- **For model checking:** A sample sd of 5.83 vs. our assumed 5 suggests the model assumption is reasonable but slightly conservative (the data show slightly more daily variation than assumed).
- **For sensitivity analysis:** We could re-run the posterior with $\sigma = 5.83$ to see how robust our conclusions are.


1. The posterior calculation has to be $\sigma = 5$ because it is a **model assumption** (the likelihood standard deviation), not something we estimate from the data. 
2. The sample sd of 5.8298 is a **descriptive statistic** that validates our assumption is in the right ballpark, but it doesn't change the posterior under the Normal-Normal conjugate model with known $\sigma$.

So the answer which sd to use lies in the provided **model specification**:



> “… they’ll assume that daily temperatures vary Normally around $\mu$ with a standard deviation of 5 degrees: $Y_i \mid \mu \sim \mathcal{N}(\mu, 5^2)$.”

This means:  
✅ **$\sigma = 5$ is a fixed, known parameter** in the model — **not estimated from the data**.


In Bayesian inference, when we say “known $\sigma$”, we are assuming:

- The variability of daily temperatures around the true mean $\mu$ is **exactly 5°C**.
- We are **not using the sample standard deviation (5.8298)** to estimate $\sigma$.
- The posterior for $\mu$ is computed **conditional on this fixed $\sigma = 5$**.

This is called the **Normal-Normal conjugate model with known variance** — a standard textbook setup.

The sample standard deviation of 5.8298 is:

- An **empirical estimate** of the population $\sigma$ based on 1000 observations.
- A **descriptive statistic** — it tells us what the data *actually* show.
- **Not used in the posterior calculation** under the assumption that $\sigma$ is known.

We could use sd = 5.8298 instead --- but that would be a **different model**:

- We would need to treat $\sigma$ as **unknown** and estimate it (e.g., using a Normal-Gamma model).
- Or, we could do a **sensitivity analysis** and recompute the posterior with $\sigma = 5.8298$ to see how much it changes.

But **under the model that was specified**, $\sigma = 5$ is fixed — so we use it.


:::


Taking the result of @nte-005-exr-perth-weather-posterior-model we get for the posterior model of $\mu$:  

- our Normal prior model of $\mu$ had mean $\theta = 30$ and standard deviation $\tau = 5$; 
- our $n = 1000$ data points shows an average temperature at Perth of 24,0009 degrees. 
- individual temperature $Y$ vary Normally around $\mu$ with a standard deviation of 5 degrees.

***

**Manual calculation of the posterior standard deviation** 


$$
\begin{align*}
\frac{1}{\tau_n^2} &= \frac{1}{\tau^2} + \frac{n}{\sigma^2} \\
&= \frac{1}{5^2} + \frac{1000}{5^2} \\
&= 0.04 + 40 = 40.04 \\
\tau_n^2 &= \frac{1}{40.04} \\
&=\boxed{0.02497502} \\
\tau_n &= \boxed{0.1580349}
\end{align*}
$$ {#eq-005-exr-perth-weather-posterior-variance}


***

**Manual calculation of the posterior mean** 


$$
\begin{align*}
&\mu_n = \tau_n^2 \left(\frac{\theta}{\tau^2} + \frac{n\bar{y}}{\sigma^2}\right) \\ &= 0.02497502\left(\frac{30}{25} + \frac{1000 \times 24.0009}{25}\right) \\
&= 0.02497502 \left(1.2 + 960.036\right) \\ 
&= 0.02497502 \times 961.236 \\
&\boxed{\approx 24.00689}
\end{align*}
$$ {#eq-005-exr-perth-weather-posterior-mean}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-verify-calculation-mena-variance}
: Verify calculation of posterior mean and variance using `bayesrules::summarize_normal_normal()`
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: tbl-005-exr-verify-calculation-mean-variance
#| tbl-cap: Posterior mean and variance of temperature in Perth at 3 p.m.

bayesrules::summarize_normal_normal(
  mean = 30,
  sd = 5,
  sigma = 5,
  y_bar = mean(weather_perth$temp3pm),
  n = 1000
) |> 
  knitr::kable()


```


::::
:::::

In the manual calculation and in using the `bayesrules::summarize_normal_normal()`function we got exactly the same values!


**d) Plot the prior pdf, likelihood function, and posterior pdf of μ. Describe the evolution in your understanding of μ from the prior to the posterior.**

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-plot-perth-weather-posterior}
: Perth weather prior pdf, likelihood function and posterior pdf of $\mu$
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: fig-005-exr-plot-perth-weather-posterior
#| fig-cap: The prior pdf, likelihood function and posterior pdf of $\mu$ for the waether in Perth (Australia)
#| fig-height: 3

bayesrules::plot_normal_normal(
  mean = 30,
  sd = 5,
  sigma = 5,
  y_bar = mean(weather_perth$temp3pm),
  n = 1000
)
```

::::
:::::

**Dramatic Shift from Prior**

The summary and the plot not only show how dramatically the data shifted our beliefs from the prior mean of 30°C down to a posterior mean of ~24°C but also that we are pretty sure of the mean temperature.

Prior: $\mu \sim \mathcal{N}(30, 25)$ — centered at 30°C with wide uncertainty
Posterior: $\mu \sim \mathcal{N}(24.01, 0.0250)$ — centered at 24°C with very tight uncertainty
The data revealed that Perth's average 3 p.m. temperature is ~6°C cooler than our friend initially believed.

Extreme Precision
The posterior standard deviation is only 0.158°C this is an extreme precision, meaning we're now 95% confident that $\mu$ lies in:

$$24.01 \pm 1.96(0.158) = 24.01 \pm 0.31 = [23.70, 24.32] \text{ degrees}$$

With 1000 observations, the sample mean (24.0009) nearly equals the posterior mean (24.01), with only minimal shrinkage toward the prior mean of 30. 

The weight on the data is ~99.9%, while the prior receives ~0.1% weight!

:::::{.my-theorem}
:::{.my-theorem-header}
:::::: {#thm-005-exr-calculate-weight-and-data-prior}
: Calculating the weight of the data in comparison with the prior pdf
::::::
:::
::::{.my-theorem-container}

::: {#big-text style="font-size: 110%"}
**The Weighted Average Formula**
:::



Recall from the derivation that the posterior mean can be written as a **precision-weighted average**:

$$
\begin{align*}
\mu_n = (1-w)\,\theta + w\,\bar{y} \\
\text{where w is the weight on the data} \\
w = \frac{n/\sigma^2}{1/\tau^2 + n/\sigma^2}
\end{align*}
$$ {#eq-005-exr-precision-weighted-average}


::: {#big-text style="font-size: 110%"}
**Plugging In the Numbers** 
:::


**Numerator** (data precision):
$$\frac{n}{\sigma^2} = \frac{1000}{25} = 40$$

**Denominator** (total precision = prior precision + data precision):
$$\frac{1}{\tau^2} + \frac{n}{\sigma^2} = \frac{1}{25} + \frac{1000}{25} = \frac{1001}{25} = 40.04$$

**Weight on the data:**
$$w = \frac{40}{40.04} \approx 0.99900 \approx 99.9\%$$

**Weight on the prior:**
$$1 - w = \frac{0.04}{40.04} \approx 0.00100 \approx 0.1\%$$

::: {#big-text style="font-size: 110%"}
**The weight of the data in comparison with the prior pdf for the weather in Perth** 
:::

| Component | Weight |
|-----------|--------|
| **Data** $\bar{y}$ | **~99.9%** |
| **Prior** $\theta$ | **~0.1%** |

**Verification**

$$
\begin{align*}
\mu_n &= (1-w)\,\theta + w\,\bar{y} \\
&= 0.001(30) + 0.999(24.0009) \\
&= 0.030 + 23.977 \\
&= 24.007 ✓
\end{align*}
$$ {#eq-005-exr-verification-precision-weighted-average}

This confirms that with 1000 observations and equal $\tau = \sigma = 5$, the data almost completely dominate the posterior.


::::
:::::

::::
:::::

::: {.callout-note #nte-005-exr-bayesian-advantages}

###### When Bayesian Inference Shines



Given the dominate role data plays the question arose for me: What is the advantage of using the Bayesian approach? Frequentist statistics would get almost the same result.

::: {#big-text style="font-size: 140%"}
**Advantages of the Bayesian approach** 
:::

::: {#big-text style="font-size: 120%"}
**1. Small Sample Sizes** 
:::


With only **10 or 50 days** of data instead of 1000, the prior would carry meaningful weight:

**Example: $n = 50$ days**

$$w = \frac{50/25}{1/25 + 50/25} = \frac{2}{2.04} \approx 0.98 \text{ (98% data)}$$

Weight on prior: **~2%** — now the prior actually matters!

The Bayesian posterior would be:
$$\mu_n \approx 0.02(30) + 0.98(\bar{y})$$

This **regularizes** the estimate when data are scarce, preventing wild swings from small samples. Frequentist confidence intervals would be much wider and less informative.

::: {#big-text style="font-size: 120%"}
**2. Incorporating Prior Knowledge** 
:::

Our friend's prior belief (30°C, sd = 5) encodes real information:
- Perth is in the Southern Hemisphere (warmer than many places)
- 3 p.m. is afternoon (warmer than morning)
- The range [10, 50] reflects reasonable bounds

**Frequentist approach:** Ignores this entirely. If the first 50 observations happened to average 18°C due to sampling variability, the frequentist estimate would be 18°C with no skepticism.

**Bayesian approach:** Would say "18°C is implausibly cold given what we know about Perth" and pull the estimate back toward 30°C.

::: {#big-text style="font-size: 120%"}
**3. Decision-Making Under Uncertainty** 
:::

Suppose your friend needs to decide: *"Should I bring a light jacket (assume μ ≥ 22°C) or a heavy coat (assume μ < 22°C)?"*

**Bayesian:** Directly answers: $P(\mu \geq 22 \mid \text{data}) = ?$

With posterior $\mu \mid \text{data} \sim \mathcal{N}(24.01, 0.158^2)$:
$$P(\mu \geq 22 \mid \text{data}) = P\left(Z \geq \frac{22 - 24.01}{0.158}\right) = P(Z \geq -12.7) \approx 1.0$$

**Frequentist:** Cannot directly answer this. A 95% CI is [23.70, 24.32], but that doesn't directly tell you $P(\mu \geq 22)$ in a frequentist framework.

::: {#big-text style="font-size: 120%"}
**4. Hierarchical/Complex Models** 
:::


With multiple professors, multiple cities, or temporal trends, Bayesian methods naturally handle:

- Sharing information across groups
- Partial pooling of estimates
- Uncertainty propagation through layers

Frequentist approaches become awkward or require ad-hoc methods.

::: {#big-text style="font-size: 120%"}
**5. Sequential Analysis** 
:::

Suppose you observe data on Day 1, then Day 2, then Day 3...

**Bayesian:** Today's posterior becomes tomorrow's prior. You naturally update beliefs as new data arrive — no multiple testing corrections needed.

**Frequentist:** Each new test inflates Type I error; you need corrections (Bonferroni, etc.), making inference less efficient.

::: {#big-text style="font-size: 140%"}
**Why This Exercise Didn't Show Bayesian Advantage** 
:::

In the Perth temperature problem:

- ✅ Large sample size ($n = 1000$) → data overwhelm the prior
- ✅ Prior and likelihood have equal information ($\tau = \sigma$)
- ✅ Simple one-parameter model
- ✅ No sequential or hierarchical structure

In this setting, Bayesian and frequentist methods converge — and that's **expected and fine**.

**Bayesian inference adds the most value when:**

- Data are limited
- Prior knowledge is informative and relevant
- You need probability statements about parameters (not just intervals)
- You're making sequential decisions
- You have complex hierarchical structures

**When data are abundant and prior is weak, Bayesian and frequentist results converge** — which is reassuring, not a weakness!
:::




#### Normal-Normal Simulation

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-exr-normal-normal-simulation}
: Normal-Normal Simulation
::::::
:::
::::{.my-exercise-container}

**a) Your friend Alex has read @sec-chap-004 of this book, but not this chapter. Explain to Alex why it’s difficult to simulate a Normal-Normal posterior using the simulation methods we have learned thus far.**

Simulating the Normal-Normal example with the methods learned so far is not possible. We have seen in @prp-005-steps-of-simulations that we must follow four steps to simulate the Normal-Normal model. The problem arises with step 3: 

> Filter out only those of the $10000$ simulated sets of $(θ, Y )$ for which the simulated $Y$ data matches the data we actually observed.

This doesn't work with the Normal-Normal model: We won't have any match, because our data values $Y_i$ are continuous!

**b) To prove your point, try (and fail) to simulate the posterior of μ for the following model upon observing a single data point $Y = 1.1$:

$$
\begin{align*}
Y \mid \mu &\sim \mathcal{N} (\mu, 1^2) \\
\mu &\sim \mathcal{N} (0.1^2)
\end{align*}
$$ {#eq-005-exr-normal-normal-example}

::: {#big-text style="font-size: 125%"}
**Why This Simulation Method Fails for Normal-Normal Models** 
:::


The fundamental problem is **the curse of continuous data**: when you simulate data from a continuous distribution, the probability of generating data that *exactly matches* your observed value is essentially **zero**.

::: {#big-text style="font-size: 110%"}
**The Core Issue** 
:::


In step 3, you're filtering for simulated $Y$ values that match your observed $Y = 1.1$ *exactly*. But since $Y$ is continuous:

$$P(Y_{\text{simulated}} = 1.1 \text{ exactly}) \approx 0$$

This means virtually **none** of your 10,000 simulated ($\theta$, $Y$) pairs will pass the filter, leaving you with an empty (or nearly empty) posterior sample.

***

::: {#big-text style="font-size: 110%"}
**Demonstration: Attempting to Simulate the Posterior** 
:::

Let me show why this fails with the following model:

**Prior:** $\mu \sim \mathcal{N}(0, 0.1^2)$  
**Likelihood:** $Y \mid \mu \sim \mathcal{N}(\mu, 1^2)$  
**Observed:** $Y = 1.1$

:::::{.my-assessment}
:::{.my-assessment-header}
:::::: {#cor-005-exr-simulate-posterior}
: Simulate the Posterior pdf 
::::::
:::
::::{.my-assessment-container}

- **Step 1**: Simulate 10,000 theta values from prior
- **Step 2**: Simulate data Y from each theta
- **Step 3**: Filter for exact match to observed Y = 1.1
- **Step 4**: Extract posterior theta values
::::
:::::



**R Code (tidyverse style)**

```{r}
#| label: tbl-005-failed-simulation-1
#| tbl-cap: "Failed simulation with 0 matches"
#| results: hold

base::set.seed(84735)
Y_observed <- 1.1
n_sims <- 10000

posterior_attempt <- tibble::tibble(
  sim_id = 1:n_sims,
# Step 1: Simulate 10,000 theta values from prior ###########
  theta = stats::rnorm(n_sims, mean = 0, sd = 0.1)
  ) |> 
  dplyr::mutate(
# Step 2: Simulate data Y from each theta ###################
    Y_simulated = stats::rnorm(n_sims, mean = theta, sd = 1.0)
  )  |> 
  
# Step 3: Filter for exact match to observed Y = 1.1 ########  
  dplyr::filter(Y_simulated == Y_observed) |> 
  dplyr::select(theta)


# Step 4: Extract posterior theta values ############
base::cat("Number of matches:", nrow(posterior_attempt), "\n")
base::cat("Posterior sample size:", nrow(posterior_attempt), "\n")
```

**You get nothing.** Even with 10,000 simulations, you'll have zero (or perhaps one or two by extreme luck) matches.

::: {#big-text style="font-size: 110%"}
**Why This Happens Mathematically** 
:::


The probability of a match in any single simulation is:

$$P(\text{match}) = \int P(Y_{\text{sim}} = 1.1 \mid Y_{\text{sim}} \sim \mathcal{N}(\theta, 1)) \, dP(\theta)$$

For continuous distributions, this integral is **zero** because the probability density at a single point is infinitesimal.

Even with a small tolerance (e.g., $|Y_{\text{sim}} - 1.1| < 0.01$), you'd need an astronomically large number of simulations to get a reasonable posterior sample.

::: {#big-text style="font-size: 110%"}
**Attempting a Fix: Adding a Tolerance Window** 
:::

To illustrate the problem more clearly, here's what happens if you add a small tolerance:

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-005-exr-unreliable-simulation}
: Unreliable simulation with only few matches
::::::
:::
::::{.my-r-code-container}

```{r}
#| label: tbl-005-failed-simulation-2
#| tbl-cap: "Unreliable simulation with just a few matches"
#| results: hold

base::set.seed(100)
Y_observed <- 1.1
n_sims <- 10000
tolerance <- 0.01  # Accept values within ±0.01

posterior_with_tolerance <- tibble::tibble(
  sim_id = 1:n_sims,
# Step 1: Simulate 10,000 theta values from prior ##########
  theta = stats::rnorm(n_sims, mean = 0, sd = 0.1)
  ) |> 
  dplyr::mutate(
# Step 2: Simulate data Y from each theta ###################
    Y_simulated = stats::rnorm(n_sims, mean = theta, sd = 1.0)
  ) |> 
# Step 3: Filter for exact match to observed Y = 1.1# tolerance  ########  
  dplyr::filter(abs(Y_simulated - Y_observed) < tolerance) |> 
  dplyr::select(theta)

# Step 4a: Number of posterior theta values ############
base::cat("Number of matches with tolerance:", nrow(posterior_with_tolerance), "\n")

# Step 4a: Extract posterior theta values ############
if (base::nrow(posterior_with_tolerance) > 0) {
  posterior_with_tolerance |> 
    dplyr::summarize(
      mean_theta = mean(theta),
      sd_theta = sd(theta),
      .groups = "drop"
    ) |> 
    print()
}

```

::::
:::::


Even with a tolerance of 0.01, you'll likely get only few matches (30-50 matches), making the posterior estimate unreliable.

**There are three different solutions: For details see: @sec-910-solution-continuous-distribution}** 



::::
:::::


### General Practice Exercises

#### Which model?

:::::{.my-assessment}
:::{.my-assessment-header}
:::::: {#cor-005-exr-which-model}
: Which model?
::::::
:::
::::{.my-assessment-container}
Below are kernels for Normal, Poisson, Gamma, Beta, and Binomial models. Identify the appropriate model with specific parameter values.

$$
\begin{align*}
\text{a)}\quad f (\theta) &\propto 0.3^{\theta}0.7^{16− \theta} \text{ for } \theta \in \{0, 1, 2, \cdots, 16\} \\
\text{b)}\quad f (\theta) &\propto 1/\theta! \text{ for } \theta \in \{0, 1, 2, \cdots, \infty \} \\
\text{c)}\quad f (\theta) &\propto \theta^4(1 − \theta)^7 \text{ for } \theta \in [0, 1] \\
\text{d)}\quad f (\theta) &\propto e^{−θ^2} \text{ for } \theta \in (−\infty, \infty) 
\end{align*}
$$ {#eq-005-exr-which-model}


Which model represents the kernel form of @eq-005-exr-which-model?

a) `r mcq(c("Normal", "Poisson", "Gamma", "Beta", answer = "Binomial"))`
b) `r mcq(c("Normal", answer = "Poisson", "Gamma", "Beta", "Binomial"))`
c) `r mcq(c("Normal", "Poisson", "Gamma", answer = "Beta", "Binomial"))`
d) `r mcq(c(answer = "Normal", "Poisson", "Gamma", "Beta", "Binomial"))`

::::
:::::

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-which-model}
: Identify the appropriate model from the kernels in @eq-005-exr-which-model with specific parameter values
::::::
:::
::::{.my-exercise-container}



The key is to match each kernel to a known distribution's **kernel form** (dropping constants).

::: {#big-text style="font-size: 125%"}
**Quick Reference: Kernel Forms for different models** 
:::

| Distribution | Full PDF | Kernel |
|---|---|---|
| **Binomial$(n,p)$** | $\binom{n}{\theta} p^\theta (1-p)^{n-\theta}$ | $p^\theta (1-p)^{n-\theta}$ |
| **Poisson$(\lambda)$** | $\frac{e^{-\lambda}\lambda^\theta}{\theta!}$ | $\frac{\lambda^\theta}{\theta!}$ |
| **Beta$(\alpha,\beta)$** | $\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\theta^{\alpha-1}(1-\theta)^{\beta-1}$ | $\theta^{\alpha-1}(1-\theta)^{\beta-1}$ |
| **Normal$(\mu,\sigma^2)$** | $\frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{(\theta-\mu)^2}{2\sigma^2}}$ | $e^{-\frac{(\theta-\mu)^2}{2\sigma^2}}$ |
| **Gamma$(\alpha,\beta)$** | $\frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{\alpha-1}e^{-\beta\theta}$ | $\theta^{\alpha-1}e^{-\beta\theta}$ |
: Kernel Forms for different models {#tbl-005-exr-kernel-forms}

***

::: {#big-text style="font-size: 125%"}
**Part a)** 
:::


$$f(\theta) \propto 0.3^{\theta} \cdot 0.7^{16-\theta}, \quad \theta \in \{0,1,\ldots,16\}$$

- Discrete support with finite upper bound → **Binomial**
- Match to kernel $p^\theta(1-p)^{n-\theta}$:
  - $p = 0.3$, $1-p = 0.7$ ✓
  - $n = 16$ ✓

$$\boxed{\theta \sim \text{Binomial}(n=16,\ p=0.3)}$$

::: {#big-text style="font-size: 125%"}
**Part b)** 
:::

$$f(\theta) \propto \frac{1}{\theta!}, \quad \theta \in \{0,1,2,\ldots\}$$

- Discrete support with infinite upper bound → **Poisson**
- Match to kernel $\frac{\lambda^\theta}{\theta!}$:
  - $\lambda^\theta$ is missing, meaning $\lambda^\theta = 1 \Rightarrow \lambda = 1$

$$\boxed{\theta \sim \text{Poisson}(\lambda=1)}$$

::: {#big-text style="font-size: 125%"}
**Part c)** 
:::

$$f(\theta) \propto \theta^4(1-\theta)^7, \quad \theta \in [0,1]$$

- Continuous support on $[0,1]$ → **Beta**
- Match to kernel $\theta^{\alpha-1}(1-\theta)^{\beta-1}$:
  - $\alpha - 1 = 4 \Rightarrow \alpha = 5$
  - $\beta - 1 = 7 \Rightarrow \beta = 8$

$$\boxed{\theta \sim \text{Beta}(\alpha=5,\ \beta=8)}$$

::: {#big-text style="font-size: 125%"}
**Part d)** 
:::

$$f(\theta) \propto e^{-\theta^2}, \quad \theta \in (-\infty, \infty)$$

- Continuous support on $\mathbb{R}$ → **Normal**
- Match to kernel $e^{-\frac{(\theta - \mu)^2}{2\sigma^2}}$:
  - $\mu = 0$ (no linear term in the exponent)
  - $\frac{1}{2\sigma^2} = 1 \Rightarrow \sigma^2 = \frac{1}{2}$

$$\boxed{\theta \sim \mathcal{N}\!\left(\mu=0,\ \sigma^2=\frac{1}{2}\right)}$$

::::
:::::

#### Which model: Back for more!




:::::{.my-assessment}
:::{.my-assessment-header}
:::::: {#cor-005-exr-which-model-2}
: Which model: Back for more!
::::::
:::
::::{.my-assessment-container}
Below are kernels for Normal, Poisson, Gamma, Beta, and Binomial models. Identify the appropriate model with specific parameter values.

$$
\begin{align*}
\text{a)}\quad f (\theta) &\propto e^{−2θ}θ^{15} \text{ for } \theta > 0\\
\text{b)}\quad f (\theta) &\propto e \frac{−(θ−12)^2}{18} \text{ for } \theta \in (-\infty, \infty) \\ 
\text{c)}\quad f (\theta) &\propto 0.3^{θ}/θ! \text{ for } \theta \in \{0, 1, 2, \cdots, \infty \}
\end{align*}
$$ {#eq-005-exr-which-model-2}

Which model represents the kernel form of @eq-005-exr-which-model-2?

a) `r mcq(c("Normal", "Poisson", answer = "Gamma", "Beta", "Binomial"))`
b) `r mcq(c(answer = "Normal", "Poisson", "Gamma", "Beta", "Binomial"))`
c) `r mcq(c("Normal", answer = "Poisson", "Gamma", "Beta", "Binomial"))`


::::
:::::


:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-which-model-2}
: Which model: Back for more!
::::::
:::
::::{.my-exercise-container}

Again the key is to match each kernel to a known distribution’s kernel form by dropping the constants. See @tbl-005-exr-kernel-forms.

::: {#big-text style="font-size: 125%"}
**Part a)** 
:::


$$f(\theta) \propto e^{-2\theta}\theta^{15}, \quad \theta > 0$$

- Continuous support on $(0, \infty)$ → **Gamma**
- Match to kernel $\theta^{\alpha-1}e^{-\beta\theta}$:
  - $\alpha - 1 = 15 \Rightarrow \alpha = 16$
  - $\beta = 2$

$$\boxed{\theta \sim \text{Gamma}(\alpha=16,\ \beta=2)}$$

::: {#big-text style="font-size: 125%"}
**Part b)** 
:::

$$f(\theta) \propto e^{-\frac{(\theta-12)^2}{18}}, \quad \theta \in (-\infty, \infty)$$

- Continuous support on $\mathbb{R}$ → **Normal**
- Match to kernel $e^{-\frac{(\theta-\mu)^2}{2\sigma^2}}$:
  - $\mu = 12$ (center of the exponent)
  - $\frac{1}{2\sigma^2} = \frac{1}{18} \Rightarrow 2\sigma^2 = 18 \Rightarrow \sigma^2 = 9$

$$\boxed{\theta \sim \mathcal{N}(\mu=12,\ \sigma^2=9)}$$

::: {#big-text style="font-size: 125%"}
**Part c)** 
:::


$$f(\theta) \propto \frac{0.3^{\theta}}{\theta!}, \quad \theta \in \{0, 1, 2, \ldots\}$$

- Discrete support with infinite upper bound → **Poisson**
- Match to kernel $\frac{\lambda^\theta}{\theta!}$:
  - $\lambda = 0.3$

$$\boxed{\theta \sim \text{Poisson}(\lambda=0.3)}$$

::::
:::::


#### Weighted average

In Section 4.3.2, we expressed the Beta-Binomial posterior mean of proportion $\pi$ as a weighted average of the prior mean and the observed sample proportion. In @sec-005-normal-normal-conjugacy, we expressed the Normal-Normal posterior mean of $\mu$ as a weighted average of the prior mean $\theta$ and the observed sample mean $y$. Similarly, express the Gamma-Poisson posterior mean of rate $\lambda, E(\lambda \mid \overrightarrow{\rm y})$, as a weighted average of the prior mean $E(\lambda)$ and sample mean $y$.

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-gamma-poisson-average}
: Gamma-Poisson posterior mean as weighted average
::::::
:::
::::{.my-exercise-container}


::: {#big-text style="font-size: 125%"}
**Setup** 
:::

- **Prior:** $\lambda \sim \text{Gamma}(\alpha, \beta)$  
- **Likelihood:** $Y_i \mid \lambda \sim \text{Poisson}(\lambda)$ for $i = 1, \ldots, n$  
- **Posterior:** $\lambda \mid \vec{y} \sim \text{Gamma}(\alpha + \sum y_i, \beta + n)$

::: {#big-text style="font-size: 125%"}
**Posterior mean** 
:::

The posterior mean is:

$$E(\lambda \mid \vec{y}) = \frac{\alpha + \sum_{i=1}^n y_i}{\beta + n}$$

Rearranging:

$$E(\lambda \mid \vec{y}) = \frac{\alpha}{\beta + n} + \frac{\sum_{i=1}^n y_i}{\beta + n}$$

$$E(\lambda \mid \vec{y}) = \frac{\beta}{\beta + n} \cdot \frac{\alpha}{\beta} + \frac{n}{\beta + n} \cdot \frac{\sum_{i=1}^n y_i}{n}$$

::: {#big-text style="font-size: 125%"}
**Weighted Average Form** 
:::


$$\boxed{E(\lambda \mid \vec{y}) = \frac{\beta}{\beta + n} \cdot E(\lambda) + \frac{n}{\beta + n} \cdot \bar{y}}$$

where:

- **$E(\lambda) = \frac{\alpha}{\beta}$** = prior mean
- **$\bar{y} = \frac{1}{n}\sum_{i=1}^n y_i$** = sample mean
- **$\frac{\beta}{\beta + n}$** = prior weight
- **$\frac{n}{\beta + n}$** = data weight

::: {#big-text style="font-size: 125%"}
**Interpretation** 
:::

The posterior mean is a **convex combination** of:

1. **Prior mean** ($E(\lambda)$), weighted by $\frac{\beta}{\beta + n}$
2. **Sample mean** ($\bar{y}$), weighted by $\frac{n}{\beta + n}$

**Key insights:**

- As $n \to \infty$: posterior mean → sample mean (data dominates)
- As $\beta \to \infty$: posterior mean → prior mean (strong prior belief)
- As $\beta \to 0$: posterior mean → sample mean (weak prior)

::::
:::::

#### Counting insects

A biologist hires you to study the density of a certain insect in the local region. Their prior idea about $\theta$, the number of insects per $m^2$ area, is well captured by a `r glossary("Gamma-Model", "Gamma model")` with an expected value of 0.5 insects per $m^2$ and a standard deviation of 0.25 insects per $m^2$ You then go out to inspect 20 separate $m^2$ areas of a nearby field. You count 3, 2, 5, 1, 2 insects in the first five square meters, and then do not encounter the insect in the next $15m^2$.

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-gamma-from-mean-and-sd}
: Tuning Gamma model given mean and standard deviation
::::::
:::
::::{.my-exercise-container}


::: {#big-text style="font-size: 125%"}
**Given Information** 
:::


- **Prior distribution:** $\theta \sim \text{Gamma}(s, r)$
- **Expected value:** $E[\theta] = 0.5$
- **Standard deviation:** $\text{SD}[\theta] = 0.25$


***


::: {#big-text style="font-size: 125%"}
**Gamma Distribution Properties** 
:::

See @eq-005-gamma-central-measures in @sec-005-gamma-formulas:

For $\theta \sim \text{Gamma}(r, s)$:

$$
\begin{align*}
E[\theta] &= \frac{s}{r} \\
\text{Var}[\theta] &= \frac{s}{r^2} \\
\text{SD}[\theta] &= \sqrt{\frac{s}{r^2}} = \frac{\sqrt{s}}{r}
\end{align*}
$$ 
***



::: {#big-text style="font-size: 125%"}
**Solving for $s$ and $r$** 
:::

::: {#big-text style="font-size: 110%"}
**From the variance relationship:** 
:::


$$\text{Var}[\theta] = (\text{SD}[\theta])^2 = (0.25)^2 = 0.0625$$

Using $\text{Var}[\theta] = \dfrac{s}{r^2}$:

$$\frac{s}{r^2} = 0.0625 \quad \Rightarrow \quad s = 0.0625 r^2 \quad \cdots (1)$$


::: {#big-text style="font-size: 110%"}
**From the mean relationship:** 
:::

$$\frac{s}{r} = 0.5 \quad \Rightarrow \quad s = 0.5r \quad \cdots (2)$$

::: {#big-text style="font-size: 110%"}
**Equate (1) and (2):** 
:::

$$0.0625 r^2 = 0.5r$$

$$0.0625 r = 0.5$$

$$r = \frac{0.5}{0.0625} = 8$$

::: {#big-text style="font-size: 110%"}
**Solve for $r$:** 
:::


$$s = 0.5 \cdot r = 0.5 \cdot 8 = 4$$

***

::: {#big-text style="font-size: 125%"}
**Answer** 
:::


$$\boxed{s = 4, \quad r = 8}$$

::: {#big-text style="font-size: 125%"}
**Verification** 
:::


- $E[\theta] = \dfrac{4}{8} = 0.5$ ✓
- $\text{SD}[\theta] = \dfrac{\sqrt{4}}{8} = \dfrac{2}{8} = 0.25$ ✓

So the biologist's prior is: $\theta \sim \text{Gamma}(4, 8)$

```{r}
#| label: fig-005-exr-plot-gamma-prior-insects
#| fig-cap: Plotting the biologist’s Gamma prior for the insect study
#| fig-height: 3
#| fig-width: 4

my_plot_gamma(4, 8, TRUE, TRUE)
```



```{r}
#| label: tbl-005-exr-summarize-insect-posterior
#| tbl-cap: Summarizing the biologist’s Gamma-Poisson posterior for the insect study

bayesrules::summarize_gamma_poisson(
  shape = 4,
  rate = 8,
  sum_y = sum(c(3, 2, 5, 1, 2)),
  n = 20
) |> 
  knitr::kable(digits = 4)
```



```{r}
#| label: fig-005-exr-summarize-insect-posterior
#| fig-cap: The biologist’s Gamma-Poisson posterior for the insect study
#| fig-height: 3

bayesrules::plot_gamma_poisson(
  shape = 4,
  rate = 8,
  sum_y = sum(c(3, 2, 5, 1, 2)),
  n = 20
)
```

::: {#big-text style="font-size: 125%"}
**Interpretation** 
:::

- The posterior mean of 0.607 insects/m² is pulled upward from the prior mean (0.5) toward the data signal.
- The posterior SD of 0.147 is notably smaller than the prior SD (0.250), reflecting the reduction in uncertainty after observing 20 areas.

::::
:::::

#### Beta-Geometric model

Consider the following new Bayesian model: 

$$
\begin{align*}
Y \mid \theta &\sim \text{Geometric}(\theta) \\
\theta &\sim \text{Beta}(\alpha, \beta) 
\end{align*}
$$ 
  
where the Geometric model has pmf $f (y \mid \theta) = \theta(1 − θ)y−1 \text{ for } y \in \{1, 2, \cdots\}$.

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-005-beta-geometric-model}
: Beta-Geometric model
::::::
:::
::::{.my-exercise-container}

::: {#big-text style="font-size: 125%"}
**a) Derive the posterior model for $\theta$ given observed data $Y = y$. If possible, identify the name of the posterior model and its parameters.**
:::



::: {#big-text style="font-size: 110%"}
**Bayes’ Theorem Setup** 
:::

The posterior is proportional to the likelihood times the prior:

$$p(\theta \mid y) \propto f(y \mid \theta) \cdot p(\theta)$$

::: {#big-text style="font-size: 110%"}
**Likelihood** 
:::

Given $Y = y$ with the Geometric distribution:

$$f(y \mid \theta) = \theta(1-\theta)^{y-1}$$

::: {#big-text style="font-size: 110%"}
**Prior** 
:::

The Beta prior:

$$p(\theta) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} \theta^{\alpha-1}(1-\theta)^{\beta-1}$$

::: {#big-text style="font-size: 110%"}
**Deriving the Posterior** 
:::


Applying Bayes' theorem (keeping only terms involving $\theta$):

$$p(\theta \mid y) \propto \theta(1-\theta)^{y-1} \cdot \theta^{\alpha-1}(1-\theta)^{\beta-1}$$

Combining the powers of $\theta$ and ($1-\theta$):

$$p(\theta \mid y) \propto \theta^{1+\alpha-1}(1-\theta)^{y-1+\beta-1}$$

$$p(\theta \mid y) \propto \theta^{\alpha}(1-\theta)^{y+\beta-2}$$

::: {#big-text style="font-size: 110%"}
**Posterior Model** 
:::


This is the kernel of a **Beta distribution**. Specifically:

$$\boxed{\theta \mid y \sim \text{Beta}(\alpha + 1, \beta + y - 1)}$$

::: {#big-text style="font-size: 110%"}
**Summary** 
:::


| Component | Distribution | Parameters |
|-----------|--------------|-----------|
| Prior | Beta | $\alpha, \beta$ |
| Likelihood | Geometric | θ |
| Posterior | Beta | $\alpha + 1, \beta + y - 1$ |

***

::: {#big-text style="font-size: 125%"}
**b) Is the Beta model a conjugate prior for the Geometric data model?** 
:::





**Key insight:** The Beta distribution is the **conjugate prior** for the Geometric likelihood, meaning the posterior remains in the same Beta family. The posterior parameters show that observing a single geometric trial with outcome y increases the first parameter by $1$ and the second parameter by $y - 1$.

::::
:::::

## Glossary Entries {.unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```



------------------------------------------------------------------------

## Session Info {.unnumbered}

::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::
