# MCMC Under the Hood {#sec-chap-007}

```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")
ggplot2::theme_set(ggplot2::theme_bw())
```



::::: {#obj-chap-007}
:::: {.my-objectives}
::: {.my-objectives-header}

- Build a strong conceptual understanding of how Markov chain algorithms work. 
- Explore the foundational Metropolis-Hastings algorithm. 
- Implement the Metropolis-Hastings algorithm in the Normal-Normal and Beta-Binomial settings.
:::

::: {.my-objectives-container}


:::
::::
:::::


Bayesian `r glossary("MCMC")` simulation is a rich field, spanning various algorithms that share a common goal: approximate the posterior model. For example, the {**rstan**} package utilized throughout this book employs a Hamiltonian Monte Carlo (`r glossary("HMC")`) algorithm. The alternative {**rjags**} package for Bayesian modeling employs a `r glossary("Gibbs")` sampling algorithm. Though the details differ, both algorithms are variations on the fundamental `r glossary("Metropolis Algorithm", "Metropolis-Hastings")` algorithm.

## The big idea

It helps to think of the illustrated Markov chain $\{\mu^{(1)}, \mu^{(2)}, \cdots , \mu^{(N)}\}$ as a **tour** around the range of posterior plausible values of $\mu$ and yourself as the **tour manager**.

As tour manager, it’s your job to ensure that the density of tour stops in each $\mu$ region is proportional to its posterior plausibility. That is, the chain should spend more time touring values where the posterior pdf is greatest, and less time visiting $\mu$ values where the posterior drops off. This consideration is crucial to producing a collection of tour stops that accurately approximate the posterior.

The Metropolis-Hastings algorithm iterates through a two-step process. Assuming the Markov chain is at location $\mu^{(i)} = \mu$ at iteration or “tour stop” $i$, the next tour stop $\mu^{(i+1)}$ is selected as follows:  

- **Step 1:** Propose a random location, $\mu′$, for the next tour stop. 
- **Step 2:** Decide whether to go to the proposed location ($\mu^{(i+1)} = \mu′$) or to stay at the current location for another iteration ($\mu^{(i)} = \mu$.

The special case that step 2 requires to go unconditionally to the proposed location is called `r glossary("Monte-Carlo Algorithm", "Monte Carlo Algorithm")`.

:::::{.my-procedure}
:::{.my-procedure-header}
:::::: {#prp-007-monte-carlo-algorithm}
: Monte Carlo Algorithm
::::::
:::
::::{.my-procedure-container}

To construct an *independent* Monte Carlo sample *directly* from posterior pdf $f \{\mu^{(1)}, \mu^{(2)}, \cdots , \mu^{(N)}\}$, select each tour stop    $\mu^{(i)} = \mu$ as follows:  

- **Step 1: Propose a location**.  Draw a location $\mu$ from the posterior model with pdf $f (\mu \mid y)$.  
- **Step 2: Go there**.


::::
:::::


Consider the following `r glossary("Normal-Normal Conjugate Family", "Normal-Normal model")` with numerical outcome $Y$ that varies Normally around an unknown mean $\mu$ with a standard deviation of 0.75:

$$
\begin{align*}
Y \mu \mu &\sim \mathcal{N}(\mu, 0.75^2) \\  
\mu &\sim \mathcal{N}(0, 1^2)
\end{align*}
$$ {#eq-007-normal-normal-example}

Suppose we observe an outcome $Y = 6.25$. Then by our Normal-Normal work in @sec-chap-005 @eq-005-normal-normal-posterior-model, the posterior model of $\mu$ is Normal with mean $4$ and standard deviation $0.6$:

$$
\begin{align*}
\mu \mid \vec{y} \sim N \left(\theta \frac{\sigma^2}{n \tau^2 + \sigma^2} + \overline{y} \frac{n\tau^2}{n \tau^2 + \sigma^2}, \frac{\tau^2 \sigma^2}{n \tau^2 + \sigma^2} \right)
\end{align*}
$$

::: {#big-text style="font-size: 110%"}
**Posterior Variance** 
:::

$$
\begin{align*}
\frac{1}{\tau_n^2} &= \frac{1}{\tau^2} + \frac{1}{\sigma^2} \\
\frac{1}{\tau_n^2} &= \frac{1}{1^2} + \frac{1}{0.75^2} \\
\frac{1}{\tau_n^2} &= 1 + \frac{1}{0.5625} \\
\frac{1}{\tau_n^2} &= 1 + 1.7\overline{7} \\
\frac{1}{\tau_n^2} &= 2.7\overline{7} \\
\frac{1}{\tau_n^2} &= \frac{25}{9} \\
\tau_n^2 &= \frac{9}{25} = 0.36 \quad \Rightarrow \quad \tau_n = 0.6
\end{align*}
$$ 


::: {#big-text style="font-size: 110%"}
**Posterior Mean** 
:::


$$
\begin{align*}
\mu_n &= \frac{\frac{\theta}{\tau^2} + \frac{y}{\sigma^2}}{\frac{1}{\tau^2} + \frac{1}{\sigma^2}} \\
&= \frac{\frac{0}{1} + \frac{6.25}{0.5625}}{2.7\overline{7}} \\
&= \frac{0 + 11.1\overline{1}}{2.7\overline{7}} \\
\mu_n &= \frac{100/9}{25/9} = \frac{100}{25} = 4 \\
\end{align*} 
$$ 


::: {#big-text style="font-size: 110%"}
**Posterior Distribution** 
:::


$$\boxed{\mu \mid Y = 6.25 \;\sim\; \mathcal{N}(4, 0.36)}$$

Or equivalently: 

$$\boxed{\mu \mid Y = 6.25 \sim \mathcal{N}(4, 0.6^2)}$$

::: {#big-text style="font-size: 110%"}
**Interpretation** 
:::

| Aspect | Value |
|--------|-------|
| **Posterior mean** | 4.0 |
| **Posterior SD** | 0.6 |
| **95% credible interval** | $[4 \pm 1.96(0.6)] = [2.82, 5.18]$ |

::: {#big-text style="font-size: 110%"}
**Key insights** 
:::

1. The posterior mean (4.0) is **pulled toward the data** (6.25) but **tempered by the prior** (0), reflecting both sources of information.
2. The posterior is **much more concentrated** than the prior (SD: 0.6 vs 1.0), because the data provides strong evidence about $\mu$.
3. We're now 95% confident that $\mu$ lies between **2.82 and 5.18**, a substantial reduction in uncertainty from the prior.

***

Posterior Variance: $\tau_n^2 = \dfrac{9}{25}$

This is simply taking the reciprocal of $\dfrac{25}{9}$:

$$\frac{1}{\tau_n^2} = \frac{25}{9} \quad \Longrightarrow \quad \tau_n^2 = \frac{9}{25} = 0.36$$

The $\dfrac{25}{9}$ itself comes from converting $2.7\overline{7}$ to a fraction:

$$1 + \frac{1}{0.5625} = 1 + \frac{16}{9} = \frac{9}{9} + \frac{16}{9} = \frac{25}{9}$$

since $0.75^2 = 0.5625 = \dfrac{9}{16}$, so $\dfrac{1}{0.5625} = \dfrac{16}{9}$.


Posterior Mean: $\mu_n = \dfrac{100/9}{25/9}$

The **numerator** $\dfrac{100}{9}$ comes from:

$$\frac{y}{\sigma^2} = \frac{6.25}{0.5625} = \frac{6.25}{9/16} = 6.25 \times \frac{16}{9} = \frac{100}{9}$$

The **denominator** $\dfrac{25}{9}$ is just $\dfrac{1}{\tau_n^2}$ from above.

So the full calculation is:

$$\mu_n = \frac{\frac{100}{9}}{\frac{25}{9}} = \frac{100}{9} \times \frac{9}{25} = \frac{100}{25} = 4$$

How to convert $\dfrac{1}{0.5625}$ into $\dfrac{16}{9}$:


**Step 1**: Convert the decimal to a fraction

First, recognize that $0.5625$ is a terminating decimal. I can express it as:

$$0.5625 = \frac{5625}{10000}$$

**Step 2**: Simplify by finding the GCD

Now I need to reduce this fraction. Finding the greatest common divisor (GCD) of 5625 and 10000:

- $5625 = 5^4 \times 9 = 625 \times 9$
- $10000 = 10^4 = 2^4 \times 5^4$

The GCD is $5^4 = 625$, so:

$$\frac{5625}{10000} = \frac{5625 \div 625}{10000 \div 625} = \frac{9}{16}$$

**Step 3**: Take the reciprocal

$$\frac{1}{0.5625} = \frac{1}{\frac{9}{16}} = \frac{16}{9}$$

***

**Alternative approach:** You could also recognize that $0.75^2 = (3/4)^2 = 9/16$, so immediately $\dfrac{1}{0.75^2} = \dfrac{16}{9}$ without converting the decimal at all!

::: {.callout-note #nte-007-stopped-here}
##### I stopped my notes for this chapter here

This chapter helped me to get a better understanding of how `r glossary("Markov chain")` algorithms work. But because it is not essential to grasp the following chapters of "Bayes rules!" I skipped writing notes for this chapter.
:::



## Glossary Entries {.unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```

------------------------------------------------------------------------

## Session Info {.unnumbered}

::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::

