# First Contact with Stan {#sec-chap-905}

```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")
ggplot2::theme_set(ggplot2::theme_bw())
```

:::::: {#obj-chap-905}
::::: my-objectives
::: my-objectives-header
Objectives
:::

::: my-objectives-container

This chapter experiments with [RStan: the R interface to
Stan](https://cran.r-project.org/web/packages/rstan/vignettes/rstan.html) and [CmdStanR: R interface to 'CmdStan'](https://mc-stan.org/cmdstanr/).
My aim is to get a better understanding of the underlying functionality
of **RStan** using {**rstan**}, and of **CmdStanR** using (**cmdstanr**). I look into both packages and investigate especially their connections to the `r glossary("Stan platform")`.

After applying the books recommendation to use the `rstan::stan()` function several times, especially
during working of the Exercises in @sec-006-exercises, I came up with
some questions such as:

-   How to use `.stan` files for applying the Stan modelling language?
-   What do the MCMC sampling messages mean?
-   How to minimize execution time for the simulations?

This appendix collects my experiments and gives answer to the questions
above. In the end I decided during the exercises for @sec-chap-006 to use the CmdStanR approach instead of RStan. To compare both packages I have used as a central example --- functioning as an
`r glossary("advanced organizer")`. I applied the
`r glossary("Beta-Binomial-Model", "Beta-Binomial model")` in
@sec-006-beta-binomial-example to demonstrate the code differences.

:::
:::::
::::::

## Introduction

::: {#nte-905-terminology-using-leo .callout-note}

###### Using LLMs for clarification of terminology and procedures

The book "Bayes rules!" aims to reduce complexity by honing the process
of Bayesian modeling, sampling, and inference to the bare bones. But
this reduction in complexity comes at the expense of conceptual clarity
and modeling flexibility.

The following content is my own summary of conversations with
`r glossary("Brave Leo AI")`. It took me several days to gradually build
my understanding of flexible Bayesian understanding.

In @sec-chap-006, I had my first contact with the {**rstan**} package.
After applying the code snippets guided by the book examples, at least
since working on the exercises, I became aware that I did not understand
the whole picture:

-   What is the relationship of {**rstan**} to the
    [Stan](https://mc-stan.org/), (another?) software for Bayesian Data
    Analysis?
-   What are the relationships to other software packages, like
    {**cmdstanr**} (R Interface to 'CmdStan'), {**rstanarm**} (Bayesian
    Applied Regression Modeling via Stan), {**brms**} (Bayesian
    Regression Models using 'Stan').

These questions are not only important conceptually but also
practically. It needs a deeper understanding to guide efficient
practice:

-   How to minimize execution time for statistical modeling and
    sampling?
-   How do the two steps (defining the model and sampling) work
    together?
-   When to use which package?

**Naming conventions:**

-   When I refer in prose to the package names I am going to use
    capitalization like `RStan` or `CmdStanR.`
-   When I refer to the package names formally I use {**rstan**} resp.
    {**cmdstanr**}.
-   When I use the package names in R code I write `rstan::<function_name>` or `cmdstanr::<function_name>`


:::

:::::: my-resource
:::: my-resource-header
::: {#lem-905-stan-resources}
: Used RStan resources
:::
::::

::: my-resource-container
-   The most authoritative resource is the manual [Stan: Software for
    Bayesian Data Analysis](https://mc-stan.org/). But I have to confess
    that many parts I do not understand currently (2026-03-17).
-   The next session draws therefore from the vignette [RStan: the R
    interface to
    Stan](https://cran.r-project.org/web/packages/rstan/vignettes/rstan.html),
    an introduction to using {**RStan**}, the R interface to
    [Stan](https://mc-stan.org/). Even the majority of the introductory
    vignette content is still too complex for my understanding. But to
    get a better understanding for the interaction between RStan and
    Stan software is exactly the goal of the exerpiments in this annex.
-   Additionally I have addressed my questions one by one asking
    `r glossary("LLMs")`. Many of my questions were about details and
    adaptions of the modeling procedure covered in [Bayes
    Rules!](https://www.bayesrulesbook.com/).
:::
::::::


## Different Software for Bayesian Analysis {#sec-905-software-for-bayesian-analysis}


### Stan ≠ R

Stan is **not** an R package. It's a separate probabilistic programming
language with its own syntax. RStan and CmdStanR are just **interfaces**
between R and Stan. When you write a Stan model, you're writing in a
language that R cannot execute directly.

-   **Stan** is the core probabilistic programming language and
    inference engine that performs Bayesian computation using algorithms
    like `r glossary("NUTS")` (No-U-Turn Sampler), optimization
    (`r glossary("BFGSx", "BFGS")`/`r glossary("L-BFGS")`), and
    variational inference (`r glossary("ADVI")`).
-   **CmdStan** is the command-line interface to Stan. It doesn’t add
    new functionality; it’s just a way to execute Stan models without
    needing R or Python wrappers. It’s how you run Stan programs from
    the terminal or via other interfaces like
    -   **CmdStanR** is an R package that provides a lightweight
        interface to Stan calling the command-line interface CmdStan for
        compilation and running algorithms. It avoids direct C++
        integration via {**Rcpp**}. This approach provides benefits like
        compatibility with the latest Stan versions, fewer installation
        errors, active maintenance, lower memory overhead, and a more
        permissive license. It's the appropriate choice for users who
        want to use the latest Stan features or work with custom Stan
        models. It’s also ideal for modular workflows where Stan handles
        computation and downstream R packages (like {**posterior**} or
        {**bayesplot**}) handle analysis. CmdStanR is currently
        generally recommended for new projects.
    -   **RStan** is an R package {**rstan**} that provides an in-memory
        interface relying on R packages like {**Rcpp**} and inline to
        call C++ code from R. While it offers familiar R syntax, it may
        lag behind in supporting the latest Stan features due to the
        complexity of maintaining compatibility with Stan releases.

### Different layers:

#### The car analogy

Think of the different terms and packages this way:

-   **Hardware (Processor)** = The actual physical engine that does
    computation — the CPU executing machine code .
-   **Operating System** = The mechanic's shop that manages how the
    engine runs, allocates resources, handles communication between
    components .
-   **Stan** = The engine design and fuel (the probabilistic programming
    language and algorithms like NUTS).
-   **RStan** = You hire a mechanic (Rcpp) to bolt Stan's engine
    directly into your R car's chassis. The mechanic must rebuild the
    engine for your specific shop (OS) and car model (R version). If
    Stan releases a new engine design, the mechanic must rebuild
    everything — this takes time and often breaks things .
-   **CmdStan** = A standalone engine factory that produces ready-made
    Stan engines. The factory handles all OS and processor details
    internally. --- Your R car sends a messenger (command-line
    interface) to the factory, orders a Stan engine run, and the factory
    delivers results via files. The factory (CmdStan) handles
    OS/processor compatibility — your R car doesn't need to know the
    details.

#### The place of R in the car analogy

Think of the role of R this way:

You’re the driver — you tell the car *where to go* (specify the model),
and the car’s systems (R + CmdStanR + Stan) handle *how to get there*
(computation) without you needing to know engine mechanics.

**R (or Python)** = The driver’s cockpit and dashboard — it’s where you
sit, steer, and interact with the car. R is your control panel. You
don’t need to understand the engine (Stan), the factory (CmdStan), or
the mechanic (Rcpp) to drive — you just use the controls (functions like
`brms::brm()` or `rstanarm::stan_glm()`).

-   **R/Python** provides the *user interface* — the commands, data
    structures, and visualization tools you use to specify models, run
    analyses, and interpret results.
-   It’s the *glue* that connects you to the engine (Stan) via
    interfaces like CmdStanR or RStan.
-   It handles data preprocessing, model diagnostics, plotting, and
    integration with tidyverse/tidymodels — all the “applied statistics”
    work you care about.

### Stan, RStan, CmdStanR and brms

#### RStan versus CmdStanR

**Stan is the core engine**, and all other packages are interfaces to
it. The key difference between **RStan** and **CmdStanR** lies in how
they connect to Stan:

-   **RStan** uses Rcpp to call C++ code directly from R, which can lead
    to installation issues and compatibility problems with newer Stan
    versions (see: [CmdStanR: the R interface to
    CmdStan](https://mc-stan.org/cmdstanr/reference/cmdstanr-package.html)).
    It may also be slower for some tasks due to overhead in reading
    output files (see: [Google
    Groups](https://groups.google.com/g/stan-dev/c/BujPDKpUDp0/m/zmImoHyNacYJ)).
-   **CmdStanR** calls the command-line interface (CmdStan) instead,
    avoiding direct C++ integration. This results in fewer installation
    errors, better compatibility with the latest Stan releases, and a
    more permissive license (see: [Package
    {cmdstanr}](https://stan-dev.r-universe.dev/cmdstanr)). CmdStanR is
    not hosted on `r glossary("CRAN")`, but available at
    <https://stan-dev.r-universe.dev/cmdstanr>. However, it may be
    slower for simple models due to file I/O overhead. But this
    performance difference (e.g., 1s vs 3s for a simple model) is
    negligible for my work. My focus is on stability and ease of use
    rather than micro-optimizations.

See also the summary section [Comparison with
RStan](https://mc-stan.org/cmdstanr/articles/cmdstanr.html#comparison-with-rstan)
of the vignette "Getting started with CmdStanR.

#### Using CmdStanR from the beginning

For learning Bayesian statistics, `r glossary("Claude Haiku")` recommends to **start
with cmdstanr from the beginning**:

**Easier installation**: {**cmdstanr**} avoids the C++ compiler
configuration issues that often plague {**rstan**}
setup[^905-rstan-tutorial-1]. This lets you focus on learning statistics
rather than troubleshooting installation problems.

[^905-rstan-tutorial-1]: Some of the installation complexities is
    explained in the [FAQ for the Aalto course
    2025](https://avehtari.github.io/BDA_course_Aalto/FAQ.html).

**Modern design**: {**cmdstanr}** is the newer, actively developed
interface. It doesn't directly call C++ from R like {**rstan**} does;
instead, it uses CmdStan behind the scenes for compilation and running
algorithms[^905-rstan-tutorial-2]. This separation actually makes
{**cmdstanr}** more reliable and easier to maintain.

[^905-rstan-tutorial-2]: See the vignette: [Getting started with
    CmdStanR](https://mc-stan.org/cmdstanr/articles/cmdstanr.html)

**Better for teaching**: {**cmdstanr}** provides a similar user
experience across different programming languages. So if you ever learn
Stan in other languages, the concepts transfer directly.

**Performance advantages**: {**cmdstanr}** supports within-chain
multithreading more intelligently and tends to compile models faster
with better detection of when recompilation is
needed[^905-rstan-tutorial-3].

[^905-rstan-tutorial-3]: See documentation for
    [parallelization](https://mc-stan.org/docs/cmdstan-guide/parallelization.html).

The only reason to learn {**rstan**} first would be if one is working with
legacy code or packages that specifically require it. Since I am
starting fresh, {**cmdstanr**} is the better choice.

::: {#nte-905-cmdstanr-over-rstan .callout-note}

###### Applying CmdStanR instad of Rstan

Brave Leo AI and other sources recommend CmdStanR {**cmdstanr**} over
RStan {**rstan**}. Summarized the reasons that matters me most are threefold:

-   speed,
-   stability,
-   and up-to-date integration with Stan.

Instead following "Bayes rules!" approach with {**rstan**} I will therefore apply
in this text mainly {**cmdstanr**}.

:::

#### And what about {**brms**}?

The only disadvantage of CmdStanR is that I have to learn Stan syntax.
There is the possibility of using {**brms**} as an even higher-level
interface on top of CmdStanR and Stan for standard models, without
learning Stan syntax.

{**brms**} is an R package that provides a flexible interface for
fitting **Bayesian generalized (non-)linear multivariate multilevel
models** using **Stan**, the most widely used probabilistic programming
language for full Bayesian inference.

-   It supports a wide range of response distributions and link
    functions, enabling users to fit models for **linear, robust linear,
    count data, survival, response times, ordinal, zero-inflated,
    hurdle, and mixture models** within a multilevel framework.
-   The package uses a formula syntax similar to {**lme4**,} making it
    accessible to users familiar with frequentist mixed models.
-   Key features include:
    -   **Distributional regression**: All parameters of the response
        distribution (e.g., mean, variance, shape) can be modeled as
        functions of predictors.
    -   **Non-linear and smooth terms**: Supports theory-driven and
        data-driven non-linear effects.
    -   **Advanced modeling options**: Auto-correlation structures,
        censoring, truncation, meta-analytic standard errors, missing
        data imputation, and spatial models (e.g., CAR structures).
    -   **Flexible prior specification**: Encourages the use of
        informative priors that reflect prior knowledge.
    -   **Model evaluation**: Offers tools for posterior predictive
        checks, leave-one-out cross-validation (LOO), WAIC, and Bayesian
        R-squared.

The package is actively maintained, with version **2.23.1** as of
January 2026, and is available on `r glossary("CRAN")` and `r glossary("GitHub")`. {**brms**} is widely used in social sciences,
biostatistics, and other fields requiring robust, flexible Bayesian
modeling.

#### Andvantages of starting with {**brms**}

- {**brms**} has the major advantage that I **don't need to learn Stan syntax** at all. The package uses familiar R formula notation (like `y ~ x1 + x2`), making Bayesian modeling accessible without Stan knowledge.
- {**brms**} is not "advanced only" — it is actually suited **for beginners and practitioners who want to fit standard Bayesian models quickly**. One can do serious applied work without touching Stan code. One gets Bayesian inference without the Stan learning curve.
- {**brms**} original **documentation is excellent** and comprehensive. Additionally there are many other tutorials to learn this package. 
- With `brms::stancode(<model>)` one can **inspect generated Stan code**. Learning through inspection (seeing what {**brms**} generates) is often more efficient than learning Stan syntax from scratch.
- Bayesian diagnostics (trace plots, Rhat, effective sample size) work the same way regardless of interface. They catch problems which is another learning possibility to apply {**brms**} correctly.

For all these advantages `r glossary("Claude Haiku")` stated: **Start with {brms}, learn {cmdstanr} only if needed.**

**When to learn CmdStanR:**

- When {**brms**} can't express your model (e.g., custom likelihood, complex hierarchical structures)
- After I have internalized Bayesian thinking through {**brms**} workflows
- To deepen conceptual understanding of what's happening "under the hood"

#### Looking into brms sporadically

In contrast to Claude Haiku’s recommendation, I will start with {**cmdstanr**}. The reason is that I am afraid I will digress too much from the book content on "Bayes rules!" This annex already gives me a headache, because I have to learn everything from scratch. It is correct that `r glossary("LLMs")` like `r glossary("Brave LEO AI")` or `r glossary("Positron Assistant")` are very helpful. But their role as tutors is limited for three reasons:

- Sometimes they are simply inaccurate or even wrong.
- Sometimes they misinterpret my questions.
- But above all: **I myself** have to ask the right questions and to determine the direction. There is no guidance from a domain expert on the next step in my learning career.

I am therefore planning three consecutive learning steps:

1. Understanding Bayesian modeling by reading "Bayes rules!" Especially trying to solve the exercises are very helpful in this respect.
2. Translating the book's code into {**cmdstanr**} functions and writing it down as notes following the structure of the book.
3. Looking into {**brms**} and trying to develop the book models with {**brms**} syntax. But my text to this part will perhaps not be complete and I will therefore write these kind of notes separately in annex chapters.

::: {.callout-note #nte-905-future-plans}

###### Future Plans

After I have finished "Bayes rules!" [@johnson2022] I plan to foster my understanding of Bayesian analysis with "Statistical Rethinking" [@mcelreath2023] followed by "Bayesian Data Analysis" [@gelman2013].

:::

## Compilation & MCMC Sampling

The function used in the book `rstan::stan()` combines these two process phases into "# STEP 2: SIMULATE the posterior". But using the nomenclature of the book there are actually three steps:

### Defining the model

Writing the data, parameters, model, and generated quantities blocks in Stan, a probabilistic programming language with its own syntax (Step 1 in the book). There are two options: Writing a character string as done in the book, or recommended -- especially with complex models -- writing the definition into a file.
- Compilation:
- Sampling

### Compilation

The compilation step translates your Stan model through **multiple
layers**:

#### Stan ->  C++ -> Machine Code

1.  **Stan → C++**: The Stan compiler (*stanc*) parses your `.stan`
    model and translates it into C++ code. This includes:
    -   Parsing the `data`, `parameters`, `model`, and
        `generated quantities` blocks
    -   Performing **automatic differentiation** setup — Stan needs
        gradients of your log-posterior for its samplers, so the C++
        code is structured to compute these efficiently
    -   Type-checking and validating the model structure
2.  **C++ → Machine code**: The generated C++ is compiled by your local
    C++ compiler (e.g., `g++` via `Rtools` on Windows) into a **native
    shared library** (`.so` / `.dll`). This is binary code your CPU can
    execute directly.

The result is a compiled model object in R — essentially a
self-contained, highly optimized executable.

#### Why This Design?

-   **Performance**: Native machine code runs orders of magnitude faster
    than interpreted code. Bayesian sampling requires evaluating the
    log-posterior and its **gradients thousands to millions of times** —
    speed is critical.
-   **Automatic differentiation**: The C++ translation sets up the
    *autodiff* tape, enabling exact gradient computation (not numerical
    approximation) needed by HMC/NUTS.

#### Relationship to the Sampling Step

The two steps have a clean **compile-once, sample-many** relationship:

| Step | What happens | Analogy |
|------------------------|------------------------|------------------------|
| **Compilation** | Translates + optimizes the model into machine code | Building a specialized calculator |
| **Sampling** | Runs HMC/NUTS using the compiled code, feeding in your data | Using that calculator repeatedly |

### Sampling (MCMC with HMC/NUTS)

There are several conceptual and computation layers in the sampling step.

#### Layer 1: **MCMC Iteration Level** (Top Layer)

- Each iteration = one proposal → acceptance/rejection decision
- You control this with `iter_sampling` (e.g., 1000 iterations)
- Each iteration generates **one posterior sample** (after warmup)
- Multiple chains run independently → better convergence diagnostics

> *Example: 4 chains × 1000 iterations = 4,000 posterior samples*

#### Layer 2: **Trajectory Construction** (HMC/NUTS Core)

- Within each iteration, Stan builds a **Hamiltonian trajectory** using **leapfrog steps**
- NUTS automatically chooses how many steps to take (avoids manual tuning)
- Each leapfrog step:
  - Updates position (parameters: e.g., `pi`)
  - Updates momentum (auxiliary variable)
  - Uses gradient of log-posterior

> *This is where the “simulation” happens — the sampler explores the parameter space*

#### Layer 3: **Gradient Evaluation** (Computational Core)

- At each leapfrog step, Stan computes:
  - **Log-posterior**: `log(p(data | parameters) × p(parameters))`
  - **Gradient**: ∇ log-posterior (via automatic differentiation)
- This is where the **compiled C++ code** runs — fast and exact
- The `r glossary("gradient")` tells the sampler which direction to move for efficient exploration

> *This is the “engine” — the reason Stan is so fast and accurate*

#### Layer 4: **Model Evaluation** (Lowest Layer)
- The compiled model evaluates:
  - Likelihood: `Y ~ binomial(n_trials, pi)`
  - Prior: `pi ~ beta(2, 2)`
  - Generated quantities (if any)
- All this is pre-compiled into C++ → runs at machine speed

> *This is the “model structure” you defined — now being executed millions of times*



#### Sampling summary

| Layer | What Happens | Controlled By | Frequency |
|-------|--------------|-------------|---------|
| 1. MCMC Iteration | One sample per iteration | `iter_sampling` | 1,000–10,000× |
| 2. Leapfrog Steps | Trajectory construction | NUTS (automatic) | 10–100× per iteration |
| 3. Gradient Eval | Compute log-posterior + gradient | Compiled model | 10–100× per iteration |
| 4. Model Eval | Evaluate likelihood + prior | Stan code | 10–100× per iteration |


### Key implications

The whole process is also called **simulating the posterior**

- An flexible defined model **does not contain data**
- The compiled object **knows nothing about your data yet** 
— Data is passed at sampling time 
- You can **reuse the compiled model** with different datasets or priors without recompiling 
- Sampling calls the compiled log-posterior + gradient function **at every leapfrog step** of HMC, which is why having it in fast native code matters so much

Compilation is a **one-time translation cost** that makes the repeated,
expensive work of sampling as fast as possible. The sampler is
essentially a tight loop calling the compiled C++ function millions of
times — without compilation, this would be prohibitively slow in
interpreted R or Python.

## CmdStanR versus RStan

In "Different Software for Bayesian Analysis" @sec-905-software-for-bayesian-analysis I decided to change from RStan {**rstan**} to CmdStanR {**cmdstanr**}. The next step is to transfer the RStan "Beta-Binomial example" @sec-006-beta-binomial-example into CmdRStanR and to compare the two approaches.

The model definition written in Stan is by and large in both approaches the same. Nevertheless I found five possible improvements.



Here is a detailed comparison of the **modern CmdStanR code** with the **"Bayes Rules!" code**:

### Model definition

The model definition written in Stan is by and large in both approaches the same. Nevertheless I found several improvements.

#### **Integration approach:** `.stan` file vs. character string

::: {.callout-note #nte-905-writelines-to-stan-file}
##### Generating `.stan` file programmatically

What follows was my first solution. I wrote the `.stan` file and included it later in the MCMC simulation.

But there is a better approach: Writing in an R chunk the model definition and store it programmatically into the `.stan` file. For the details of this approach see: @lst-exr-612-define-and-write-bb-model-to-file.
:::


The first improvement does not address the content of the model definition but the mode of integration. In contrast to the "Bayes rules!" book which uses a character string within R to describe the model, the recommended approach is to use a separate file with a `.stan` extension.

To make the comparison between the book and my version I will show in the first part of the following experiment the code from "Bayes rules!".

:::::: my-experiment
:::: my-experiment-header
::: {#def-905-stan-file}
: Display `.stan` file with model code in Quarto document
:::
::::

::: my-experiment-container
There are two options:

1.  Using R `base::cat(base::readLines())` function: You need to apply the `#| class-output: stan` chunk option. The following code lines would the `.stan` file include into the Quarto document:

```r
#| class-output: stan

cat(readLines("include/stan/905-exp-rstan-beta-binomial.stan"), sep = "\n")

```

The final `\n` is necessary, because when a Quarto file is missing a trailing empty line the R parser may issue an "incomplete final line" warning or fail to read the last row correctly, as the parser expects the file to end with a newline character.

The above code results in the following code snippet.


```{r}
#| class-output: stan
#| label: rstan-beta-binomial
#| lst-label: lst-rstan-beta-binomial

cat(readLines("include/stan/905-exp-rstan-beta-binomial.stan"), sep = "\n")
```


2.  Using the
    [include-code-files](https://github.com/quarto-ext/include-code-files)
    filter extension. This method renders the Stan code with syntax
    highlighting, treating it as a Stan code block. It supports other
    features too, like line numbers, or not inserting the whole file, but a range of lines.  (This time I refer to the file with the improved version)

````
``` {.stan include="include/stan/905-exp-cmdstanr-beta-binomial.stan" code-line-numbers="true"}
```
````

The above code results in the following code snippet:

```{.stan include="include/stan/905-exp-cmdstanr-beta-binomial.stan" code-line-numbers="true"}

```



:::
::::::



#### **Side-by-Side Comparison:** CmdStanR vs. RStan

| Aspect | Modern Code (CmdStanR) | "Bayes Rules!" Code |
|--------|------------------------|-------------------|
| **Data block structure** | Parameterized with `n`, `n_trials` | Hardcoded with `upper = 10` |
| **Array syntax** | `array[n] int<lower=0, upper=n_trials> Y` | `int<lower = 0, upper = 10> Y` |
| **Flexibility** | Reusable for any number of trials | Fixed for 10 trials only |
| **Likelihood** | `Y ~ binomial(n_trials, pi)` | `Y ~ binomial(10, pi)` |
| **Prior/Likelihood order** | Prior first, then likelihood | Likelihood first, then prior |
| **Spaces in syntax** | No spaces around `=` and `<` | Spaces around `=` and `<` |


#### **Data Block:** Flexibility vs. Simplicity

**Modern Code:**
```stan
data {
  int<lower=0> n;             // number of observations
  int<lower=0> n_trials;      // number of trials
  array[n] int<lower=0, upper=n_trials> Y;
}
```

**"Bayes Rules!" Code:**
```stan
data {
  int<lower = 0, upper = 10> Y;
}
```

**Difference:**

- Modern code is **parameterized** — you can pass `n_trials` as data, making the model reusable.
- "Bayes Rules!" is **hardcoded** — the number of trials (10) is fixed in the model definition.
- Modern code declares `n` (number of observations) for potential multiple observations.


#### **Array Syntax:** Modern vs. Legacy

**Modern Code:**
```stan
array[n] int<lower=0, upper=n_trials> Y;
```

**"Bayes Rules!" Code:**
```stan
int<lower = 0, upper = 10> Y;
```

**Difference:**

- Modern uses **`array[n]` syntax** (Stan 2.26+) — more intuitive, size comes first.
- "Bayes Rules!" uses **scalar declaration** — only one observation `Y`.
- Modern code scales to multiple observations; "Bayes Rules!" is single-observation only.


#### **Model order:** Prior first, then Likelihood

**Modern Code:**
```stan
model {
  // Prior
  pi ~ beta(2, 2);

  // Likelihood
  Y ~ binomial(n_trials, pi);
}
```

**"Bayes Rules!" Code:**
```stan
model {
  Y ~ binomial(10, pi);
  pi ~ beta(2, 2);
}
```

**Difference:**

- Modern code lists **prior first, then likelihood** (pedagogically clearer for Bayesian thinking).
- "Bayes Rules!" lists **likelihood first, then prior** (mathematically equivalent, but less intuitive).
- **Functionally identical** — Stan doesn't care about order.


#### **Spacing Convention:** No spaces around operators

**Modern Code:**
```stan
int<lower=0, upper=n_trials> Y;
```

**"Bayes Rules!" Code:**
```stan
int<lower = 0, upper = 10> Y;
```

**Difference:**

- Modern code uses **no spaces** around `=` and `<` (modern Stan style).
- "Bayes Rules!" uses **spaces** (older style, still valid).
- **Functionally identical** — both are valid syntax.


#### **Processing:** Separate steps vs. combined call

**Modern Code uses CmdStanR:**

```r
model <- cmdstan_model("model.stan")
fit <- model$sample(data = data_list, chains = 4, ...)
```

**"Bayes Rules!" likely uses RStan:**
```r
fit <- stan(file = "model.stan", data = data_list, chains = 4, ...)
```

**Difference:**

- **CmdStanR** — separate compilation and sampling steps: more explicit, and faster.
- **RStan** — combined in one `rstan::stan()` call: simpler, but less transparent. (But one could in Rstan also separate compilation and sampling with `rstan::stan_model()` and `rstan::sampling()`.)


#### **Summary Table:** CmdStanR vs. RStan

| Feature | Modern Code | "Bayes Rules!" |
|---------|------------|----------------|
| **Generalizability** | ✅ Reusable for any `n_trials` | ❌ Fixed to 10 trials |
| **Multiple observations** | ✅ Supports `n > 1` | ❌ Single observation only |
| **Stan version** | ✅ Modern (2.26+) | ✅ Works in all versions |
| **Readability** | ✅ Clear prior/likelihood structure | ✅ Simple and concise |
| **Production use** | ✅ Flexible and scalable | ❌ Limited to specific case |


::: {.callout-important #imp-905-lesson-learned-model-definition}

###### Model definition: Lesson learned

- **For learning** (like "Bayes Rules!") — The simple version is fine; it's easier to understand.
- **For research/production** — Use the modern, parameterized version for flexibility and reusability.
- **Best practice** — Always parameterize fixed values like `n_trials` so your model generalizes.

:::


## Compilation Caching with CmdStanR

It was important for my understanding to separate compilation and sampling. Beside of the conceptual reason separating compilation from sampling has another advantage: You can `r glossary("cache")` the compiled model and run it many times with different data without repeating the time-consuming compilation step.

### Why This Matters

| Aspect | CmdStanR | rstan |
|--------|----------|-------|
| **Compilation** | Once per session | Every `stan()` call |
| **Time saved** | Significant (especially for complex models) | None |
| **Efficiency** | ✅ Excellent for multiple analyses | ❌ Wasteful |
| **Use case** | Fitting same model to different data | Single fit |


For instance I could run my Beta-Binomial example model with a different number of trials. See as an example the following `r glossary("pseudocode")`: 


```r
## Pseudocode Snippet

library(cmdstanr)

# Compile once
model <- cmdstan_model("model.stan")

# Fit to different datasets without recompilation
fit_10_trials <- model$sample(data = list(n = 1, n_trials = 10, Y = array(9)), ...)
fit_20_trials <- model$sample(data = list(n = 1, n_trials = 20, Y = array(18)), ...)
fit_100_trials <- model$sample(data = list(n = 1, n_trials = 100, Y = array(90)), ...)
```

**All three fits use the same compiled model** — this arrangement is a huge time saver!


### Default Behavior (Without Explicit Caching)

**By default, re-knitting a Quarto document will recompile the model** because:

- Each code chunk runs in a fresh R session (or at least a new environment).
- The `model` object from `cmdstan_model()` is not automatically persisted between knits.

```r
## Pseudocode Snippet

# This gets re-run every time you knit
model <- cmdstan_model("model.stan")  # ← Recompiles each time!

fit <- model$sample(data = data_list, chains = 4, ...)
```

### How to Cache the Compiled Model

#### **Option 1:** Use Quarto Code Block Options (Recommended)

**Pseudocode snippet**

```r
#| cache: true

library(cmdstanr)
model <- cmdstan_model("model.stan")
```

With `cache: true`:

- ✅ First knit: compiles the model and caches it.
- ✅ Subsequent knits: uses the cached compiled model (no recompilation).
- ⚠️ Cache invalidates if you modify the Stan file or the R code.



#### **Option 2:** Use `refresh = 0` (Fastest)

**Pseudocode snippet**

```r
model <- cmdstan_model("model.stan", refresh = 0)
```

- If the model is already compiled, `refresh = 0` skips recompilation.
- ✅ Faster than recompiling.

---

#### **Recommended Workflow** for Quarto

I show the pseudocode for three consecutive step:

**Pseudocode Step 1**

```r
#| cache: true
#| label: compile-model

library(cmdstanr)
model <- cmdstan_model("model.stan")
```

**Pseudocode Step 2**

```r
#| label: fit-model-1

fit1 <- model$sample(data = data_list_1, chains = 4, ...)
```

**Pseudocode Step 3**

```r
#| label: fit-model-2

fit2 <- model$sample(data = data_list_2, chains = 4, ...)
```

**More fitting steps**

…


**Benefits:**

- ✅ Model compiles once and is cached.
- ✅ Subsequent knits use the cached version.
- ✅ If you modify the Stan file, the cache invalidates automatically.
- ✅ Clean and reproducible.


#### **Summary**

| Scenario | Behavior |
|----------|----------|
| **Default (no caching)** | ❌ Recompiles every knit |
| **With `cache: true`** | ✅ Compiles once, cached on subsequent knits |
| **With `refresh = 0`** | ✅ Skips recompilation if already compiled |


**Best practice for Quarto:** 

Use `#| cache: true` on the code chunk that compiles the model. This gives you automatic caching with minimal setup!


### Saving the compiled model to disk?

#### Compiled binaries are OS-specific

At first I thought that it would be a great idea to save the compiled model to the disk. This would also save much time. Sure, it is slightly more complex than just add `#| cache: true` because

- you have to delete the file for recompilation and 
- you have to write the code for saving and loading.

Despite these small disadvantages --- so I thought --- file saving shines form the perspective of reproducibility: Everybody gets the same compiled model in each session. 

But what I didn't thought: **Compiled binaries are OS-specific** — Windows, Mac, Linux binaries differ.

#### Other disadvantages with GitHub

Using version control there are other disadvantages. Since I want others to reproduce and use my work from `r glossary("GitHub")`, `cache: true` is the better choice because:

Since you want others to reproduce your work from GitHub, **`cache: true` is the better choice** because:

✅ **Compiled binaries are NOT committed to GitHub**

- Compiled models are OS-specific (Windows, Mac, Linux binaries differ).
- Committing them creates bloat and breaks cross-platform reproducibility.

✅ **Quarto cache is automatically ignored**

- Quarto's cache directory (`.quarto/`) is typically in `.gitignore`.
- Others won't download OS-specific compiled files.

✅ **Others can reproduce on their OS**

- When someone clones my repo and knits the document, it will:
  1. Compile the Stan model for their OS.
  2. Cache it locally on their machine.
  3. Reuse the cache on subsequent knits.

✅ **Simple and transparent**

- No manual file management needed.
- The `.stan` file is all that's needed in the repo.



#### Summary

**Problems:**

- ❌ Committing the compiled binary, it's OS-specific. 
- ❌ If one doesn't commit it, others must recompile anyway (defeats the purpose).
- ❌ Adds unnecessary complexity.



## CmdStanR Code to Simulate the Posterior


To compare the two approaches I will present (but not evaluate) the RStan code snippet:

### Code snippet RStan

:::::{.my-experiment}
:::{.my-experiment-header}
:::::: {#def-905-rstan-beta-binomial-code}
: Code snippet for the RStan version of the Beta-Binomial model of chapter 6
::::::
:::
::::{.my-experiment-container}

```{r}
#| label: rstan-posterior-simulation
#| tbl-cap: "RStan posterior simulation of the Beta-Binomial example"
#| code-fold: show

# Prepare data
data_list <- list(
  n = 1,            # number of observations
  n_trials = 10,    # number of trials
  Y = array(9)      # observed successes (9 out of 10 trials)
)

# Simulate the posterior
fit_rstan <- rstan::stan(
  file = "include/stan/905-exp-cmdstanr-beta-binomial.stan", 
  data = data_list, 
  chains = 4, 
  iter = 5000*2, 
  seed = 84735,
  verbose = FALSE # output only 1000 iterations
)

# Summarize posterior
print(fit_rstan)

# Extract posterior samples
posterior_samples_rstan <- rstan::extract(fit_rstan)
pi_samples_rstan <- posterior_samples_rstan$pi
```

::::
:::::



### Code snippet CmdStanR

The following code chunk translates the RStan simulation into CmdStanR coding. 

:::::{.my-experiment}
:::{.my-experiment-header}
:::::: {#def-905-cmdstanr-beta-binomial-code}
: Code snippet for the CmdStanR version of the Beta-Binomial model of chapter 6
::::::
:::
::::{.my-experiment-container}

```{r}
#| label: cmdstanr-posterior-simulation
#| tbl-cap: "CmdStanR posterior simulation of the Beta-Binomial example"
#| code-fold: show
#| cache: true

# Prepare data
data_list <- list(
  n = 1,            # number of observations
  n_trials = 10,    # number of trials
  Y = array(9)      # observed successes (9 out of 10 trials)
)

# 1. Compile the Stan model
model <- 
  cmdstanr::cmdstan_model("include/stan/905-exp-cmdstanr-beta-binomial.stan")


# 2. Fit the model
fit_cmdstanr <- model$sample(
  data = data_list,
  chains = 4,
  iter_warmup = 5000,
  iter_sampling = 5000,
  seed = 84735,
  refresh = 0 # output only running time for each chain
)

# Summarize posterior
fit_cmdstanr$summary()

# Extract posterior samples
posterior_samples_cmdstanr <- fit_cmdstanr$draws()
pi_samples_cmdstanr <- posterior_samples_cmdstanr[, , "pi"]
```

::::
:::::


Let me explain the different code lines applying the introductory  ["Getting started with CmdStanR" vignette](https://mc-stan.org/cmdstanr/articles/cmdstanr.html).

**1. Compile the Stan model**

The `cmdstanr::cmdstan_model()` function creates a new CmdStanModel object from a file containing a Stan program. Under the hood, CmdStan is called to translate a Stan program to C++ and create a compiled executable. 

The object `model` is an [R6](https://r6.r-lib.org/) reference object (see: [Chapter R6 of Advanced R](https://adv-r.hadley.nz/r6.html)) of class [CmdStanModel](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html) and behaves similarly to R’s reference class objects and those in object oriented programming languages. Methods are accessed using the `$` operator. This design choice allows for CmdStanR and [CmdStanPy](https://github.com/stan-dev/cmdstanpy) to provide a similar user experience and share many implementation details.

Examples for the application of the `$` operator are:

- `model$print()`: prints the Stan program (the model definition in the `.stan` file).
- `model$exe_file()`: prints the path to the compiled executable.

**2. Fit the model**

The `$sample()` method for CmdStanModel objects runs Stan’s default MCMC algorithm. The `$sample()` method creates R6 [CmdStanMCMC](https://mc-stan.org/cmdstanr/reference/CmdStanMCMC.html) objects, which have many associated methods. 

Examples for the application of the `$summary()` method are:

- `fit$summary()` calls `posterior::summarise_draws()` from the {**posterior**} package. The first argument specifies the variables to summarize and any arguments after that are passed on to posterior::summarise_draws() to specify which summaries to compute, whether to use multiple cores, etc.
- `fit$draws()` extracts the posterior draws in formats provided by the posterior package.

:::::{.my-resource}
:::{.my-resource-header}
:::::: {#lem-905-posterior-draws-and-plots}
: Posterior draws and plots
::::::
:::
::::{.my-resource-container}

- The vignette [Working with Posteriors](https://mc-stan.org/cmdstanr/articles/posterior.html) has more details on posterior draws, including how to reproduce the structured output RStan users are accustomed to getting from `rstan::extract()`.

- Plotting posterior distributions is as easy as passing the object returned by the `$draws()` method directly to plotting functions in our [{**bayesplot**} package](https://mc-stan.org/bayesplot/).
::::
:::::



### Key Differences: RStan vs. CmdStanR


These two approaches accomplish the same goal — simulating from the posterior — but differ in some important practical ways:

| Aspect | RStan | CmdStanR |
|--------|-------|----------|
| **Compilation** | Automatic, happens within R | Explicit, separate step (`cmdstan_model()`) |
| **Speed** | Slower (compiles each time) | Faster (compiled model reusable) |
| **Memory** | Loads entire chains into R memory | More efficient memory handling |
| **Iteration specification** | `iter` = total iterations (includes warmup) | `iter_warmup` + `iter_sampling` (explicit separation) |
| **Posterior extraction** | `rstan::extract()` returns list | `$draws()` returns 3D array (chains × iterations × parameters) |
| **Output format** | List-based | Array-based (tidyverse-friendly) |

### Practical Workflow Implications

1. **CmdStanR is generally preferred** for modern Bayesian workflows. It's faster, more transparent about iterations, and the array output integrates better with tidyverse operations.
2. **The iteration counts differ slightly**: RStan's `iter = 10000` means 5000 warmup + 5000 sampling per chain. CmdStanR's specification is explicit, which reduces confusion.
3. **Posterior extraction**: CmdStanR's `$draws()` returns a structure that's easier to wrangle with tidyverse tools (you can convert it to a tibble more naturally).


## Glossary Entries {.unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```

------------------------------------------------------------------------

## Session Info {.unnumbered}

::::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::::
