# RStan Tutorial {#sec-chap-905}

```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")
ggplot2::theme_set(ggplot2::theme_bw())
```

:::::: {#obj-chap-905}
::::: my-objectives
::: my-objectives-header
Objectives
:::

::: my-objectives-container
This chapter follows the vignette [RStan: the R interface to
Stan](https://cran.r-project.org/web/packages/rstan/vignettes/rstan.html),
an introduction to using {**RStan**}, the R interface to
[Stan](https://mc-stan.org/).

My aim is to get a better understanding of the underlying functionality
of {**RStan**} in connection to the `r glossary("Stan platform")`.
:::
:::::
::::::

Stan is a C++ library for Bayesian modeling and inference that primarily
uses the No-U-Turn sampler (`r glossary("NUTS")`) [@hoffman2014] to obtain posterior
simulations given a user-specified model and data. Alternatively, Stan
can utilize the `r glossary("L-BFGS", "LBFGS")` optimization algorithm to maximize an objective function, such as a log-likelihood. The R package {**rstan**} provides RStan, the R interface to Stan. The {**rstan**} package allows one to conveniently
fit Stan models from R and access the output,
including posterior inferences and intermediate quantities such as
evaluations of the log posterior density and its gradients.

:::::{.my-procedure}
:::{.my-procedure-header}
:::::: {#prp-905-workflow-stan-via-rstan}
: Typical workflow for using Stan via RStan for Bayesian inference
::::::
:::
::::{.my-procedure-container}

1. Represent a statistical model by writing its log posterior density (up to an normalizing constant that does not depend on the unknown parameters in the model) using the Stan modeling language. We recommend using a separate file with a .stan extension, although it can also be done using a character string within R.
2. Fit a model with `rstan::stan()`: Following steps are performed implicitly by a single call to the `rstan::stan()` function.
    a) Translate the Stan program to C++ code using the `rstan::stanc()` function.
    b) Compile the C++ code to create a `r glossary("DSO")` (also called a dynamic link library (DLL)) that can be loaded by R.
    c) Run the DSO to sample from the posterior distribution.
3. Diagnose non-convergence of the `r glossary("MCMC")` chains.
4. Conduct inference based on the posterior sample (the MCMC draws from the posterior distribution).



::::
:::::


## Glossary Entries {.unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```

------------------------------------------------------------------------

## Session Info {.unnumbered}

::::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::::
